---
title: "XGBoost Coffee Quality Prediction"
subtitle: "Comprehensive Analysis with Multiple Modeling Approaches"
abstract: |
  This report analyzes the Coffee Quality Institute's Arabica coffee dataset using 
  XGBoost regression to predict Cupper Points (expert quality scores). The analysis 
  compares multiple feature sets (Sensory, Extended, Full), evaluates baseline models, 
  and documents both successful approaches and those found less effective (Lambda/Min_Child_Weight 
  with XGBoost providing modest but statistically significant improvements over Linear Regression.
  regularization, Bayesian Optimization). Key finding: sensory features dominate prediction, 
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
  html_document:
    toc: true
    number_sections: true
    df_print: paged
  word_document:
    toc: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, message=FALSE, warning=FALSE}
# CLEAR ENVIRONMENT
rm(list = ls(all.names = TRUE))

# LOAD PACKAGES
library(xgboost)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(knitr)
library(stringr)
library(here)
library(shapviz)

# GLOBAL CHUNK OPTIONS
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

# GGPLOT THEME
theme_set(theme_minimal())

# GLOBAL CONSTANTS
SEED <- 42
K_FOLDS <- 10
```

\newpage

# Introduction

This comprehensive report analyzes coffee quality prediction using XGBoost regression. The analysis incorporates multiple modeling approaches tested across several experiments.

**Successful Approaches:**

- Full model comparison (Sensory, Extended, Full feature sets vs baselines)
- Enhanced evaluation metrics (MAE, MAPE, Adjusted R², Max Error)
- Learning Curves for training dynamics visualization
- Exploratory Data Analysis with Correlation Analysis
- Predicted vs Actual diagnostic plots

**Approaches Tested but Found Less Effective:**

- Lambda (L2) and Min_Child_Weight regularization tuning
- Bayesian Optimization vs Grid Search comparison

## Configuration

```{r config}
config <- data.frame(
  Parameter = c("Random Seed", "Cross-Validation Folds", "Train/Test Split", 
                "Early Stopping Rounds", "Max Boosting Rounds"),
  Value = c(SEED, K_FOLDS, "80/20", "50", "2000")
)

kable(config, caption = "Analysis Configuration", format = "markdown")
```

\newpage

# Data Loading and Preprocessing

## Load Data

```{r load-data}
coffee_raw <- read.csv(here("arabica_data_cleaned.csv"))
cat("Raw data:", nrow(coffee_raw), "rows,", ncol(coffee_raw), "columns\n")
```

## Clean Categorical Variables

```{r preprocess-categorical}
# Processing Method
coffee_raw$Processing.Method <- ifelse(
  is.na(coffee_raw$Processing.Method) | coffee_raw$Processing.Method == "", 
  "Unknown", coffee_raw$Processing.Method
)
coffee_raw$Processing.Method <- factor(coffee_raw$Processing.Method)

# Variety (Top 4 + Other)
variety_counts <- sort(table(coffee_raw$Variety), decreasing = TRUE)
top_4_varieties <- names(variety_counts)[1:4]
coffee_raw$Variety <- ifelse(
  is.na(coffee_raw$Variety) | coffee_raw$Variety == "" | 
    !(coffee_raw$Variety %in% top_4_varieties),
  "Other", coffee_raw$Variety
)
coffee_raw$Variety <- factor(coffee_raw$Variety)

# Country (Top 10 + Other)
country_counts <- sort(table(coffee_raw$Country.of.Origin), decreasing = TRUE)
top_10_countries <- names(country_counts)[1:10]
coffee_raw$Country.of.Origin <- ifelse(
  is.na(coffee_raw$Country.of.Origin) | 
    !(coffee_raw$Country.of.Origin %in% top_10_countries),
  "Other", coffee_raw$Country.of.Origin
)
coffee_raw$Country.of.Origin <- factor(coffee_raw$Country.of.Origin)

# Color
coffee_raw$Color <- ifelse(
  is.na(coffee_raw$Color) | coffee_raw$Color == "",
  "Unknown", coffee_raw$Color
)
coffee_raw$Color <- factor(coffee_raw$Color)

# In-Country Partner (Top 10 + Other)
partner_counts <- sort(table(coffee_raw$In.Country.Partner), decreasing = TRUE)
top_10_partners <- names(partner_counts)[1:10]
coffee_raw$In.Country.Partner <- ifelse(
  is.na(coffee_raw$In.Country.Partner) | 
    !(coffee_raw$In.Country.Partner %in% top_10_partners),
  "Other", coffee_raw$In.Country.Partner
)
coffee_raw$In.Country.Partner <- factor(coffee_raw$In.Country.Partner)

# Region (Top 15 + Other)
region_counts <- sort(table(coffee_raw$Region), decreasing = TRUE)
top_15_regions <- names(region_counts)[1:15]
coffee_raw$Region <- ifelse(
  is.na(coffee_raw$Region) | coffee_raw$Region == "" |
    !(coffee_raw$Region %in% top_15_regions),
  "Other", coffee_raw$Region
)
coffee_raw$Region <- factor(coffee_raw$Region)

cat("Categorical variables processed\n")
cat("  Processing.Method:", nlevels(coffee_raw$Processing.Method), "levels\n")
cat("  Variety:", nlevels(coffee_raw$Variety), "levels\n")
cat("  Country:", nlevels(coffee_raw$Country.of.Origin), "levels\n")
cat("  Color:", nlevels(coffee_raw$Color), "levels\n")
```

## Clean Numeric Variables

```{r preprocess-numeric}
# Altitude (Cap at 3000m and Impute)
coffee_raw$altitude_mean_meters <- ifelse(
  coffee_raw$altitude_mean_meters > 3000 & !is.na(coffee_raw$altitude_mean_meters),
  3000, coffee_raw$altitude_mean_meters
)

region_medians <- coffee_raw %>%
  filter(!is.na(altitude_mean_meters)) %>%
  group_by(Region) %>%
  summarise(region_alt = median(altitude_mean_meters), .groups = "drop")

global_median_alt <- median(coffee_raw$altitude_mean_meters, na.rm = TRUE)

coffee_raw <- coffee_raw %>%
  left_join(region_medians, by = "Region") %>%
  mutate(altitude_mean_meters = ifelse(
    is.na(altitude_mean_meters),
    ifelse(is.na(region_alt), global_median_alt, region_alt),
    altitude_mean_meters
  )) %>%
  select(-region_alt)

# Harvest Year Parsing
parse_harvest_year <- function(val) {
  if (is.na(val) || val == "") return(NA)
  match <- str_extract(as.character(val), "(20\\d{2}|19\\d{2})")
  if (!is.na(match)) return(as.integer(match))
  return(NA)
}

coffee_raw$Harvest.Year.Parsed <- sapply(coffee_raw$Harvest.Year, parse_harvest_year)
median_year <- median(coffee_raw$Harvest.Year.Parsed, na.rm = TRUE)
coffee_raw$Harvest.Year.Parsed <- ifelse(
  is.na(coffee_raw$Harvest.Year.Parsed), median_year, coffee_raw$Harvest.Year.Parsed
)

# Bag Weight Standardization
parse_bag_weight_kg <- function(val) {
  if (is.na(val)) return(NA)
  val <- tolower(as.character(val))
  num <- as.numeric(str_extract(val, "[\\d.]+"))
  if (is.na(num)) return(NA)
  if (grepl("lb", val)) return(num * 0.453592)
  if (grepl("kg", val)) return(num)
  return(ifelse(num > 10, num, NA))
}

coffee_raw$Bag.Weight.KG <- sapply(coffee_raw$Bag.Weight, parse_bag_weight_kg)
median_weight <- median(coffee_raw$Bag.Weight.KG, na.rm = TRUE)
coffee_raw$Bag.Weight.KG <- ifelse(
  is.na(coffee_raw$Bag.Weight.KG), median_weight, coffee_raw$Bag.Weight.KG
)

cat("Numeric variables processed\n")
cat("  Altitude missing:", sum(is.na(coffee_raw$altitude_mean_meters)), "\n")
cat("  Harvest.Year range:", min(coffee_raw$Harvest.Year.Parsed), "-", 
    max(coffee_raw$Harvest.Year.Parsed), "\n")
```

## Remove Leakage Columns

```{r remove-leakage}
coffee_raw$Total.Cup.Points <- NULL
coffee_raw$Altitude <- NULL
coffee_raw$Species <- NULL
coffee_raw$Harvest.Year <- NULL
coffee_raw$Bag.Weight <- NULL

cat("Leakage columns removed: Total.Cup.Points, Altitude, Species, Harvest.Year, Bag.Weight\n")
```

\newpage

# Exploratory Data Analysis

## Feature Descriptions

```{r feature-desc}
feature_desc <- data.frame(
  Feature = c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", "Balance",
              "Uniformity", "Clean.Cup", "Sweetness", "Moisture", 
              "Category.One.Defects", "Quakers"),
  Description = c("Smell/fragrance of the coffee",
                  "Overall taste profile",
                  "Lingering taste after swallowing",
                  "Bright, tangy quality",
                  "Weight/thickness on the tongue",
                  "How well flavors harmonize",
                  "Consistency across multiple cups",
                  "Absence of off-flavors/defects",
                  "Natural sweetness perception",
                  "Moisture content of beans (%)",
                  "Count of major defects",
                  "Unripe/underdeveloped beans count"),
  Scale = c(rep("0-10", 9), "%", "Count", "Count")
)

kable(feature_desc, caption = "Sensory Feature Descriptions", format = "markdown")
```

## Target Variable Analysis

```{r target-stats}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")

target_stats <- data.frame(
  Statistic = c("Mean", "Std Dev", "Median", "Min", "Max", "IQR"),
  Value = c(
    round(mean(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(sd(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(median(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    min(coffee_raw$Cupper.Points, na.rm = TRUE),
    max(coffee_raw$Cupper.Points, na.rm = TRUE),
    round(IQR(coffee_raw$Cupper.Points, na.rm = TRUE), 4)
  )
)
kable(target_stats, caption = "Target Variable (Cupper.Points) Summary", format = "markdown")
```

```{r target-histogram, fig.cap="Distribution of Cupper Points. Red dashed line indicates mean."}
ggplot(coffee_raw, aes(x = Cupper.Points)) +
  geom_histogram(bins = 30, fill = "#2E86AB", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(coffee_raw$Cupper.Points, na.rm = TRUE), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(x = "Cupper Points", y = "Count")
```

## Correlation Analysis

```{r correlation-table}
numeric_cols <- coffee_raw[, c(sensory_cols, "Cupper.Points")]
numeric_cols <- numeric_cols[complete.cases(numeric_cols), ]

correlations <- cor(numeric_cols)
target_corr <- correlations["Cupper.Points", ]
target_corr_sorted <- sort(target_corr[names(target_corr) != "Cupper.Points"], 
                           decreasing = TRUE)

corr_df <- data.frame(
  Feature = names(target_corr_sorted),
  Correlation = round(as.numeric(target_corr_sorted), 4)
)
kable(corr_df, caption = "Feature Correlations with Cupper.Points", format = "markdown")
```

```{r correlation-barplot, fig.cap="Feature correlations with Cupper Points."}
ggplot(corr_df, aes(x = reorder(Feature, Correlation), y = Correlation, 
                    fill = Correlation > 0)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "#E63946"), 
                    guide = "none") +
  labs(x = "Feature", y = "Correlation") +
  geom_hline(yintercept = 0, linetype = "dashed")
```

```{r correlation-heatmap, fig.cap="Correlation heatmap of sensory features and target variable.", fig.width=10, fig.height=8}
cor_matrix <- cor(numeric_cols)
cor_melted <- as.data.frame(as.table(cor_matrix))
names(cor_melted) <- c("Var1", "Var2", "Correlation")

ggplot(cor_melted, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#E63946", mid = "white", high = "#2E86AB",
                       midpoint = 0, limits = c(-1, 1)) +
  geom_text(aes(label = round(Correlation, 2)), size = 2.5) +
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\newpage

# Feature Engineering

## Feature Set Definitions

```{r feature-sets}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")

extended_cols <- c(sensory_cols, 
                   "Processing.Method", "Variety", 
                   "altitude_mean_meters", "Category.Two.Defects")

full_cols <- c(extended_cols,
               "Country.of.Origin", "Color", "Region",
               "Harvest.Year.Parsed", "Bag.Weight.KG", 
               "Number.of.Bags", "In.Country.Partner")
```

```{r feature-sets-table}
feature_sets <- data.frame(
  Feature.Set = c("Sensory", "Extended", "Full"),
  N.Features = c(length(sensory_cols), length(extended_cols), length(full_cols)),
  Description = c("Core sensory quality indicators only",
                  "Sensory + key metadata (processing, variety, altitude)",
                  "All available features including origin and logistics")
)
kable(feature_sets, caption = "Feature Set Definitions", format = "markdown")
```

## Prepare Clean Dataset

```{r prepare-clean}
coffee_clean <- coffee_raw[, c(full_cols, "Cupper.Points")]
coffee_clean <- na.omit(coffee_clean)
cat("Clean dataset:", nrow(coffee_clean), "rows\n")
```

\newpage

# Train/Test Split

```{r train-test-split}
set.seed(SEED)

n <- nrow(coffee_clean)
train_idx <- sample(1:n, floor(0.8 * n))
test_idx <- setdiff(1:n, train_idx)

y <- coffee_clean$Cupper.Points
y_train <- y[train_idx]
y_test <- y[test_idx]

# Sensory features
X_sensory <- as.matrix(coffee_clean[, sensory_cols])
X_sensory_train <- X_sensory[train_idx, ]
X_sensory_test <- X_sensory[test_idx, ]

# Extended features (one-hot encoded)
X_extended <- model.matrix(~ . - 1, data = coffee_clean[, extended_cols])
X_extended_train <- X_extended[train_idx, ]
X_extended_test <- X_extended[test_idx, ]

# Full features (one-hot encoded)
X_full <- model.matrix(~ . - 1, data = coffee_clean[, full_cols])
X_full_train <- X_full[train_idx, ]
X_full_test <- X_full[test_idx, ]

train_df <- coffee_clean[train_idx, ]
test_df <- coffee_clean[test_idx, ]

# CV folds
set.seed(SEED + 42)
fold_ids <- sample(rep(1:K_FOLDS, length.out = n))
```

```{r split-summary}
split_summary <- data.frame(
  Set = c("Training", "Test", "Total"),
  Observations = c(length(train_idx), length(test_idx), n),
  Percentage = c("80%", "20%", "100%")
)
kable(split_summary, caption = "Train/Test Split Summary", format = "markdown")
```

# Evaluation Metrics

```{r metrics-functions}
get_metrics <- function(pred, obs) {
  residuals <- obs - pred
  rmse <- sqrt(mean(residuals^2))
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  mae <- mean(abs(residuals))
  return(list(RMSE = rmse, R2 = r2, MAE = mae))
}

get_enhanced_metrics <- function(pred, obs, n_features = 12) {
  residuals <- obs - pred
  n <- length(obs)
  
  rmse <- sqrt(mean(residuals^2))
  mae <- mean(abs(residuals))
  median_ae <- median(abs(residuals))
  max_error <- max(abs(residuals))
  mape <- mean(abs(residuals / (obs + 1e-10))) * 100
  
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  adj_r2 <- 1 - (1 - r2) * (n - 1) / (n - n_features - 1)
  
  return(list(
    RMSE = rmse, MAE = mae, MAPE = mape, R2 = r2,
    Adjusted_R2 = adj_r2, Median_AE = median_ae, Max_Error = max_error
  ))
}

cohens_d <- function(x, y) {
  diff <- mean(x) - mean(y)
  pooled_sd <- sqrt((var(x) + var(y)) / 2)
  diff / pooled_sd
}
```

\newpage

# Baseline Models

## Mean Predictor

```{r baseline-mean}
mean_train <- mean(y_train)
preds_mean <- rep(mean_train, length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)
metrics_mean_enh <- get_enhanced_metrics(preds_mean, y_test, 0)

cat("Training Mean:", round(mean_train, 4), "\n")
cat("Test RMSE:", round(metrics_mean$RMSE, 4), "\n")
cat("Test R²:", round(metrics_mean$R2, 4), "\n")
```

## Linear Regression

```{r baseline-lm}
model_lm <- lm(Cupper.Points ~ ., data = train_df)
preds_lm <- predict(model_lm, newdata = test_df)
metrics_lm <- get_metrics(preds_lm, y_test)
metrics_lm_enh <- get_enhanced_metrics(preds_lm, y_test, ncol(X_full))

cat("Linear Regression Test RMSE:", round(metrics_lm$RMSE, 4), "\n")
cat("Linear Regression Test R²:", round(metrics_lm$R2, 4), "\n")
```

```{r lm-coefficients}
lm_coefs <- summary(model_lm)$coefficients
lm_coefs_df <- data.frame(
  Feature = rownames(lm_coefs),
  Estimate = round(lm_coefs[, 1], 4),
  Std_Error = round(lm_coefs[, 2], 4),
  P_Value = format(lm_coefs[, 4], digits = 3)
)
kable(head(lm_coefs_df[order(abs(lm_coefs_df$Estimate), decreasing = TRUE), ], 10), 
      caption = "Top 10 Linear Regression Coefficients",
      format = "markdown", row.names = FALSE)
```

\newpage
# XGBoost XAI 
## ALE
### XGBoost Load Model

```{r train-final}
model_xgb_f <- xgb.load("models/model_xgb_f.xgb")

dtest_f <- xgb.DMatrix(data = X_full_test, label = y_test)
preds_xgb_f <- predict(model_xgb_f, dtest_f)
```


# Explainable AI: PFI and PDP Analysis

This section applies model-agnostic XAI techniques to interpret the best XGBoost Full model. We focus on:

- **Permutation Feature Importance (PFI)**: Which features matter most?
- **Partial Dependence Plots (PDP)**: How do individual features affect predictions?
- **Accumulated Local Effects (ALE)**: Similar to PDP but more robust to multicollinearities.
- **Shapley Additive Explanations (SHAP)**: 
- **Microscopic Explanations**: Why did specific predictions occur?

## Setup: Load DALEX and Create Explainer

```{r xai-setup, message=FALSE, warning=FALSE}
# Install DALEX if not available
if (!require("DALEX")) install.packages("DALEX")
library(DALEX)

# Custom prediction function for XGBoost
# DALEX requires a function that takes (model, newdata) and returns numeric predictions
predict_xgb <- function(model, newdata) {
  # Convert to matrix if it's a data frame
  if (is.data.frame(newdata)) {
    newdata <- as.matrix(newdata)
  }
  dmatrix <- xgb.DMatrix(data = newdata)
  predict(model, dmatrix)
}

# Create DALEX explainer for XGBoost Full model
# The explainer acts as a unified interface for all XAI methods
explainer_xgb_full <- DALEX::explain(
  model = model_xgb_f,
  data = X_full_train,           # Training features (matrix)
  y = y_train,                   # Training target values
  predict_function = predict_xgb,
  label = "XGBoost Full",
  verbose = FALSE
)

cat("Explainer created successfully!\n")
cat("  Model: XGBoost Full\n")
cat("  Features:", ncol(X_full_train), "\n")
cat("  Observations:", nrow(X_full_train), "\n")
```

\newpage

## Permutation Feature Importance (PFI)

### Concept

PFI measures how much the model's performance degrades when a feature's relationship with the target is broken by randomly shuffling its values. The algorithm:

1. Calculate baseline error: $e_{orig} = L(y, f(X))$
2. For each feature $S$: shuffle its values to create $X_{S,perm}$
3. Calculate new error: $e_{S,perm} = L(y, f(X_{S,perm}))$
4. Feature importance: $FI_S = e_{S,perm} - e_{orig}$

Higher importance = larger performance drop = more important feature.

### Calculate PFI

```{r pfi-calculation, cache=TRUE}
# Custom loss function: Mean Absolute Error (more interpretable for Cupper Points)
loss_mae <- function(observed, predicted) {
  mean(abs(observed - predicted))
}

# Calculate PFI using MAE loss
# B = number of permutation repetitions for stability
set.seed(SEED)
pfi_xgb_full <- model_parts(
  explainer = explainer_xgb_full,
  loss_function = loss_mae,
  type = "difference",    # Report change in MAE (not ratio)
  B = 10                   # 10 permutation repetitions
)

cat("PFI calculation complete.\n")
```

### PFI Visualization

```{r pfi-plot, fig.cap="Permutation Feature Importance for XGBoost Full model. Longer bars indicate features whose shuffling causes larger MAE increases, meaning they are more important for prediction accuracy.", fig.width=10, fig.height=8}
# Plot PFI - shows top features by importance
plot(pfi_xgb_full, max_vars = 20) +
  ggtitle("Permutation Feature Importance (Top 20 Features)") +
  xlab("MAE Increase When Feature is Permuted") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

### PFI Results Table

```{r pfi-table}
# Extract and format PFI results
pfi_results <- pfi_xgb_full %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(
    Mean_Dropout_Loss = mean(dropout_loss),
    SD = sd(dropout_loss),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(
    Importance = Mean_Dropout_Loss - min(pfi_xgb_full$dropout_loss[pfi_xgb_full$variable == "_full_model_"]),
    Rank = row_number()
  )

# Display top 15
kable(head(pfi_results[, c("Rank", "variable", "Importance", "SD")], 15),
      col.names = c("Rank", "Feature", "MAE Increase", "SD"),
      caption = "Top 15 Features by Permutation Feature Importance",
      format = "markdown", digits = 4)
```


### PFI Interpretation

```{r pfi-interpretation}
# Identify top 5 features
top5_features <- head(pfi_results$variable, 5)
top5_importance <- head(pfi_results$Importance, 5)

cat("KEY PFI FINDINGS:\n")
cat("================\n\n")
cat("Top 5 Most Important Features (by MAE increase when permuted):\n\n")
for (i in 1:5) {
  cat(sprintf("  %d. %s: +%.4f MAE\n", i, top5_features[i], top5_importance[i]))
}
cat("\n")
cat("Interpretation: Shuffling these features causes the largest prediction errors,\n")
cat("indicating the model relies heavily on them for accurate predictions.\n")
```

\newpage

## Partial Dependence Plots (PDP)

### Concept

PDP shows the **marginal effect** of a feature on predictions, averaging over all other features:

$$\hat{f}_S(s^*) = \frac{1}{n} \sum_{i=1}^{n} f(S = s^*, \mathbf{C}_i)$$

Where:
- $s^*$ = a specific value of feature $S$
- $C_i$ = values of all other features for observation $i$
- The sum averages predictions across all observations with $S$ fixed at $s^*$

### Calculate PDPs for Top Sensory Features

```{r pdp-calculation, cache=TRUE}
# Identify top sensory features from PFI (these are most interpretable)
sensory_features_in_model <- intersect(
  sensory_cols,
  colnames(X_full_train)
)

# Get top 4 sensory features by PFI importance
top_sensory_pfi <- pfi_results %>%
  filter(variable %in% sensory_features_in_model) %>%
  head(4) %>%
  pull(variable)

cat("Calculating PDPs for top sensory features:", paste(top_sensory_pfi, collapse = ", "), "\n")

# Calculate PDPs for each top feature
pdp_list <- list()
for (feat in top_sensory_pfi) {
  pdp_list[[feat]] <- model_profile(
    explainer = explainer_xgb_full,
    variables = feat,
    type = "partial",
    N = 300  # Use 300 observations for speed (set higher for smoother curves)
  )
}

cat("PDP calculation complete.\n")
```


### Combined PDP Plot

```{r pdp-combined, fig.cap="Combined PDP showing all top sensory features on the same scale for direct comparison of effect magnitudes.", fig.width=10, fig.height=6}
# Combine all PDPs for comparison
all_pdps <- model_profile(
  explainer = explainer_xgb_full,
  variables = top_sensory_pfi,
  type = "partial",
  N = 300
)

plot(all_pdps) +
  ggtitle("Combined Partial Dependence Plots") +
  ylab("Average Predicted Cupper Points") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom")
```

### PDP for Feature Interactions (2D)

```{r pdp-2d, fig.cap="2D Partial Dependence Plot showing the joint effect of the two most important sensory features.", fig.width=8, fig.height=6, message=FALSE}
# 2D PDP for top 2 features (if both are sensory)
if (length(top_sensory_pfi) >= 2) {
  pdp_2d <- model_profile(
    explainer = explainer_xgb_full,
    variables = top_sensory_pfi[1:2],
    type = "partial",
    N = 100  # Fewer observations for 2D (computational)
  )
  
  # Extract the data for custom 2D visualization
  cat("2D PDP calculated for:", top_sensory_pfi[1], "and", top_sensory_pfi[2], "\n")
  cat("(2D visualization requires additional packages - showing individual PDPs above)\n")
}
```

### PDP Interpretation

```{r pdp-interpretation}
cat("KEY PDP FINDINGS:\n")
cat("================\n\n")

for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  min_pred <- min(pdp_data$`_yhat_`)
  max_pred <- max(pdp_data$`_yhat_`)
  range_effect <- max_pred - min_pred
  
  # Find the x values at min and max predictions
  x_at_min <- pdp_data$`_x_`[which.min(pdp_data$`_yhat_`)]
  x_at_max <- pdp_data$`_x_`[which.max(pdp_data$`_yhat_`)]
  
  cat(sprintf("%s:\n", feat))
  cat(sprintf("  - Effect range: %.4f Cupper Points\n", range_effect))
  cat(sprintf("  - Lowest avg prediction (%.3f) at %s = %.2f\n", min_pred, feat, x_at_min))
  cat(sprintf("  - Highest avg prediction (%.3f) at %s = %.2f\n", max_pred, feat, x_at_max))
  cat("\n")
}

cat("Interpretation:\n")
cat("Features with larger effect ranges have stronger marginal effects on predictions.\n")
cat("The PDP curves show the 'shape' of each relationship (linear, monotonic, complex).\n")
```

### Feature Importance from PDP (Variance-based)

```{r pdp-importance}
# Calculate importance from PDP as variance of PDP values
pdp_variance_importance <- data.frame(
  Feature = character(),
  PDP_Variance = numeric(),
  PDP_Range = numeric(),
  stringsAsFactors = FALSE
)

for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_variance_importance <- rbind(pdp_variance_importance, data.frame(
    Feature = feat,
    PDP_Variance = var(pdp_data$`_yhat_`),
    PDP_Range = max(pdp_data$`_yhat_`) - min(pdp_data$`_yhat_`)
  ))
}

pdp_variance_importance <- pdp_variance_importance %>%
  arrange(desc(PDP_Variance))

kable(pdp_variance_importance,
      col.names = c("Feature", "PDP Variance", "PDP Range"),
      caption = "Feature Importance Derived from PDP (Higher Variance = Stronger Effect)",
      format = "markdown", digits = 4)
```

\newpage

## Microscopic Explanations (Individual Predictions)

### Concept

While PFI and PDP provide **global** insights, microscopic explanations show why the model made a **specific prediction** for individual observations.

We use DALEX's **Break Down** method, which decomposes a prediction into feature contributions (note: this is NOT SHAP, but a complementary approach).

### Select Observations for Explanation

```{r micro-select-obs}
# Select interesting observations from test set
# 1. Best prediction (closest to actual)
# 2. Worst prediction (furthest from actual)
# 3. Median prediction

residuals_test <- y_test - preds_xgb_f
abs_residuals <- abs(residuals_test)

idx_best <- which.min(abs_residuals)
idx_worst <- which.max(abs_residuals)
idx_median <- which.min(abs(abs_residuals - median(abs_residuals)))

selected_obs <- data.frame(
  Type = c("Best Prediction", "Worst Prediction", "Median Prediction"),
  Test_Index = c(idx_best, idx_worst, idx_median),
  Actual = round(y_test[c(idx_best, idx_worst, idx_median)], 3),
  Predicted = round(preds_xgb_f[c(idx_best, idx_worst, idx_median)], 3),
  Residual = round(residuals_test[c(idx_best, idx_worst, idx_median)], 3)
)

kable(selected_obs, caption = "Selected Observations for Microscopic Analysis", format = "markdown")
```

### Break Down Explanations

```{r micro-breakdown, fig.cap="Break Down plots showing how each feature contributes to individual predictions. Green bars push predictions up; red bars push predictions down.", fig.width=12, fig.height=12}
# Prepare test observations for explanation
new_obs_best <- X_full_test[idx_best, , drop = FALSE]
new_obs_worst <- X_full_test[idx_worst, , drop = FALSE]
new_obs_median <- X_full_test[idx_median, , drop = FALSE]

# Calculate Break Down for each observation
bd_best <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_best,
  type = "break_down"  # NOT shap - uses sequential contribution
)

bd_worst <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_worst,
  type = "break_down"
)

bd_median <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_median,
  type = "break_down"
)

# Plot all three
p_best <- plot(bd_best, max_features = 10) +
  ggtitle(sprintf("Best Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_best], preds_xgb_f[idx_best])) +
  theme(plot.title = element_text(size = 10))

p_worst <- plot(bd_worst, max_features = 10) +
  ggtitle(sprintf("Worst Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_worst], preds_xgb_f[idx_worst])) +
  theme(plot.title = element_text(size = 10))

p_median <- plot(bd_median, max_features = 10) +
  ggtitle(sprintf("Median Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_median], preds_xgb_f[idx_median])) +
  theme(plot.title = element_text(size = 10))

grid.arrange(p_best, p_median, p_worst, ncol = 1)
```

### Microscopic Explanation Tables

```{r micro-tables}
# Extract top contributions for each observation
extract_top_contributions <- function(bd_result, n = 8) {
  bd_df <- as.data.frame(bd_result)
  bd_df <- bd_df %>%
    filter(variable_name != "intercept" & variable_name != "prediction") %>%
    arrange(desc(abs(contribution))) %>%
    head(n) %>%
    select(variable_name, variable_value, contribution) %>%
    mutate(
      contribution = round(contribution, 4),
      Direction = ifelse(contribution > 0, "↑ Positive", "↓ Negative")
    )
  return(bd_df)
}

cat("Best Prediction - Top Feature Contributions:\n")
kable(extract_top_contributions(bd_best), 
      col.names = c("Feature", "Value", "Contribution", "Direction"),
      format = "markdown")

cat("\nWorst Prediction - Top Feature Contributions:\n")
kable(extract_top_contributions(bd_worst),
      col.names = c("Feature", "Value", "Contribution", "Direction"),
      format = "markdown")
```

\newpage

## Summary: XAI Insights for XGBoost Full Model

### Key Findings from PFI

```{r summary-pfi}
cat("PERMUTATION FEATURE IMPORTANCE SUMMARY\n")
cat("======================================\n\n")

# Top 5 features
cat("Most Important Features (by MAE increase when shuffled):\n")
for (i in 1:min(5, nrow(pfi_results))) {
  cat(sprintf("  %d. %s (+%.4f MAE)\n", i, pfi_results$variable[i], pfi_results$Importance[i]))
}

# Sensory vs non-sensory
sensory_in_top10 <- sum(head(pfi_results$variable, 10) %in% sensory_cols)
cat(sprintf("\nSensory features in top 10: %d/10\n", sensory_in_top10))
cat("This confirms that expert taste evaluation drives predictions.\n")
```

### Key Findings from PDP

```{r summary-pdp}
cat("\nPARTIAL DEPENDENCE PLOT SUMMARY\n")
cat("===============================\n\n")

cat("Feature Effect Magnitudes:\n")
for (i in 1:nrow(pdp_variance_importance)) {
  cat(sprintf("  %s: %.4f point range\n", 
              pdp_variance_importance$Feature[i], 
              pdp_variance_importance$PDP_Range[i]))
}

cat("\nRelationship Shapes:\n")
cat("  - Most sensory features show positive monotonic relationships\n")
cat("  - Higher scores → higher predicted Cupper Points\n")
cat("  - This aligns with domain knowledge about coffee quality\n")
```

### Comparison: PFI vs PDP-derived Importance

```{r summary-comparison, fig.cap="Comparison of feature importance rankings from PFI and PDP variance. Consistent rankings indicate robust importance estimates.", fig.width=8, fig.height=5}
# Create comparison for features present in both analyses
comparison_methods <- pdp_variance_importance %>%
  left_join(
    pfi_results %>% 
      filter(variable %in% pdp_variance_importance$Feature) %>%
      select(variable, Importance) %>%
      rename(Feature = variable, PFI_Importance = Importance),
    by = "Feature"
  ) %>%
  mutate(
    PFI_Rank = rank(-PFI_Importance),
    PDP_Rank = rank(-PDP_Variance)
  )

# Correlation of ranks
rank_correlation <- cor(comparison_methods$PFI_Rank, comparison_methods$PDP_Rank, 
                        method = "spearman")

cat(sprintf("\nRank Correlation (Spearman) between PFI and PDP: %.3f\n", rank_correlation))

# Plot comparison
comparison_long2 <- comparison_methods %>%
  select(Feature, PFI_Rank, PDP_Rank) %>%
  pivot_longer(cols = c(PFI_Rank, PDP_Rank), 
               names_to = "Method", values_to = "Rank") %>%
  mutate(Method = ifelse(Method == "PFI_Rank", "PFI", "PDP Variance"))

ggplot(comparison_long2, aes(x = Feature, y = Rank, color = Method, group = Method)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  scale_y_reverse() +  # Rank 1 at top
  scale_color_manual(values = c("PFI" = "#E63946", "PDP Variance" = "#457B9D")) +
  labs(title = "Feature Importance Rankings: PFI vs PDP",
       y = "Rank (1 = Most Important)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Practical Implications

```{r practical-implications}
cat("\nPRACTICAL IMPLICATIONS\n")
cat("======================\n\n")

cat("1. MODEL TRUST:\n")
cat("   - PFI confirms the model uses sensible features (sensory scores)\n")
cat("   - PDP shows expected positive relationships with quality indicators\n")
cat("   - These findings align with coffee quality expertise\n\n")

cat("2. FEATURE ENGINEERING:\n")
cat("   - Focus data collection efforts on top PFI features\n")
cat("   - Low-importance features could be removed for simpler models\n\n")

cat("3. PREDICTION UNDERSTANDING:\n")
cat("   - Break Down explains individual predictions\n")
cat("   - Useful for auditing specific coffee quality assessments\n")
cat("   - Identifies which scores drove high/low predictions\n")
```



## Accumulated Local Effects (ALE)

### Concept

ALE plots show how features influence predictions by measuring the effect of small, local changes in feature values. Unlike PDP, ALE:

- **Handles correlated features correctly**: Evaluates effects along the actual data distribution
- **Shows local effects**: Accumulates changes in predictions within small intervals
- **Avoids extrapolation**: Only uses realistic feature combinations from the data

The ALE algorithm:

1. Divide feature range into intervals $[z_{k-1}, z_k]$
2. For each interval, calculate the average prediction difference when moving from $z_{k-1}$ to $z_k$
3. Accumulate these differences: $\text{ALE}(x) = \sum_{k=1}^{k(x)} \Delta_k$
4. Center the curve so the mean effect is zero

### Calculate ALE for Top Features
```{r ale-calculation, cache=TRUE}
# Select top 6 features from PFI for ALE analysis
top_features <- head(pfi_results$variable, 6)

cat("Calculating ALE for top 6 features:\n")
cat(paste("  -", top_features, collapse = "\n"), "\n")

# Calculate ALE for each top feature
set.seed(SEED)
ale_plots <- lapply(top_features, function(feature) {
  cat("  Processing:", feature, "...\n")
  model_profile(
    explainer = explainer_xgb_full,
    variables = feature,
    type = "accumulated",  # ALE method
    N = 1000               # Sample 1000 observations for efficiency
  )
})

names(ale_plots) <- top_features

# Manual correction so that ALE plots are centered around 0 which should be the default 
ale_plots <- lapply(ale_plots, function(mp) {
  mp$agr_profiles$`_yhat_` <-
    mp$agr_profiles$`_yhat_` - mean(mp$agr_profiles$`_yhat_`)
  mp
})

cat("\nALE calculation complete.\n")
```

### ALE Visualizations
```{r ale-plots, fig.cap="Accumulated Local Effects for the top 6 most important features. The y-axis shows the accumulated effect on predicted Cupper Points. Zero-centered curves indicate positive (above 0) or negative (below 0) influence on predictions.", fig.width=12, fig.height=10}
# Create a grid of ALE plots
ale_plot_list <- lapply(seq_along(ale_plots), function(i) {
  feature <- names(ale_plots)[i]
  plot(ale_plots[[i]]) +
    ggtitle(paste0(i, ". ", feature)) +
    xlab(feature) +
    ylab("Accumulated Effect") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.title = element_text(size = 10)
    )
})

# Arrange plots in a 2x3 grid
library(gridExtra)
do.call(grid.arrange, c(ale_plot_list, ncol = 2))
```

### ALE Interpretation Guide
```{r ale-interpretation, results='asis'}
cat("
**How to read ALE plots:**

- **Y-axis (Accumulated Effect)**: The impact on predicted Cupper Points
  - Values above 0: Feature values in this range increase predictions
  - Values below 0: Feature values in this range decrease predictions
  - The magnitude shows the strength of the effect

- **X-axis**: The range of the feature's values in the training data

- **Curve shape**: 
  - Steep slopes = strong local effects
  - Flat sections = feature has little effect in that range
  - Non-linear curves = complex, non-monotonic relationships

- **Zero-centering**: All curves are centered so their average effect is zero, making relative comparisons easier
")
```

### ALE Summary Statistics
```{r ale-summary}
# Extract summary statistics for ALE plots
ale_summary <- data.frame(
  Feature = character(),
  Min_Effect = numeric(),
  Max_Effect = numeric(),
  Range = numeric(),
  Mean_Abs_Effect = numeric(),
  stringsAsFactors = FALSE
)

for (feature in names(ale_plots)) {
  ale_data <- ale_plots[[feature]]$agr_profiles
  effects <- ale_data$`_yhat_`
  
  ale_summary <- rbind(ale_summary, data.frame(
    Feature = feature,
    Min_Effect = min(effects),
    Max_Effect = max(effects),
    Range = max(effects) - min(effects),
    Mean_Abs_Effect = mean(abs(effects))
  ))
}

kable(ale_summary,
      col.names = c("Feature", "Min Effect", "Max Effect", "Effect Range", "Mean |Effect|"),
      caption = "ALE Effect Summary: Range and magnitude of accumulated effects for top features",
      format = "markdown", digits = 3)
```

### Key Insights from ALE
```{r ale-insights, results='asis'}
# Identify features with largest effect ranges
largest_range <- ale_summary[which.max(ale_summary$Range), ]
most_variable <- ale_summary[which.max(ale_summary$Mean_Abs_Effect), ]

cat(sprintf("
**Notable patterns:**

1. **Strongest overall effect**: %s (effect range: %.3f Cupper Points)
   - This feature shows the largest difference between its minimum and maximum accumulated effects

2. **Most variable effect**: %s (mean absolute effect: %.3f)
   - This feature shows the most consistent deviation from zero across its range

3. **Comparison with PFI**: 
   - PFI tells us *which* features are important (breaking them hurts performance)
   - ALE tells us *how* they're important (the direction and magnitude of their effects)
   - Features can have high PFI but small ALE ranges if they're important for fine-tuning predictions
", 
largest_range$Feature, largest_range$Range,
most_variable$Feature, most_variable$Mean_Abs_Effect))
```

\newpage

## Shapley Additive Explanations (SHAP)

### Concept

SHAP values provide a unified measure of feature importance based on cooperative game theory. For each prediction, SHAP assigns each feature a value that represents its contribution to moving the prediction away from the baseline (average prediction).

Key properties of SHAP:

- **Local explanations**: Shows feature contributions for individual predictions
- **Global importance**: Aggregating SHAP values across observations reveals overall feature importance
- **Additive**: The sum of all SHAP values equals the difference between the prediction and the baseline
- **Fair attribution**: Based on Shapley values from game theory, ensuring fair credit assignment
- **Model-agnostic**: Works with any machine learning model

The intuition: SHAP answers "How much did each feature contribute to this specific prediction?"

```{r shap-microscopic}
set.seed(SEED)
# Find observation where flavor is closest to your target value - values 0.0, 10.0
target_val_flav <- 10.0

# Get the column index for flavor
flavor_col <- which(colnames(X_full_train) == "Flavor")

# Find observation closest to target value
obs_id <- which.min(abs(X_full_train[, flavor_col] - target_val_flav))

# Extract the observation as a numeric matrix row
new_obs_matrix <- X_full_train[obs_id, , drop = FALSE]

# Ensure it's numeric
storage.mode(new_obs_matrix) <- "numeric"

# Convert to data frame for predict_parts
new_obs <- as.data.frame(new_obs_matrix)

cat(paste0("Testing Observation #", obs_id, " where Flavor = ", 
           round(X_full_train[obs_id, flavor_col], 2), "\n"))

# Run SHAP for this observation
shap <- predict_parts(explainer = explainer_xgb_full,
                      new_observation = new_obs, 
                      type = "shap", B = 25)
plot(shap) + ggtitle(paste("Flavor target =", target_val_flav))
```
```{r, shap-micro-two, fig.width=12, fig.height=8}
set.seed(SEED)

# Target flavor values
target_vals <- c(0.0, 10.0)

# Get the column index for flavor
flavor_col <- which(colnames(X_full_train) == "Flavor")

shap_plots <- vector("list", length(target_vals))

for (i in seq_along(target_vals)) {
  
  target_val_flav <- target_vals[i]
  
  # Find observation closest to target value
  obs_id <- which.min(abs(X_full_train[, flavor_col] - target_val_flav))
  
  # Extract observation
  new_obs_matrix <- X_full_train[obs_id, , drop = FALSE]
  storage.mode(new_obs_matrix) <- "numeric"
  new_obs <- as.data.frame(new_obs_matrix)
  
  cat(paste0(
    "Testing Observation #", obs_id,
    " where Flavor = ",
    round(X_full_train[obs_id, flavor_col], 2), "\n"
  ))
  
  shap <- predict_parts(
    explainer = explainer_xgb_full,
    new_observation = new_obs,
    type = "shap",
    B = 25
  )
  
  shap_plots[[i]] <- plot(shap) +
    ggtitle(paste("Flavor target =", target_val_flav))
}

grid.arrange(
  grobs = shap_plots,
  ncol = 2
)

```
```{r, shapviz-plots}
set.seed(SEED)

# Target flavor values
target_vals <- c(0.0, 10.0)

# Column index for Flavor
flavor_col <- which(colnames(X_full_train) == "Flavor")

# Store plots
wf_plots <- vector("list", length(target_vals))

for (i in seq_along(target_vals)) {
  
  target_val_flav <- target_vals[i]
  
  # Find observation closest to target value
  obs_id <- which.min(abs(X_full_train[, flavor_col] - target_val_flav))
  
  # Extract observation
  new_obs_matrix <- X_full_train[obs_id, , drop = FALSE]
  storage.mode(new_obs_matrix) <- "numeric"
  #new_obs <- as.data.frame(new_obs_matrix)
  
  cat(paste0(
    "Testing Observation #", obs_id,
    " where Flavor = ",
    round(X_full_train[obs_id, flavor_col], 2), "\n"
  ))
  
  # Create shapviz object
  sv <- shapviz(
    explainer_xgb_full$model,
    X_pred = new_obs_matrix
  )
  
  # Waterfall plot
  wf_plots[[i]] <- sv_waterfall(sv, max_display = 12) +
    ggtitle(paste("Flavor target =", target_val_flav))
}

# Arrange plots side-by-side
grid.arrange(
  grobs = wf_plots,
  ncol = 2
)
```


