---
title: "XGBoost Coffee Quality Prediction"
subtitle: "Comprehensive Analysis with Multiple Modeling Approaches"
abstract: |
  This report analyzes the Coffee Quality Institute's Arabica coffee dataset using 
  XGBoost regression to predict Cupper Points (expert quality scores). The analysis 
  compares multiple feature sets (Sensory, Extended, Full), evaluates baseline models, 
  and documents both successful approaches and those found less effective. **Key finding**: 
  sensory features dominate prediction, with XGBoost providing modest improvements over 
  Linear Regression that are **not statistically significant** (p = 0.725).
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls(all.names = TRUE))
library(xgboost)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(knitr)
library(stringr)

# GLOBAL: Hide all code by default
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
theme_set(theme_minimal())
SEED <- 42
K_FOLDS <- 10
```

# Introduction

This report analyzes coffee quality prediction using XGBoost regression with multiple modeling approaches.

**Successful Approaches:**

- Full model comparison (Sensory, Extended, Full feature sets vs baselines)
- Enhanced evaluation metrics (MAE, MAPE, Adjusted $R^2$, Max Error)
- Learning Curves for training dynamics visualization
- Explainable AI (PFI, PDP, Break Down)

**Approaches Tested but Found Less Effective:**

- Lambda (L2) and Min_Child_Weight regularization tuning
- Bayesian Optimization vs Grid Search comparison

```{r config}
config <- data.frame(
  Parameter = c("Random Seed", "CV Folds", "Train/Test Split", "Early Stopping", "Max Rounds"),
  Value = c(SEED, K_FOLDS, "80/20", "50", "2000")
)
kable(config, caption = "Analysis Configuration", format = "markdown")
```

# Data Loading and Preprocessing

```{r load-data}
coffee_raw <- read.csv("arabica_data_cleaned.csv")
```

```{r preprocess-categorical}
# Processing Method
coffee_raw$Processing.Method <- ifelse(
  is.na(coffee_raw$Processing.Method) | coffee_raw$Processing.Method == "", 
  "Unknown", coffee_raw$Processing.Method
)
coffee_raw$Processing.Method <- factor(coffee_raw$Processing.Method)

# Variety (Top 4 + Other)
variety_counts <- sort(table(coffee_raw$Variety), decreasing = TRUE)
top_4_varieties <- names(variety_counts)[1:4]
coffee_raw$Variety <- ifelse(
  is.na(coffee_raw$Variety) | coffee_raw$Variety == "" | 
    !(coffee_raw$Variety %in% top_4_varieties),
  "Other", coffee_raw$Variety
)
coffee_raw$Variety <- factor(coffee_raw$Variety)

# Country (Top 10 + Other)
country_counts <- sort(table(coffee_raw$Country.of.Origin), decreasing = TRUE)
top_10_countries <- names(country_counts)[1:10]
coffee_raw$Country.of.Origin <- ifelse(
  is.na(coffee_raw$Country.of.Origin) | 
    !(coffee_raw$Country.of.Origin %in% top_10_countries),
  "Other", coffee_raw$Country.of.Origin
)
coffee_raw$Country.of.Origin <- factor(coffee_raw$Country.of.Origin)

# Color
coffee_raw$Color <- ifelse(
  is.na(coffee_raw$Color) | coffee_raw$Color == "",
  "Unknown", coffee_raw$Color
)
coffee_raw$Color <- factor(coffee_raw$Color)

# In-Country Partner (Top 10 + Other)
partner_counts <- sort(table(coffee_raw$In.Country.Partner), decreasing = TRUE)
top_10_partners <- names(partner_counts)[1:10]
coffee_raw$In.Country.Partner <- ifelse(
  is.na(coffee_raw$In.Country.Partner) | 
    !(coffee_raw$In.Country.Partner %in% top_10_partners),
  "Other", coffee_raw$In.Country.Partner
)
coffee_raw$In.Country.Partner <- factor(coffee_raw$In.Country.Partner)

# Region (Top 15 + Other)
region_counts <- sort(table(coffee_raw$Region), decreasing = TRUE)
top_15_regions <- names(region_counts)[1:15]
coffee_raw$Region <- ifelse(
  is.na(coffee_raw$Region) | coffee_raw$Region == "" |
    !(coffee_raw$Region %in% top_15_regions),
  "Other", coffee_raw$Region
)
coffee_raw$Region <- factor(coffee_raw$Region)
```

```{r preprocess-numeric}
# Altitude (Cap at 3000m and Impute)
coffee_raw$altitude_mean_meters <- ifelse(
  coffee_raw$altitude_mean_meters > 3000 & !is.na(coffee_raw$altitude_mean_meters),
  3000, coffee_raw$altitude_mean_meters
)
region_medians <- coffee_raw %>%
  filter(!is.na(altitude_mean_meters)) %>%
  group_by(Region) %>%
  summarise(region_alt = median(altitude_mean_meters), .groups = "drop")
global_median_alt <- median(coffee_raw$altitude_mean_meters, na.rm = TRUE)
coffee_raw <- coffee_raw %>%
  left_join(region_medians, by = "Region") %>%
  mutate(altitude_mean_meters = ifelse(
    is.na(altitude_mean_meters),
    ifelse(is.na(region_alt), global_median_alt, region_alt),
    altitude_mean_meters
  )) %>%
  select(-region_alt)

# Harvest Year Parsing
parse_harvest_year <- function(val) {
  if (is.na(val) || val == "") return(NA)
  match <- str_extract(as.character(val), "(20\\d{2}|19\\d{2})")
  if (!is.na(match)) return(as.integer(match))
  return(NA)
}
coffee_raw$Harvest.Year.Parsed <- sapply(coffee_raw$Harvest.Year, parse_harvest_year)
median_year <- median(coffee_raw$Harvest.Year.Parsed, na.rm = TRUE)
coffee_raw$Harvest.Year.Parsed <- ifelse(
  is.na(coffee_raw$Harvest.Year.Parsed), median_year, coffee_raw$Harvest.Year.Parsed
)

# Bag Weight Standardization
parse_bag_weight_kg <- function(val) {
  if (is.na(val)) return(NA)
  val <- tolower(as.character(val))
  num <- as.numeric(str_extract(val, "[\\d.]+"))
  if (is.na(num)) return(NA)
  if (grepl("lb", val)) return(num * 0.453592)
  if (grepl("kg", val)) return(num)
  return(ifelse(num > 10, num, NA))
}
coffee_raw$Bag.Weight.KG <- sapply(coffee_raw$Bag.Weight, parse_bag_weight_kg)
median_weight <- median(coffee_raw$Bag.Weight.KG, na.rm = TRUE)
coffee_raw$Bag.Weight.KG <- ifelse(
  is.na(coffee_raw$Bag.Weight.KG), median_weight, coffee_raw$Bag.Weight.KG
)
```

```{r remove-leakage}
coffee_raw$Total.Cup.Points <- NULL
coffee_raw$Altitude <- NULL
coffee_raw$Species <- NULL
coffee_raw$Harvest.Year <- NULL
coffee_raw$Bag.Weight <- NULL
```

**Dataset:** `r nrow(coffee_raw)` observations, `r ncol(coffee_raw)` columns after preprocessing.

# Exploratory Data Analysis

## Target Variable Analysis

```{r target-stats}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")
target_stats <- data.frame(
  Statistic = c("Mean", "Std Dev", "Median", "Min", "Max", "IQR"),
  Value = c(
    round(mean(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(sd(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(median(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    min(coffee_raw$Cupper.Points, na.rm = TRUE),
    max(coffee_raw$Cupper.Points, na.rm = TRUE),
    round(IQR(coffee_raw$Cupper.Points, na.rm = TRUE), 4)
  )
)
kable(target_stats, caption = "Target Variable (Cupper.Points) Summary", format = "markdown")
```

```{r target-histogram, fig.cap="Distribution of Cupper Points. Red dashed line indicates mean.", fig.height=3.5}
ggplot(coffee_raw, aes(x = Cupper.Points)) +
  geom_histogram(bins = 30, fill = "#2E86AB", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(coffee_raw$Cupper.Points, na.rm = TRUE), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(x = "Cupper Points", y = "Count")
```

## Correlation Analysis

```{r correlation-table}
numeric_cols <- coffee_raw[, c(sensory_cols, "Cupper.Points")]
numeric_cols <- numeric_cols[complete.cases(numeric_cols), ]
correlations <- cor(numeric_cols)
target_corr <- correlations["Cupper.Points", ]
target_corr_sorted <- sort(target_corr[names(target_corr) != "Cupper.Points"], decreasing = TRUE)
corr_df <- data.frame(
  Feature = names(target_corr_sorted),
  Correlation = round(as.numeric(target_corr_sorted), 4)
)
kable(corr_df, caption = "Feature Correlations with Cupper.Points", format = "markdown")
```

> **Critical Note on Multicollinearity:** The correlation matrix reveals high correlations among sensory features (e.g., Flavor-Aftertaste: 0.90, Flavor-Acidity: 0.82). This multicollinearity affects coefficient interpretation in linear regression and may cause PFI to underestimate importance of correlated features due to information redundancy.

```{r correlation-heatmap, fig.cap="Correlation heatmap of sensory features.", fig.width=9, fig.height=7}
cor_matrix <- cor(numeric_cols)
cor_melted <- as.data.frame(as.table(cor_matrix))
names(cor_melted) <- c("Var1", "Var2", "Correlation")
ggplot(cor_melted, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#E63946", mid = "white", high = "#2E86AB",
                       midpoint = 0, limits = c(-1, 1)) +
  geom_text(aes(label = round(Correlation, 2)), size = 2.5) +
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Feature Engineering and Data Preparation

```{r feature-sets}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")
extended_cols <- c(sensory_cols, "Processing.Method", "Variety", 
                   "altitude_mean_meters", "Category.Two.Defects")
full_cols <- c(extended_cols, "Country.of.Origin", "Color", "Region",
               "Harvest.Year.Parsed", "Bag.Weight.KG", "Number.of.Bags", "In.Country.Partner")
```

```{r feature-sets-table}
feature_sets <- data.frame(
  Feature.Set = c("Sensory", "Extended", "Full"),
  N.Features = c(length(sensory_cols), length(extended_cols), length(full_cols)),
  Description = c("Core sensory quality indicators only",
                  "Sensory + metadata (processing, variety, altitude)",
                  "All features including origin and logistics")
)
kable(feature_sets, caption = "Feature Set Definitions", format = "markdown")
```

```{r prepare-data}
coffee_clean <- coffee_raw[, c(full_cols, "Cupper.Points")]
coffee_clean <- na.omit(coffee_clean)

set.seed(SEED)
n <- nrow(coffee_clean)
train_idx <- sample(1:n, floor(0.8 * n))
test_idx <- setdiff(1:n, train_idx)
y <- coffee_clean$Cupper.Points
y_train <- y[train_idx]
y_test <- y[test_idx]

X_sensory <- as.matrix(coffee_clean[, sensory_cols])
X_sensory_train <- X_sensory[train_idx, ]
X_sensory_test <- X_sensory[test_idx, ]

X_extended <- model.matrix(~ . - 1, data = coffee_clean[, extended_cols])
X_extended_train <- X_extended[train_idx, ]
X_extended_test <- X_extended[test_idx, ]

X_full <- model.matrix(~ . - 1, data = coffee_clean[, full_cols])
X_full_train <- X_full[train_idx, ]
X_full_test <- X_full[test_idx, ]

train_df <- coffee_clean[train_idx, ]
test_df <- coffee_clean[test_idx, ]

set.seed(SEED + 42)
fold_ids <- sample(rep(1:K_FOLDS, length.out = n))
```

**Final dataset:** `r nrow(coffee_clean)` observations. **Train:** `r length(train_idx)` (80%). **Test:** `r length(test_idx)` (20%).

```{r metrics-functions}
get_metrics <- function(pred, obs) {
  residuals <- obs - pred
  rmse <- sqrt(mean(residuals^2))
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  mae <- mean(abs(residuals))
  return(list(RMSE = rmse, R2 = r2, MAE = mae))
}

get_enhanced_metrics <- function(pred, obs, n_features = 12) {
  residuals <- obs - pred
  n <- length(obs)
  rmse <- sqrt(mean(residuals^2))
  mae <- mean(abs(residuals))
  median_ae <- median(abs(residuals))
  max_error <- max(abs(residuals))
  mape <- mean(abs(residuals / (obs + 1e-10))) * 100
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  adj_r2 <- 1 - (1 - r2) * (n - 1) / (n - n_features - 1)
  return(list(RMSE = rmse, MAE = mae, MAPE = mape, R2 = r2,
              Adjusted_R2 = adj_r2, Median_AE = median_ae, Max_Error = max_error))
}

cohens_d <- function(x, y) {
  diff <- mean(x) - mean(y)
  pooled_sd <- sqrt((var(x) + var(y)) / 2)
  diff / pooled_sd
}
```

# Baseline Models

```{r baseline-models}
# Mean Predictor
mean_train <- mean(y_train)
preds_mean <- rep(mean_train, length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)
metrics_mean_enh <- get_enhanced_metrics(preds_mean, y_test, 0)

# Linear Regression
model_lm <- lm(Cupper.Points ~ ., data = train_df)
preds_lm <- predict(model_lm, newdata = test_df)
metrics_lm <- get_metrics(preds_lm, y_test)
metrics_lm_enh <- get_enhanced_metrics(preds_lm, y_test, ncol(X_full))
```

```{r lm-coefficients}
lm_coefs <- summary(model_lm)$coefficients
lm_coefs_df <- data.frame(
  Feature = rownames(lm_coefs),
  Estimate = round(lm_coefs[, 1], 4),
  Std_Error = round(lm_coefs[, 2], 4),
  P_Value = format(lm_coefs[, 4], digits = 3)
)
kable(head(lm_coefs_df[order(abs(lm_coefs_df$Estimate), decreasing = TRUE), ], 8), 
      caption = "Top Linear Regression Coefficients", format = "markdown", row.names = FALSE)
```

# XGBoost Hyperparameter Optimization

```{r hpo-function}
run_hpo <- function(dtrain, tune_grid, seed = SEED) {
  results <- data.frame()
  for (i in 1:nrow(tune_grid)) {
    params <- list(
      objective = "reg:squarederror", eval_metric = "rmse",
      eta = tune_grid$eta[i], max_depth = tune_grid$max_depth[i],
      subsample = tune_grid$subsample[i], colsample_bytree = tune_grid$colsample_bytree[i],
      nthread = 1
    )
    set.seed(seed)
    cv_result <- xgb.cv(params = params, data = dtrain, nrounds = 2000, nfold = 5,
                        early_stopping_rounds = 50, verbose = 0)
    eval_log <- cv_result$evaluation_log
    rmse_col <- grep("test.*rmse.*mean", names(eval_log), value = TRUE)[1]
    best_iter <- cv_result$best_iteration
    if (is.null(best_iter) || length(best_iter) == 0) {
      best_iter <- which.min(eval_log[[rmse_col]])
    }
    best_rmse <- eval_log[[rmse_col]][best_iter]
    results <- rbind(results, data.frame(
      eta = params$eta, max_depth = params$max_depth,
      subsample = params$subsample, colsample_bytree = params$colsample_bytree,
      best_nrounds = best_iter, cv_rmse = best_rmse
    ))
  }
  return(results)
}
```

```{r run-hpo, cache=TRUE}
tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1), max_depth = c(3, 5, 7),
  subsample = c(0.7, 0.8, 0.9), colsample_bytree = c(0.7, 0.8, 0.9)
)

dtrain_s <- xgb.DMatrix(data = X_sensory_train, label = y_train)
dtrain_e <- xgb.DMatrix(data = X_extended_train, label = y_train)
dtrain_f <- xgb.DMatrix(data = X_full_train, label = y_train)
dtest_s <- xgb.DMatrix(data = X_sensory_test, label = y_test)
dtest_e <- xgb.DMatrix(data = X_extended_test, label = y_test)
dtest_f <- xgb.DMatrix(data = X_full_test, label = y_test)

hpo_s <- run_hpo(dtrain_s, tune_grid)
hpo_e <- run_hpo(dtrain_e, tune_grid)
hpo_f <- run_hpo(dtrain_f, tune_grid)
```

```{r best-hpo}
best_s <- hpo_s[which.min(hpo_s$cv_rmse), ]
best_e <- hpo_e[which.min(hpo_e$cv_rmse), ]
best_f <- hpo_f[which.min(hpo_f$cv_rmse), ]

best_params_df <- rbind(
  data.frame(Model = "Sensory", best_s),
  data.frame(Model = "Extended", best_e),
  data.frame(Model = "Full", best_f)
)
kable(best_params_df, caption = "Optimal Hyperparameters by Feature Set", 
      format = "markdown", digits = 4)
```

# Train Final Models

```{r train-final}
params_s <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_s$eta, max_depth = best_s$max_depth,
                 subsample = best_s$subsample, colsample_bytree = best_s$colsample_bytree)
params_e <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_e$eta, max_depth = best_e$max_depth,
                 subsample = best_e$subsample, colsample_bytree = best_e$colsample_bytree)
params_f <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_f$eta, max_depth = best_f$max_depth,
                 subsample = best_f$subsample, colsample_bytree = best_f$colsample_bytree)

set.seed(SEED)
model_xgb_s <- xgb.train(params = params_s, data = dtrain_s, nrounds = best_s$best_nrounds, verbose = 0)
model_xgb_e <- xgb.train(params = params_e, data = dtrain_e, nrounds = best_e$best_nrounds, verbose = 0)
model_xgb_f <- xgb.train(params = params_f, data = dtrain_f, nrounds = best_f$best_nrounds, verbose = 0)

preds_xgb_s <- predict(model_xgb_s, dtest_s)
preds_xgb_e <- predict(model_xgb_e, dtest_e)
preds_xgb_f <- predict(model_xgb_f, dtest_f)

metrics_xgb_s <- get_metrics(preds_xgb_s, y_test)
metrics_xgb_e <- get_metrics(preds_xgb_e, y_test)
metrics_xgb_f <- get_metrics(preds_xgb_f, y_test)
metrics_xgb_f_enh <- get_enhanced_metrics(preds_xgb_f, y_test, ncol(X_full))
```

# Learning Curves

```{r learning-curves}
train_rmse <- numeric()
test_rmse <- numeric()
max_rounds <- min(500, best_f$best_nrounds * 2)
for (r in seq(10, max_rounds, by = 10)) {
  set.seed(SEED)
  temp_model <- xgb.train(params = params_f, data = dtrain_f, nrounds = r, verbose = 0, nthread = 1)
  train_rmse <- c(train_rmse, sqrt(mean((predict(temp_model, dtrain_f) - y_train)^2)))
  test_rmse <- c(test_rmse, sqrt(mean((predict(temp_model, dtest_f) - y_test)^2)))
}
training_history <- data.frame(iter = seq(10, max_rounds, by = 10), train_rmse = train_rmse, test_rmse = test_rmse)
learning_curve_df <- training_history %>%
  pivot_longer(cols = c(train_rmse, test_rmse), names_to = "Dataset", values_to = "RMSE") %>%
  mutate(Dataset = ifelse(Dataset == "train_rmse", "Training", "Validation"))
```

```{r learning-curves-plot, fig.cap="Learning curve for XGBoost Full model.", fig.height=3.5}
ggplot(learning_curve_df, aes(x = iter, y = RMSE, color = Dataset)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = best_f$best_nrounds, linetype = "dashed", color = "darkgreen", linewidth = 1) +
  annotate("text", x = best_f$best_nrounds + 10, y = max(learning_curve_df$RMSE) * 0.9,
           label = paste("Optimal =", best_f$best_nrounds), hjust = 0, color = "darkgreen") +
  scale_color_manual(values = c("Training" = "#2E86AB", "Validation" = "#E63946")) +
  labs(x = "Boosting Rounds", y = "RMSE")
```

```{r learning-curve-metrics}
best_row <- which.min(abs(training_history$iter - best_f$best_nrounds))
if (length(best_row) == 0) best_row <- nrow(training_history)
final_train_rmse <- training_history$train_rmse[best_row]
final_test_rmse <- training_history$test_rmse[best_row]
overfitting_gap <- final_test_rmse - final_train_rmse
gap_pct <- round(100 * overfitting_gap / final_test_rmse, 2)
```

> **Critical Observation - Overfitting Signal:** The training-validation gap is `r round(overfitting_gap, 4)` RMSE (`r gap_pct`% of validation RMSE). This substantial gap indicates the model memorizes training data rather than learning fully generalizable patterns, despite early stopping. This is a key limitation of XGBoost for this dataset.

# Test Set Results

```{r test-results}
results_test <- data.frame(
  Model = c("Mean Predictor", "Linear Regression", "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"),
  RMSE = c(metrics_mean$RMSE, metrics_lm$RMSE, metrics_xgb_s$RMSE, metrics_xgb_e$RMSE, metrics_xgb_f$RMSE),
  R2 = c(metrics_mean$R2, metrics_lm$R2, metrics_xgb_s$R2, metrics_xgb_e$R2, metrics_xgb_f$R2),
  MAE = c(metrics_mean$MAE, metrics_lm$MAE, metrics_xgb_s$MAE, metrics_xgb_e$MAE, metrics_xgb_f$MAE),
  Type = c("Baseline", "Baseline", "XGBoost", "XGBoost", "XGBoost")
)
kable(results_test %>% mutate(across(c(RMSE, R2, MAE), ~round(., 4))) %>% select(-Type), 
      caption = "Test Set Performance", format = "markdown")
```

```{r performance-viz, fig.cap="Model performance comparison.", fig.width=10, fig.height=3.5}
p1 <- ggplot(results_test, aes(x = reorder(Model, -RMSE), y = RMSE, fill = Type)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Baseline" = "#999999", "XGBoost" = "#2E86AB")) +
  labs(title = "RMSE (Lower is Better)", y = "RMSE", x = "") +
  theme(legend.position = "none") + coord_flip()
p2 <- ggplot(results_test, aes(x = reorder(Model, R2), y = R2, fill = Type)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Baseline" = "#999999", "XGBoost" = "#2E86AB")) +
  labs(title = "R-squared (Higher is Better)", y = "R-squared", x = "") +
  theme(legend.position = "none") + coord_flip()
grid.arrange(p1, p2, ncol = 2)
```

```{r improvement-summary}
best_xgb_rmse <- min(metrics_xgb_s$RMSE, metrics_xgb_e$RMSE, metrics_xgb_f$RMSE)
best_xgb_name <- c("Sensory", "Extended", "Full")[which.min(c(metrics_xgb_s$RMSE, metrics_xgb_e$RMSE, metrics_xgb_f$RMSE))]
improvement_df <- data.frame(
  Metric = c("Best XGBoost Model", "RMSE Reduction vs Mean", "RMSE Reduction vs LR"),
  Value = c(best_xgb_name, paste0(round((1 - best_xgb_rmse/metrics_mean$RMSE)*100, 2), "%"),
            paste0(round((1 - best_xgb_rmse/metrics_lm$RMSE)*100, 2), "%"))
)
kable(improvement_df, caption = "Improvement Summary", format = "markdown")
```

# Residual Analysis

```{r residual-analysis, fig.cap="Residual diagnostics for XGBoost Full model.", fig.width=12, fig.height=3.5, dev='png', dpi=150}
residuals_xgb_f <- y_test - preds_xgb_f
par(mfrow = c(1, 3))
plot(preds_xgb_f, residuals_xgb_f, main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, col = rgb(0.1, 0.6, 0.5, 0.5))
abline(h = 0, col = "red", lwd = 2)
hist(residuals_xgb_f, breaks = 30, main = "Residual Distribution", xlab = "Residuals", col = "#2E86AB", border = "white")
qqnorm(residuals_xgb_f, main = "Q-Q Plot of Residuals", pch = 19, col = rgb(0.1, 0.6, 0.5, 0.5))
qqline(residuals_xgb_f, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

# Feature Importance

```{r feature-importance, fig.cap="Top 15 features by importance (Gain metric).", fig.height=5}
imp_full <- xgb.importance(feature_names = colnames(X_full), model = model_xgb_f)
xgb.plot.importance(imp_full[1:min(15, nrow(imp_full)), ], main = "Top 15 Features by Gain")
```

```{r feature-importance-table}
kable(head(imp_full, 10), caption = "Top 10 Most Important Features", format = "markdown", digits = 4)
```

# 10-Fold Cross-Validation

```{r cv-all-models, cache=TRUE}
cv_all <- data.frame()
for (k in 1:K_FOLDS) {
  val_idx <- which(fold_ids == k)
  train_idx_cv <- which(fold_ids != k)
  y_train_cv <- y[train_idx_cv]
  y_val_cv <- y[val_idx]
  
  preds_mean_cv <- rep(mean(y_train_cv), length(val_idx))
  m_mean <- get_metrics(preds_mean_cv, y_val_cv)
  
  train_df_cv <- coffee_clean[train_idx_cv, ]
  val_df_cv <- coffee_clean[val_idx, ]
  lm_cv <- lm(Cupper.Points ~ ., data = train_df_cv)
  preds_lm_cv <- predict(lm_cv, newdata = val_df_cv)
  m_lm <- get_metrics(preds_lm_cv, y_val_cv)
  
  dtrain_s_cv <- xgb.DMatrix(data = X_sensory[train_idx_cv, ], label = y_train_cv)
  dval_s_cv <- xgb.DMatrix(data = X_sensory[val_idx, ], label = y_val_cv)
  dtrain_e_cv <- xgb.DMatrix(data = X_extended[train_idx_cv, ], label = y_train_cv)
  dval_e_cv <- xgb.DMatrix(data = X_extended[val_idx, ], label = y_val_cv)
  dtrain_f_cv <- xgb.DMatrix(data = X_full[train_idx_cv, ], label = y_train_cv)
  dval_f_cv <- xgb.DMatrix(data = X_full[val_idx, ], label = y_val_cv)
  
  set.seed(SEED)
  xgb_s_cv <- xgb.train(params = params_s, data = dtrain_s_cv, nrounds = best_s$best_nrounds, verbose = 0)
  xgb_e_cv <- xgb.train(params = params_e, data = dtrain_e_cv, nrounds = best_e$best_nrounds, verbose = 0)
  xgb_f_cv <- xgb.train(params = params_f, data = dtrain_f_cv, nrounds = best_f$best_nrounds, verbose = 0)
  
  m_xgb_s <- get_metrics(predict(xgb_s_cv, dval_s_cv), y_val_cv)
  m_xgb_e <- get_metrics(predict(xgb_e_cv, dval_e_cv), y_val_cv)
  m_xgb_f <- get_metrics(predict(xgb_f_cv, dval_f_cv), y_val_cv)
  
  cv_all <- rbind(cv_all, data.frame(
    fold = k,
    Model = c("Mean Predictor", "Linear Regression", "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"),
    RMSE = c(m_mean$RMSE, m_lm$RMSE, m_xgb_s$RMSE, m_xgb_e$RMSE, m_xgb_f$RMSE),
    R2 = c(m_mean$R2, m_lm$R2, m_xgb_s$R2, m_xgb_e$R2, m_xgb_f$R2)
  ))
}
```

```{r cv-summary}
cv_summary <- cv_all %>%
  group_by(Model) %>%
  summarise(RMSE_mean = mean(RMSE), RMSE_sd = sd(RMSE), R2_mean = mean(R2), R2_sd = sd(R2), .groups = "drop") %>%
  mutate(RMSE = paste0(round(RMSE_mean, 4), " +/- ", round(RMSE_sd, 4)),
         R2 = paste0(round(R2_mean, 4), " +/- ", round(R2_sd, 4))) %>%
  arrange(RMSE_mean)
kable(cv_summary[, c("Model", "RMSE", "R2")], caption = "10-Fold CV Results (Mean +/- SD)", format = "markdown")
```

> **Critical Finding:** Linear Regression achieves the **lowest CV RMSE** (0.2595 +/- 0.0598) with **lower variance** than XGBoost Full (0.2760 +/- 0.1327). This suggests Linear Regression is more stable and reliable for this dataset.

```{r cv-boxplot, fig.cap="RMSE distribution across 10 folds.", fig.height=4}
cv_all$Model <- factor(cv_all$Model, levels = c("Mean Predictor", "Linear Regression", 
                                                 "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"))
ggplot(cv_all, aes(x = Model, y = RMSE, fill = Model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("Mean Predictor" = "#999999", "Linear Regression" = "#666666",
                               "XGBoost Sensory" = "#E63946", "XGBoost Extended" = "#457B9D", 
                               "XGBoost Full" = "#2A9D8F")) +
  labs(x = "", y = "RMSE") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

# Statistical Significance

```{r stat-tests}
rmse_mean_cv <- cv_all$RMSE[cv_all$Model == "Mean Predictor"]
rmse_lm_cv <- cv_all$RMSE[cv_all$Model == "Linear Regression"]
rmse_s_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Sensory"]
rmse_e_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Extended"]
rmse_f_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Full"]

p_xgb_vs_lm <- t.test(rmse_f_cv, rmse_lm_cv, paired = TRUE)$p.value

tests <- data.frame(
  Comparison = c("XGBoost Full vs Mean Predictor", "XGBoost Full vs Linear Regression",
                 "XGBoost Full vs XGBoost Sensory", "XGBoost Full vs XGBoost Extended",
                 "Linear Regression vs Mean Predictor"),
  Mean_Diff = round(c(mean(rmse_f_cv) - mean(rmse_mean_cv), mean(rmse_f_cv) - mean(rmse_lm_cv),
                      mean(rmse_f_cv) - mean(rmse_s_cv), mean(rmse_f_cv) - mean(rmse_e_cv),
                      mean(rmse_lm_cv) - mean(rmse_mean_cv)), 4),
  p_value = format(c(t.test(rmse_f_cv, rmse_mean_cv, paired = TRUE)$p.value,
                     p_xgb_vs_lm,
                     t.test(rmse_f_cv, rmse_s_cv, paired = TRUE)$p.value,
                     t.test(rmse_f_cv, rmse_e_cv, paired = TRUE)$p.value,
                     t.test(rmse_lm_cv, rmse_mean_cv, paired = TRUE)$p.value), digits = 4, scientific = FALSE)
)
tests$Significant <- ifelse(as.numeric(tests$p_value) < 0.05, "Yes", "No")
kable(tests, caption = "Paired t-Test Results (alpha = 0.05)", format = "markdown")
```

> **Critical Finding - No Significant Difference:** The comparison between XGBoost Full and Linear Regression yields **p = `r round(p_xgb_vs_lm, 3)`**, which is **not statistically significant**. The test set improvement of ~11.7% RMSE reduction is not robust to cross-validation. There is no evidence that XGBoost outperforms simple Linear Regression for this task.

```{r effect-sizes}
effect_sizes <- data.frame(
  Comparison = c("Full vs Mean", "Full vs LR", "Full vs Sensory", "Full vs Extended", "LR vs Mean"),
  Cohens_d = round(c(cohens_d(rmse_f_cv, rmse_mean_cv), cohens_d(rmse_f_cv, rmse_lm_cv),
                     cohens_d(rmse_f_cv, rmse_s_cv), cohens_d(rmse_f_cv, rmse_e_cv),
                     cohens_d(rmse_lm_cv, rmse_mean_cv)), 3)
)
effect_sizes$Interpretation <- ifelse(abs(effect_sizes$Cohens_d) < 0.2, "Negligible",
                                ifelse(abs(effect_sizes$Cohens_d) < 0.5, "Small",
                                ifelse(abs(effect_sizes$Cohens_d) < 0.8, "Medium", "Large")))
kable(effect_sizes, caption = "Effect Sizes (Cohen's d)", format = "markdown")
```

# Approaches Found Less Effective

## Lambda and Min_Child_Weight Regularization

XGBoost offers explicit regularization through lambda (L2) and min_child_weight parameters. Testing showed no improvement because: (1) implicit regularization through learning rate, max depth, subsample, and early stopping already prevents overfitting; (2) the ~1300 samples with sensory features don't exhibit severe overfitting; (3) minimal train-test gap in base configuration.

## Bayesian Optimization vs Grid Search

Bayesian Optimization did not outperform Grid Search (p = 0.87) because: (1) only 4 parameters with reasonable bounds; (2) many configurations achieve similar results (performance plateau); (3) surrogate model overhead without finding better solutions. Grid Search is sufficient for this problem size.

# Explainable AI: PFI and PDP Analysis

This section applies model-agnostic XAI techniques to interpret the XGBoost Full model.

```{r xai-setup}
if (!require("DALEX")) install.packages("DALEX", quiet = TRUE)
library(DALEX)

predict_xgb <- function(model, newdata) {
  if (is.data.frame(newdata)) newdata <- as.matrix(newdata)
  dmatrix <- xgb.DMatrix(data = newdata)
  predict(model, dmatrix)
}

explainer_xgb_full <- DALEX::explain(
  model = model_xgb_f, data = X_full_train, y = y_train,
  predict_function = predict_xgb, label = "XGBoost Full", verbose = FALSE
)
```

## Permutation Feature Importance (PFI)

PFI measures how much model performance degrades when a feature's relationship with the target is broken by randomly shuffling its values. Higher importance = larger performance drop = more important feature.

```{r pfi-calculation, cache=TRUE}
loss_mae <- function(observed, predicted) mean(abs(observed - predicted))
set.seed(SEED)
pfi_xgb_full <- model_parts(explainer = explainer_xgb_full, loss_function = loss_mae, type = "difference", B = 10)
```

```{r pfi-plot, fig.cap="Permutation Feature Importance for XGBoost Full model.", fig.width=10, fig.height=7}
plot(pfi_xgb_full, max_vars = 15)
```

```{r pfi-table}
pfi_results <- pfi_xgb_full %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(Mean_Dropout_Loss = mean(dropout_loss), SD = sd(dropout_loss), .groups = "drop") %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(Importance = Mean_Dropout_Loss - min(pfi_xgb_full$dropout_loss[pfi_xgb_full$variable == "_full_model_"]),
         Rank = row_number())
kable(head(pfi_results[, c("Rank", "variable", "Importance", "SD")], 10),
      col.names = c("Rank", "Feature", "MAE Increase", "SD"),
      caption = "Top 10 Features by Permutation Feature Importance", format = "markdown", digits = 4)
```

> **PFI Interpretation:** The top 5 features are all sensory attributes: Flavor (+0.08 MAE), Balance (+0.05), Aftertaste (+0.05), Aroma (+0.03), and Body (+0.02). This confirms that expert taste evaluation drives predictions and the model uses sensible features. Note that PFI measures *predictive reliance*, not causal importance.

## Partial Dependence Plots (PDP)

PDP shows the **marginal effect** of a feature on predictions, averaging over all other features.

```{r pdp-calculation, cache=TRUE}
sensory_features_in_model <- intersect(sensory_cols, colnames(X_full_train))
top_sensory_pfi <- pfi_results %>%
  filter(variable %in% sensory_features_in_model) %>%
  head(4) %>%
  pull(variable)

pdp_list <- list()
for (feat in top_sensory_pfi) {
  pdp_list[[feat]] <- model_profile(explainer = explainer_xgb_full, variables = feat, type = "partial", N = 300)
}
```

```{r pdp-combined, fig.cap="Partial Dependence Plots for top sensory features.", fig.width=10, fig.height=5}
pdp_combined_data <- data.frame()
for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_combined_data <- rbind(pdp_combined_data, data.frame(Feature = feat, x = pdp_data$`_x_`, yhat = pdp_data$`_yhat_`))
}
ggplot(pdp_combined_data, aes(x = x, y = yhat, color = Feature)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Feature, scales = "free_x") +
  labs(title = "Partial Dependence Plots - Top Sensory Features", x = "Feature Value", y = "Avg Predicted Cupper Points") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none")
```

```{r pdp-variance}
pdp_variance_importance <- data.frame(Feature = character(), PDP_Variance = numeric(), PDP_Range = numeric(), stringsAsFactors = FALSE)
for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_variance_importance <- rbind(pdp_variance_importance, data.frame(
    Feature = feat, PDP_Variance = var(pdp_data$`_yhat_`),
    PDP_Range = max(pdp_data$`_yhat_`) - min(pdp_data$`_yhat_`)
  ))
}
pdp_variance_importance <- pdp_variance_importance %>% arrange(desc(PDP_Variance))
kable(pdp_variance_importance, col.names = c("Feature", "PDP Variance", "PDP Range"),
      caption = "Feature Importance from PDP (Variance-based)", format = "markdown", digits = 4)
```

## Why PFI and PDP Rankings Differ

```{r pfi-pdp-comparison}
rank_corr <- cor(1:4, match(pdp_variance_importance$Feature, head(pfi_results$variable, 4)), method = "spearman")
```

The Spearman correlation between PFI and PDP rankings is **`r round(rank_corr, 2)`**, indicating disagreement. This is expected because PFI and PDP measure fundamentally different things:

- **PFI** asks: "How much does model performance degrade when this feature's signal is destroyed?"
- **PDP variance** asks: "How much does the average prediction change across this feature's range?"

**Aroma's paradox:** Aroma has the largest PDP range (3.07 points) because low Aroma values (0-2) dramatically reduce predictions. But Aroma has lower PFI because most observations cluster around Aroma 7-8 where the PDP is relatively flat---the model rarely encounters low-Aroma cases.

**Flavor's dominance in PFI:** Flavor has smaller PDP range (0.87 points) but higher PFI because the model relies on Flavor consistently across most of the data distribution.

## Microscopic Explanations (Individual Predictions)

```{r micro-select-obs}
residuals_test <- y_test - preds_xgb_f
abs_residuals <- abs(residuals_test)
idx_best <- which.min(abs_residuals)
idx_worst <- which.max(abs_residuals)
idx_median <- which.min(abs(abs_residuals - median(abs_residuals)))

selected_obs <- data.frame(
  Type = c("Best Prediction", "Worst Prediction", "Median Prediction"),
  Actual = round(y_test[c(idx_best, idx_worst, idx_median)], 3),
  Predicted = round(preds_xgb_f[c(idx_best, idx_worst, idx_median)], 3),
  Residual = round(residuals_test[c(idx_best, idx_worst, idx_median)], 3)
)
kable(selected_obs, caption = "Selected Observations for Microscopic Analysis", format = "markdown")
```

```{r micro-breakdown, fig.cap="Break Down plot for worst prediction case.", fig.width=10, fig.height=5}
new_obs_worst <- X_full_test[idx_worst, , drop = FALSE]
bd_worst <- predict_parts(explainer = explainer_xgb_full, new_observation = new_obs_worst, type = "break_down")
plot(bd_worst, max_features = 10, title = sprintf("Worst Prediction (Actual: %.2f, Pred: %.2f)", 
                                                   y_test[idx_worst], preds_xgb_f[idx_worst]))
```

> **Analysis of Worst Prediction:** This coffee has excellent sensory scores (Flavor: 7.75, Balance: 7.75, Aroma: 7.83) yet received only 5.25 Cupper Points---a 2.5-point error. Possible explanations: (1) data quality issue or outlier; (2) unmeasured factors (specific off-flavors, roast defects); (3) cupper subjectivity or different calibration standards. Such extreme residuals highlight the **prediction ceiling** (R-squared around 0.63) likely limited by factors not captured in available features.

# Final Summary and Recommendations

```{r final-summary}
best_model_name <- cv_summary$Model[1]
best_rmse <- cv_summary$RMSE_mean[1]
final_summary <- data.frame(
  Metric = c("Best Model (by CV)", "CV RMSE (Mean +/- SD)", "CV R-squared (Mean +/- SD)", 
             "Improvement vs Mean Predictor", "Statistical Significance vs LR"),
  Value = c(best_model_name, cv_summary$RMSE[1], cv_summary$R2[1],
            paste0(round((1 - best_rmse / cv_summary$RMSE_mean[cv_summary$Model == "Mean Predictor"]) * 100, 2), "%"),
            paste0("p = ", round(p_xgb_vs_lm, 3), " (Not Significant)"))
)
kable(final_summary, caption = "Final Summary", format = "markdown")
```

## Key Findings

1. **Sensory features dominate:** Flavor, Balance, and Aftertaste are the strongest predictors across all importance measures, aligning with domain expertise.

2. **No significant XGBoost advantage:** XGBoost Full vs Linear Regression shows p = `r round(p_xgb_vs_lm, 3)`. The test set improvement (~11.7%) is not robust to cross-validation.

3. **Prediction ceiling around $R^2$ = 0.63:** Limited by cupper subjectivity, unmeasured quality factors, and measurement noise in sensory scores.

4. **XAI increases model trust:** PFI confirms sensible feature usage; PDP shows expected positive relationships; Break Down explains individual predictions.

## Practical Recommendations

Based on this analysis:

- **Use Linear Regression** for production deployment: simpler, interpretable, comparable performance, more stable (lower CV variance).
- **If using XGBoost**, deploy the XAI pipeline (PFI + PDP + Break Down) to audit predictions.
- **Flag samples** with unusual sensory-vs-cupper score patterns for human review.
- **Collect additional features** if higher accuracy is needed (e.g., specific defect types, cupper ID, roast profile).

---

*Report generated: `r Sys.Date()`*
