---
title: "XGBoost Coffee Quality Prediction"
subtitle: "Comprehensive Analysis with Multiple Modeling Approaches"
abstract: |
  This report analyzes the Coffee Quality Institute's Arabica coffee dataset using 
  XGBoost regression to predict Cupper Points (expert quality scores). The analysis 
  compares multiple feature sets (Sensory, Extended, Full), evaluates baseline models, 
  and documents both successful approaches and those found less effective (Lambda/Min_Child_Weight 
  with XGBoost providing modest but statistically significant improvements over Linear Regression.
  regularization, Bayesian Optimization). Key finding: sensory features dominate prediction, 
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    number_sections: true
    df_print: paged
  word_document:
    toc: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# CLEAR ENVIRONMENT
rm(list = ls(all.names = TRUE))

# LOAD PACKAGES
library(xgboost)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(knitr)
library(stringr)

# GLOBAL CHUNK OPTIONS
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

# GGPLOT THEME
theme_set(theme_minimal())

# GLOBAL CONSTANTS
SEED <- 42
K_FOLDS <- 10
```

\newpage

# Introduction

This comprehensive report analyzes coffee quality prediction using XGBoost regression. The analysis incorporates multiple modeling approaches tested across several experiments.

**Successful Approaches:**

- Full model comparison (Sensory, Extended, Full feature sets vs baselines)
- Enhanced evaluation metrics (MAE, MAPE, Adjusted R², Max Error)
- Learning Curves for training dynamics visualization
- Exploratory Data Analysis with Correlation Analysis
- Predicted vs Actual diagnostic plots

**Approaches Tested but Found Less Effective:**

- Lambda (L2) and Min_Child_Weight regularization tuning
- Bayesian Optimization vs Grid Search comparison

## Configuration

```{r config, echo=FALSE}
config <- data.frame(
  Parameter = c("Random Seed", "Cross-Validation Folds", "Train/Test Split", 
                "Early Stopping Rounds", "Max Boosting Rounds"),
  Value = c(SEED, K_FOLDS, "80/20", "50", "2000")
)

kable(config, caption = "Analysis Configuration", format = "markdown")
```

\newpage

# Data Loading and Preprocessing

## Load Data

```{r load-data}
coffee_raw <- read.csv("arabica_data_cleaned.csv")
cat("Raw data:", nrow(coffee_raw), "rows,", ncol(coffee_raw), "columns\n")
```

## Clean Categorical Variables

```{r preprocess-categorical}
# Processing Method
coffee_raw$Processing.Method <- ifelse(
  is.na(coffee_raw$Processing.Method) | coffee_raw$Processing.Method == "", 
  "Unknown", coffee_raw$Processing.Method
)
coffee_raw$Processing.Method <- factor(coffee_raw$Processing.Method)

# Variety (Top 4 + Other)
variety_counts <- sort(table(coffee_raw$Variety), decreasing = TRUE)
top_4_varieties <- names(variety_counts)[1:4]
coffee_raw$Variety <- ifelse(
  is.na(coffee_raw$Variety) | coffee_raw$Variety == "" | 
    !(coffee_raw$Variety %in% top_4_varieties),
  "Other", coffee_raw$Variety
)
coffee_raw$Variety <- factor(coffee_raw$Variety)

# Country (Top 10 + Other)
country_counts <- sort(table(coffee_raw$Country.of.Origin), decreasing = TRUE)
top_10_countries <- names(country_counts)[1:10]
coffee_raw$Country.of.Origin <- ifelse(
  is.na(coffee_raw$Country.of.Origin) | 
    !(coffee_raw$Country.of.Origin %in% top_10_countries),
  "Other", coffee_raw$Country.of.Origin
)
coffee_raw$Country.of.Origin <- factor(coffee_raw$Country.of.Origin)

# Color
coffee_raw$Color <- ifelse(
  is.na(coffee_raw$Color) | coffee_raw$Color == "",
  "Unknown", coffee_raw$Color
)
coffee_raw$Color <- factor(coffee_raw$Color)

# In-Country Partner (Top 10 + Other)
partner_counts <- sort(table(coffee_raw$In.Country.Partner), decreasing = TRUE)
top_10_partners <- names(partner_counts)[1:10]
coffee_raw$In.Country.Partner <- ifelse(
  is.na(coffee_raw$In.Country.Partner) | 
    !(coffee_raw$In.Country.Partner %in% top_10_partners),
  "Other", coffee_raw$In.Country.Partner
)
coffee_raw$In.Country.Partner <- factor(coffee_raw$In.Country.Partner)

# Region (Top 15 + Other)
region_counts <- sort(table(coffee_raw$Region), decreasing = TRUE)
top_15_regions <- names(region_counts)[1:15]
coffee_raw$Region <- ifelse(
  is.na(coffee_raw$Region) | coffee_raw$Region == "" |
    !(coffee_raw$Region %in% top_15_regions),
  "Other", coffee_raw$Region
)
coffee_raw$Region <- factor(coffee_raw$Region)

cat("Categorical variables processed\n")
cat("  Processing.Method:", nlevels(coffee_raw$Processing.Method), "levels\n")
cat("  Variety:", nlevels(coffee_raw$Variety), "levels\n")
cat("  Country:", nlevels(coffee_raw$Country.of.Origin), "levels\n")
cat("  Color:", nlevels(coffee_raw$Color), "levels\n")
```

## Clean Numeric Variables

```{r preprocess-numeric}
# Altitude (Cap at 3000m and Impute)
coffee_raw$altitude_mean_meters <- ifelse(
  coffee_raw$altitude_mean_meters > 3000 & !is.na(coffee_raw$altitude_mean_meters),
  3000, coffee_raw$altitude_mean_meters
)

region_medians <- coffee_raw %>%
  filter(!is.na(altitude_mean_meters)) %>%
  group_by(Region) %>%
  summarise(region_alt = median(altitude_mean_meters), .groups = "drop")

global_median_alt <- median(coffee_raw$altitude_mean_meters, na.rm = TRUE)

coffee_raw <- coffee_raw %>%
  left_join(region_medians, by = "Region") %>%
  mutate(altitude_mean_meters = ifelse(
    is.na(altitude_mean_meters),
    ifelse(is.na(region_alt), global_median_alt, region_alt),
    altitude_mean_meters
  )) %>%
  select(-region_alt)

# Harvest Year Parsing
parse_harvest_year <- function(val) {
  if (is.na(val) || val == "") return(NA)
  match <- str_extract(as.character(val), "(20\\d{2}|19\\d{2})")
  if (!is.na(match)) return(as.integer(match))
  return(NA)
}

coffee_raw$Harvest.Year.Parsed <- sapply(coffee_raw$Harvest.Year, parse_harvest_year)
median_year <- median(coffee_raw$Harvest.Year.Parsed, na.rm = TRUE)
coffee_raw$Harvest.Year.Parsed <- ifelse(
  is.na(coffee_raw$Harvest.Year.Parsed), median_year, coffee_raw$Harvest.Year.Parsed
)

# Bag Weight Standardization
parse_bag_weight_kg <- function(val) {
  if (is.na(val)) return(NA)
  val <- tolower(as.character(val))
  num <- as.numeric(str_extract(val, "[\\d.]+"))
  if (is.na(num)) return(NA)
  if (grepl("lb", val)) return(num * 0.453592)
  if (grepl("kg", val)) return(num)
  return(ifelse(num > 10, num, NA))
}

coffee_raw$Bag.Weight.KG <- sapply(coffee_raw$Bag.Weight, parse_bag_weight_kg)
median_weight <- median(coffee_raw$Bag.Weight.KG, na.rm = TRUE)
coffee_raw$Bag.Weight.KG <- ifelse(
  is.na(coffee_raw$Bag.Weight.KG), median_weight, coffee_raw$Bag.Weight.KG
)

cat("Numeric variables processed\n")
cat("  Altitude missing:", sum(is.na(coffee_raw$altitude_mean_meters)), "\n")
cat("  Harvest.Year range:", min(coffee_raw$Harvest.Year.Parsed), "-", 
    max(coffee_raw$Harvest.Year.Parsed), "\n")
```

## Remove Leakage Columns

```{r remove-leakage}
coffee_raw$Total.Cup.Points <- NULL
coffee_raw$Altitude <- NULL
coffee_raw$Species <- NULL
coffee_raw$Harvest.Year <- NULL
coffee_raw$Bag.Weight <- NULL

cat("Leakage columns removed: Total.Cup.Points, Altitude, Species, Harvest.Year, Bag.Weight\n")
```

\newpage

# Exploratory Data Analysis

## Feature Descriptions

```{r feature-desc, echo=FALSE}
feature_desc <- data.frame(
  Feature = c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", "Balance",
              "Uniformity", "Clean.Cup", "Sweetness", "Moisture", 
              "Category.One.Defects", "Quakers"),
  Description = c("Smell/fragrance of the coffee",
                  "Overall taste profile",
                  "Lingering taste after swallowing",
                  "Bright, tangy quality",
                  "Weight/thickness on the tongue",
                  "How well flavors harmonize",
                  "Consistency across multiple cups",
                  "Absence of off-flavors/defects",
                  "Natural sweetness perception",
                  "Moisture content of beans (%)",
                  "Count of major defects",
                  "Unripe/underdeveloped beans count"),
  Scale = c(rep("0-10", 9), "%", "Count", "Count")
)

kable(feature_desc, caption = "Sensory Feature Descriptions", format = "markdown")
```

## Target Variable Analysis

```{r target-stats, echo=FALSE}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")

target_stats <- data.frame(
  Statistic = c("Mean", "Std Dev", "Median", "Min", "Max", "IQR"),
  Value = c(
    round(mean(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(sd(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    round(median(coffee_raw$Cupper.Points, na.rm = TRUE), 4),
    min(coffee_raw$Cupper.Points, na.rm = TRUE),
    max(coffee_raw$Cupper.Points, na.rm = TRUE),
    round(IQR(coffee_raw$Cupper.Points, na.rm = TRUE), 4)
  )
)
kable(target_stats, caption = "Target Variable (Cupper.Points) Summary", format = "markdown")
```

```{r target-histogram, echo=FALSE, fig.cap="Distribution of Cupper Points. Red dashed line indicates mean."}
ggplot(coffee_raw, aes(x = Cupper.Points)) +
  geom_histogram(bins = 30, fill = "#2E86AB", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(coffee_raw$Cupper.Points, na.rm = TRUE), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(x = "Cupper Points", y = "Count")
```

## Correlation Analysis

```{r correlation-table, echo=FALSE}
numeric_cols <- coffee_raw[, c(sensory_cols, "Cupper.Points")]
numeric_cols <- numeric_cols[complete.cases(numeric_cols), ]

correlations <- cor(numeric_cols)
target_corr <- correlations["Cupper.Points", ]
target_corr_sorted <- sort(target_corr[names(target_corr) != "Cupper.Points"], 
                           decreasing = TRUE)

corr_df <- data.frame(
  Feature = names(target_corr_sorted),
  Correlation = round(as.numeric(target_corr_sorted), 4)
)
kable(corr_df, caption = "Feature Correlations with Cupper.Points", format = "markdown")
```

```{r correlation-barplot, echo=FALSE, fig.cap="Feature correlations with Cupper Points."}
ggplot(corr_df, aes(x = reorder(Feature, Correlation), y = Correlation, 
                    fill = Correlation > 0)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2E86AB", "FALSE" = "#E63946"), 
                    guide = "none") +
  labs(x = "Feature", y = "Correlation") +
  geom_hline(yintercept = 0, linetype = "dashed")
```

```{r correlation-heatmap, echo=FALSE, fig.cap="Correlation heatmap of sensory features and target variable.", fig.width=10, fig.height=8}
cor_matrix <- cor(numeric_cols)
cor_melted <- as.data.frame(as.table(cor_matrix))
names(cor_melted) <- c("Var1", "Var2", "Correlation")

ggplot(cor_melted, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#E63946", mid = "white", high = "#2E86AB",
                       midpoint = 0, limits = c(-1, 1)) +
  geom_text(aes(label = round(Correlation, 2)), size = 2.5) +
  labs(x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\newpage

# Feature Engineering

## Feature Set Definitions

```{r feature-sets}
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")

extended_cols <- c(sensory_cols, 
                   "Processing.Method", "Variety", 
                   "altitude_mean_meters", "Category.Two.Defects")

full_cols <- c(extended_cols,
               "Country.of.Origin", "Color", "Region",
               "Harvest.Year.Parsed", "Bag.Weight.KG", 
               "Number.of.Bags", "In.Country.Partner")
```

```{r feature-sets-table, echo=FALSE}
feature_sets <- data.frame(
  Feature.Set = c("Sensory", "Extended", "Full"),
  N.Features = c(length(sensory_cols), length(extended_cols), length(full_cols)),
  Description = c("Core sensory quality indicators only",
                  "Sensory + key metadata (processing, variety, altitude)",
                  "All available features including origin and logistics")
)
kable(feature_sets, caption = "Feature Set Definitions", format = "markdown")
```

## Prepare Clean Dataset

```{r prepare-clean}
coffee_clean <- coffee_raw[, c(full_cols, "Cupper.Points")]
coffee_clean <- na.omit(coffee_clean)
cat("Clean dataset:", nrow(coffee_clean), "rows\n")
```

\newpage

# Train/Test Split

```{r train-test-split}
set.seed(SEED)

n <- nrow(coffee_clean)
train_idx <- sample(1:n, floor(0.8 * n))
test_idx <- setdiff(1:n, train_idx)

y <- coffee_clean$Cupper.Points
y_train <- y[train_idx]
y_test <- y[test_idx]

# Sensory features
X_sensory <- as.matrix(coffee_clean[, sensory_cols])
X_sensory_train <- X_sensory[train_idx, ]
X_sensory_test <- X_sensory[test_idx, ]

# Extended features (one-hot encoded)
X_extended <- model.matrix(~ . - 1, data = coffee_clean[, extended_cols])
X_extended_train <- X_extended[train_idx, ]
X_extended_test <- X_extended[test_idx, ]

# Full features (one-hot encoded)
X_full <- model.matrix(~ . - 1, data = coffee_clean[, full_cols])
X_full_train <- X_full[train_idx, ]
X_full_test <- X_full[test_idx, ]

train_df <- coffee_clean[train_idx, ]
test_df <- coffee_clean[test_idx, ]

# CV folds
set.seed(SEED + 42)
fold_ids <- sample(rep(1:K_FOLDS, length.out = n))
```

```{r split-summary, echo=FALSE}
split_summary <- data.frame(
  Set = c("Training", "Test", "Total"),
  Observations = c(length(train_idx), length(test_idx), n),
  Percentage = c("80%", "20%", "100%")
)
kable(split_summary, caption = "Train/Test Split Summary", format = "markdown")
```

# Evaluation Metrics

```{r metrics-functions}
get_metrics <- function(pred, obs) {
  residuals <- obs - pred
  rmse <- sqrt(mean(residuals^2))
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  mae <- mean(abs(residuals))
  return(list(RMSE = rmse, R2 = r2, MAE = mae))
}

get_enhanced_metrics <- function(pred, obs, n_features = 12) {
  residuals <- obs - pred
  n <- length(obs)
  
  rmse <- sqrt(mean(residuals^2))
  mae <- mean(abs(residuals))
  median_ae <- median(abs(residuals))
  max_error <- max(abs(residuals))
  mape <- mean(abs(residuals / (obs + 1e-10))) * 100
  
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  adj_r2 <- 1 - (1 - r2) * (n - 1) / (n - n_features - 1)
  
  return(list(
    RMSE = rmse, MAE = mae, MAPE = mape, R2 = r2,
    Adjusted_R2 = adj_r2, Median_AE = median_ae, Max_Error = max_error
  ))
}

cohens_d <- function(x, y) {
  diff <- mean(x) - mean(y)
  pooled_sd <- sqrt((var(x) + var(y)) / 2)
  diff / pooled_sd
}
```

\newpage

# Baseline Models

## Mean Predictor

```{r baseline-mean}
mean_train <- mean(y_train)
preds_mean <- rep(mean_train, length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)
metrics_mean_enh <- get_enhanced_metrics(preds_mean, y_test, 0)

cat("Training Mean:", round(mean_train, 4), "\n")
cat("Test RMSE:", round(metrics_mean$RMSE, 4), "\n")
cat("Test R²:", round(metrics_mean$R2, 4), "\n")
```

## Linear Regression

```{r baseline-lm}
model_lm <- lm(Cupper.Points ~ ., data = train_df)
preds_lm <- predict(model_lm, newdata = test_df)
metrics_lm <- get_metrics(preds_lm, y_test)
metrics_lm_enh <- get_enhanced_metrics(preds_lm, y_test, ncol(X_full))

cat("Linear Regression Test RMSE:", round(metrics_lm$RMSE, 4), "\n")
cat("Linear Regression Test R²:", round(metrics_lm$R2, 4), "\n")
```

```{r lm-coefficients, echo=FALSE}
lm_coefs <- summary(model_lm)$coefficients
lm_coefs_df <- data.frame(
  Feature = rownames(lm_coefs),
  Estimate = round(lm_coefs[, 1], 4),
  Std_Error = round(lm_coefs[, 2], 4),
  P_Value = format(lm_coefs[, 4], digits = 3)
)
kable(head(lm_coefs_df[order(abs(lm_coefs_df$Estimate), decreasing = TRUE), ], 10), 
      caption = "Top 10 Linear Regression Coefficients",
      format = "markdown", row.names = FALSE)
```

\newpage

# XGBoost Hyperparameter Optimization

## Grid Search Setup

```{r hpo-setup}
tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(3, 5, 7),
  subsample = c(0.7, 0.8, 0.9),
  colsample_bytree = c(0.7, 0.8, 0.9)
)

cat("Grid Search: Testing", nrow(tune_grid), "hyperparameter combinations\n")
```

```{r hpo-function, echo=FALSE}
run_hpo <- function(dtrain, tune_grid, seed = SEED) {
  results <- data.frame()
  
  for (i in 1:nrow(tune_grid)) {
    params <- list(
      objective = "reg:squarederror",
      eval_metric = "rmse",
      eta = tune_grid$eta[i],
      max_depth = tune_grid$max_depth[i],
      subsample = tune_grid$subsample[i],
      colsample_bytree = tune_grid$colsample_bytree[i]
    )
    
    set.seed(seed)
    cv_result <- xgb.cv(
      params = params, data = dtrain,
      nrounds = 2000, nfold = 5,
      early_stopping_rounds = 50, verbose = 0
    )
    
    results <- rbind(results, data.frame(
      eta = params$eta,
      max_depth = params$max_depth,
      subsample = params$subsample,
      colsample_bytree = params$colsample_bytree,
      best_nrounds = cv_result$best_iteration,
      cv_rmse = cv_result$evaluation_log$test_rmse_mean[cv_result$best_iteration]
    ))
  }
  return(results)
}
```

## Run HPO

```{r run-hpo, cache=TRUE}
dtrain_s <- xgb.DMatrix(data = X_sensory_train, label = y_train)
dtrain_e <- xgb.DMatrix(data = X_extended_train, label = y_train)
dtrain_f <- xgb.DMatrix(data = X_full_train, label = y_train)

dtest_s <- xgb.DMatrix(data = X_sensory_test, label = y_test)
dtest_e <- xgb.DMatrix(data = X_extended_test, label = y_test)
dtest_f <- xgb.DMatrix(data = X_full_test, label = y_test)

cat("Running HPO for Sensory model...\n")
hpo_s <- run_hpo(dtrain_s, tune_grid)

cat("Running HPO for Extended model...\n")
hpo_e <- run_hpo(dtrain_e, tune_grid)

cat("Running HPO for Full model...\n")
hpo_f <- run_hpo(dtrain_f, tune_grid)
```

## Optimal Hyperparameters

```{r best-hpo, echo=FALSE}
best_s <- hpo_s[which.min(hpo_s$cv_rmse), ]
best_e <- hpo_e[which.min(hpo_e$cv_rmse), ]
best_f <- hpo_f[which.min(hpo_f$cv_rmse), ]

best_params_df <- rbind(
  data.frame(Model = "Sensory", best_s),
  data.frame(Model = "Extended", best_e),
  data.frame(Model = "Full", best_f)
)

kable(best_params_df, caption = "Optimal Hyperparameters by Feature Set", 
      format = "markdown", digits = 4)
```

\newpage

# Train Final Models

```{r train-final}
params_s <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_s$eta, max_depth = best_s$max_depth,
                 subsample = best_s$subsample, colsample_bytree = best_s$colsample_bytree)

params_e <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_e$eta, max_depth = best_e$max_depth,
                 subsample = best_e$subsample, colsample_bytree = best_e$colsample_bytree)

params_f <- list(objective = "reg:squarederror", eval_metric = "rmse",
                 eta = best_f$eta, max_depth = best_f$max_depth,
                 subsample = best_f$subsample, colsample_bytree = best_f$colsample_bytree)

set.seed(SEED)
model_xgb_s <- xgb.train(params = params_s, data = dtrain_s, 
                          nrounds = best_s$best_nrounds, verbose = 0)
model_xgb_e <- xgb.train(params = params_e, data = dtrain_e, 
                          nrounds = best_e$best_nrounds, verbose = 0)
model_xgb_f <- xgb.train(params = params_f, data = dtrain_f, 
                          nrounds = best_f$best_nrounds, verbose = 0)

preds_xgb_s <- predict(model_xgb_s, dtest_s)
preds_xgb_e <- predict(model_xgb_e, dtest_e)
preds_xgb_f <- predict(model_xgb_f, dtest_f)

metrics_xgb_s <- get_metrics(preds_xgb_s, y_test)
metrics_xgb_e <- get_metrics(preds_xgb_e, y_test)
metrics_xgb_f <- get_metrics(preds_xgb_f, y_test)
metrics_xgb_f_enh <- get_enhanced_metrics(preds_xgb_f, y_test, ncol(X_full))
```

# Learning Curves

```{r learning-curves, echo=FALSE}
watchlist <- list(train = dtrain_f, test = dtest_f)

set.seed(SEED)
model_verbose <- xgb.train(
  params = params_f, data = dtrain_f,
  nrounds = min(500, best_f$best_nrounds * 2),
  watchlist = watchlist, verbose = 0
)

training_history <- model_verbose$evaluation_log

learning_curve_df <- training_history %>%
  pivot_longer(cols = c(train_rmse, test_rmse),
               names_to = "Dataset", values_to = "RMSE") %>%
  mutate(Dataset = ifelse(Dataset == "train_rmse", "Training", "Validation"))
```

```{r learning-curves-plot, echo=FALSE, fig.cap="Learning curve for XGBoost Full model. Vertical line indicates optimal stopping point."}
ggplot(learning_curve_df, aes(x = iter, y = RMSE, color = Dataset)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = best_f$best_nrounds, linetype = "dashed", 
             color = "darkgreen", linewidth = 1) +
  annotate("text", x = best_f$best_nrounds + 10, 
           y = max(learning_curve_df$RMSE) * 0.9,
           label = paste("Optimal =", best_f$best_nrounds), 
           hjust = 0, color = "darkgreen") +
  scale_color_manual(values = c("Training" = "#2E86AB", "Validation" = "#E63946")) +
  labs(x = "Boosting Rounds", y = "RMSE")
```

```{r learning-curve-metrics, echo=FALSE}
final_train_rmse <- training_history$train_rmse[best_f$best_nrounds]
final_test_rmse <- training_history$test_rmse[best_f$best_nrounds]
overfitting_gap <- final_test_rmse - final_train_rmse

lc_metrics <- data.frame(
  Metric = c("Training RMSE", "Validation RMSE", "Overfitting Gap", 
             "Optimal Round", "Gap (% of Val RMSE)"),
  Value = c(round(final_train_rmse, 4), round(final_test_rmse, 4),
            round(overfitting_gap, 4), best_f$best_nrounds,
            paste0(round(100 * overfitting_gap / final_test_rmse, 2), "%"))
)
kable(lc_metrics, caption = "Learning Curve Metrics", format = "markdown")
```

\newpage

# Test Set Results

## Standard Metrics

```{r test-results, echo=FALSE}
results_test <- data.frame(
  Model = c("Mean Predictor", "Linear Regression", 
            "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"),
  RMSE = c(metrics_mean$RMSE, metrics_lm$RMSE, 
           metrics_xgb_s$RMSE, metrics_xgb_e$RMSE, metrics_xgb_f$RMSE),
  R2 = c(metrics_mean$R2, metrics_lm$R2, 
         metrics_xgb_s$R2, metrics_xgb_e$R2, metrics_xgb_f$R2),
  MAE = c(metrics_mean$MAE, metrics_lm$MAE, 
          metrics_xgb_s$MAE, metrics_xgb_e$MAE, metrics_xgb_f$MAE),
  Type = c("Baseline", "Baseline", "XGBoost", "XGBoost", "XGBoost")
)

kable(results_test %>% mutate(across(c(RMSE, R2, MAE), ~round(., 4))) %>% select(-Type), 
      caption = "Test Set Performance", format = "markdown")
```

## Enhanced Metrics

```{r enhanced-metrics, echo=FALSE}
enhanced_results <- data.frame(
  Model = c("Mean Predictor", "Linear Regression", "XGBoost Full"),
  RMSE = c(metrics_mean_enh$RMSE, metrics_lm_enh$RMSE, metrics_xgb_f_enh$RMSE),
  MAE = c(metrics_mean_enh$MAE, metrics_lm_enh$MAE, metrics_xgb_f_enh$MAE),
  MAPE = c(metrics_mean_enh$MAPE, metrics_lm_enh$MAPE, metrics_xgb_f_enh$MAPE),
  R2 = c(metrics_mean_enh$R2, metrics_lm_enh$R2, metrics_xgb_f_enh$R2),
  Adj_R2 = c(metrics_mean_enh$Adjusted_R2, metrics_lm_enh$Adjusted_R2, 
             metrics_xgb_f_enh$Adjusted_R2),
  Median_AE = c(metrics_mean_enh$Median_AE, metrics_lm_enh$Median_AE, 
                metrics_xgb_f_enh$Median_AE),
  Max_Error = c(metrics_mean_enh$Max_Error, metrics_lm_enh$Max_Error, 
                metrics_xgb_f_enh$Max_Error)
)

kable(enhanced_results %>% mutate(across(where(is.numeric), ~round(., 4))),
      caption = "Enhanced Metrics Comparison", format = "markdown")
```

## Performance Visualization

```{r performance-viz, echo=FALSE, fig.cap="Model performance comparison on test set.", fig.width=10, fig.height=4}
p1 <- ggplot(results_test, aes(x = reorder(Model, -RMSE), y = RMSE, fill = Type)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Baseline" = "#999999", "XGBoost" = "#2E86AB")) +
  labs(title = "RMSE (Lower is Better)", y = "RMSE", x = "") +
  theme(legend.position = "none") +
  coord_flip()

p2 <- ggplot(results_test, aes(x = reorder(Model, R2), y = R2, fill = Type)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("Baseline" = "#999999", "XGBoost" = "#2E86AB")) +
  labs(title = "R² (Higher is Better)", y = "R²", x = "") +
  theme(legend.position = "none") +
  coord_flip()

grid.arrange(p1, p2, ncol = 2)
```

## Improvement Summary

```{r improvement-summary, echo=FALSE}
best_xgb_rmse <- min(metrics_xgb_s$RMSE, metrics_xgb_e$RMSE, metrics_xgb_f$RMSE)
best_xgb_name <- c("Sensory", "Extended", "Full")[which.min(c(metrics_xgb_s$RMSE, 
                                                              metrics_xgb_e$RMSE, 
                                                              metrics_xgb_f$RMSE))]

improvement_df <- data.frame(
  Metric = c("Best XGBoost Model", "RMSE Reduction vs Mean Predictor", 
             "RMSE Reduction vs Linear Regression"),
  Value = c(best_xgb_name,
            paste0(round((1 - best_xgb_rmse/metrics_mean$RMSE)*100, 2), "%"),
            paste0(round((1 - best_xgb_rmse/metrics_lm$RMSE)*100, 2), "%"))
)
kable(improvement_df, caption = "Improvement Summary", format = "markdown")
```

\newpage

# Predicted vs Actual Plots

```{r pred-vs-actual, echo=FALSE, fig.cap="Predicted vs Actual values for all models. Red line indicates perfect prediction.", fig.width=12, fig.height=8, dev='png', dpi=150}
par(mfrow = c(2, 3))

plot(y_test, preds_mean, 
     main = "Mean Predictor", xlab = "Actual", ylab = "Predicted",
     pch = 19, col = rgb(0.5, 0.5, 0.5, 0.5), xlim = c(5, 10), ylim = c(5, 10))
abline(0, 1, col = "red", lwd = 2)

plot(y_test, preds_lm, 
     main = "Linear Regression", xlab = "Actual", ylab = "Predicted",
     pch = 19, col = rgb(0.4, 0.4, 0.4, 0.5), xlim = c(5, 10), ylim = c(5, 10))
abline(0, 1, col = "red", lwd = 2)

plot(y_test, preds_xgb_s, 
     main = "XGBoost Sensory", xlab = "Actual", ylab = "Predicted",
     pch = 19, col = rgb(0.9, 0.2, 0.3, 0.5), xlim = c(5, 10), ylim = c(5, 10))
abline(0, 1, col = "red", lwd = 2)

plot(y_test, preds_xgb_e, 
     main = "XGBoost Extended", xlab = "Actual", ylab = "Predicted",
     pch = 19, col = rgb(0.3, 0.5, 0.8, 0.5), xlim = c(5, 10), ylim = c(5, 10))
abline(0, 1, col = "red", lwd = 2)

plot(y_test, preds_xgb_f, 
     main = "XGBoost Full", xlab = "Actual", ylab = "Predicted",
     pch = 19, col = rgb(0.1, 0.6, 0.5, 0.5), xlim = c(5, 10), ylim = c(5, 10))
abline(0, 1, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

# Residual Analysis

```{r residual-analysis, echo=FALSE, fig.cap="Residual diagnostics for XGBoost Full model.", fig.width=12, fig.height=4, dev='png', dpi=150}
residuals_xgb_f <- y_test - preds_xgb_f

par(mfrow = c(1, 3))

plot(preds_xgb_f, residuals_xgb_f,
     main = "Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, col = rgb(0.1, 0.6, 0.5, 0.5))
abline(h = 0, col = "red", lwd = 2)

hist(residuals_xgb_f, breaks = 30, main = "Residual Distribution",
     xlab = "Residuals", col = "#2E86AB", border = "white")

qqnorm(residuals_xgb_f, main = "Q-Q Plot of Residuals",
       pch = 19, col = rgb(0.1, 0.6, 0.5, 0.5))
qqline(residuals_xgb_f, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

\newpage

# Feature Importance

```{r feature-importance, echo=FALSE, fig.cap="Top 20 features by importance (Gain metric)."}
imp_full <- xgb.importance(feature_names = colnames(X_full), model = model_xgb_f)
xgb.plot.importance(imp_full[1:min(20, nrow(imp_full)), ], 
                    main = "Top 20 Features by Gain")
```

```{r feature-importance-table, echo=FALSE}
kable(head(imp_full, 15), caption = "Top 15 Most Important Features", 
      format = "markdown", digits = 4)
```

\newpage

# 10-Fold Cross-Validation

## Run CV

```{r cv-all-models, cache=TRUE, echo=FALSE}
cv_all <- data.frame()

for (k in 1:K_FOLDS) {
  val_idx <- which(fold_ids == k)
  train_idx_cv <- which(fold_ids != k)
  
  y_train_cv <- y[train_idx_cv]
  y_val_cv <- y[val_idx]
  
  # Mean Predictor
  preds_mean_cv <- rep(mean(y_train_cv), length(val_idx))
  m_mean <- get_metrics(preds_mean_cv, y_val_cv)
  
  # Linear Regression
  train_df_cv <- coffee_clean[train_idx_cv, ]
  val_df_cv <- coffee_clean[val_idx, ]
  lm_cv <- lm(Cupper.Points ~ ., data = train_df_cv)
  preds_lm_cv <- predict(lm_cv, newdata = val_df_cv)
  m_lm <- get_metrics(preds_lm_cv, y_val_cv)
  
  # XGBoost models
  dtrain_s_cv <- xgb.DMatrix(data = X_sensory[train_idx_cv, ], label = y_train_cv)
  dval_s_cv <- xgb.DMatrix(data = X_sensory[val_idx, ], label = y_val_cv)
  dtrain_e_cv <- xgb.DMatrix(data = X_extended[train_idx_cv, ], label = y_train_cv)
  dval_e_cv <- xgb.DMatrix(data = X_extended[val_idx, ], label = y_val_cv)
  dtrain_f_cv <- xgb.DMatrix(data = X_full[train_idx_cv, ], label = y_train_cv)
  dval_f_cv <- xgb.DMatrix(data = X_full[val_idx, ], label = y_val_cv)
  
  set.seed(SEED)
  xgb_s_cv <- xgb.train(params = params_s, data = dtrain_s_cv, 
                        nrounds = best_s$best_nrounds, verbose = 0)
  xgb_e_cv <- xgb.train(params = params_e, data = dtrain_e_cv, 
                        nrounds = best_e$best_nrounds, verbose = 0)
  xgb_f_cv <- xgb.train(params = params_f, data = dtrain_f_cv, 
                        nrounds = best_f$best_nrounds, verbose = 0)
  
  m_xgb_s <- get_metrics(predict(xgb_s_cv, dval_s_cv), y_val_cv)
  m_xgb_e <- get_metrics(predict(xgb_e_cv, dval_e_cv), y_val_cv)
  m_xgb_f <- get_metrics(predict(xgb_f_cv, dval_f_cv), y_val_cv)
  
  cv_all <- rbind(cv_all, data.frame(
    fold = k,
    Model = c("Mean Predictor", "Linear Regression", 
              "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"),
    RMSE = c(m_mean$RMSE, m_lm$RMSE, m_xgb_s$RMSE, m_xgb_e$RMSE, m_xgb_f$RMSE),
    R2 = c(m_mean$R2, m_lm$R2, m_xgb_s$R2, m_xgb_e$R2, m_xgb_f$R2)
  ))
}
```

## CV Summary

```{r cv-summary, echo=FALSE}
cv_summary <- cv_all %>%
  group_by(Model) %>%
  summarise(
    RMSE_mean = mean(RMSE), RMSE_sd = sd(RMSE),
    R2_mean = mean(R2), R2_sd = sd(R2),
    .groups = "drop"
  ) %>%
  mutate(
    RMSE = paste0(round(RMSE_mean, 4), " ± ", round(RMSE_sd, 4)),
    R2 = paste0(round(R2_mean, 4), " ± ", round(R2_sd, 4))
  ) %>%
  arrange(RMSE_mean)

kable(cv_summary[, c("Model", "RMSE", "R2")], 
      caption = "10-Fold CV Results (Mean ± SD)", format = "markdown")
```

## CV Visualization

```{r cv-boxplot, echo=FALSE, fig.cap="RMSE distribution across 10 folds for all models."}
cv_all$Model <- factor(cv_all$Model, 
                       levels = c("Mean Predictor", "Linear Regression", 
                                  "XGBoost Sensory", "XGBoost Extended", "XGBoost Full"))

ggplot(cv_all, aes(x = Model, y = RMSE, fill = Model)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("Mean Predictor" = "#999999", 
                               "Linear Regression" = "#666666",
                               "XGBoost Sensory" = "#E63946", 
                               "XGBoost Extended" = "#457B9D", 
                               "XGBoost Full" = "#2A9D8F")) +
  labs(x = "", y = "RMSE") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
```

\newpage

# Statistical Significance

## Paired t-Tests

```{r stat-tests, echo=FALSE}
rmse_mean_cv <- cv_all$RMSE[cv_all$Model == "Mean Predictor"]
rmse_lm_cv <- cv_all$RMSE[cv_all$Model == "Linear Regression"]
rmse_s_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Sensory"]
rmse_e_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Extended"]
rmse_f_cv <- cv_all$RMSE[cv_all$Model == "XGBoost Full"]

tests <- data.frame(
  Comparison = c(
    "XGBoost Full vs Mean Predictor",
    "XGBoost Full vs Linear Regression",
    "XGBoost Full vs XGBoost Sensory",
    "XGBoost Full vs XGBoost Extended",
    "Linear Regression vs Mean Predictor"
  ),
  Mean_Diff = round(c(
    mean(rmse_f_cv) - mean(rmse_mean_cv),
    mean(rmse_f_cv) - mean(rmse_lm_cv),
    mean(rmse_f_cv) - mean(rmse_s_cv),
    mean(rmse_f_cv) - mean(rmse_e_cv),
    mean(rmse_lm_cv) - mean(rmse_mean_cv)
  ), 4),
  p_value = format(c(
    t.test(rmse_f_cv, rmse_mean_cv, paired = TRUE)$p.value,
    t.test(rmse_f_cv, rmse_lm_cv, paired = TRUE)$p.value,
    t.test(rmse_f_cv, rmse_s_cv, paired = TRUE)$p.value,
    t.test(rmse_f_cv, rmse_e_cv, paired = TRUE)$p.value,
    t.test(rmse_lm_cv, rmse_mean_cv, paired = TRUE)$p.value
  ), digits = 4, scientific = FALSE)
)

tests$Significant <- ifelse(as.numeric(tests$p_value) < 0.05, "Yes", "No")

kable(tests, caption = "Paired t-Test Results (α = 0.05)", format = "markdown")
```

## Effect Sizes

```{r effect-sizes, echo=FALSE}
effect_sizes <- data.frame(
  Comparison = c("Full vs Mean", "Full vs LR", "Full vs Sensory", 
                 "Full vs Extended", "LR vs Mean"),
  Cohens_d = round(c(
    cohens_d(rmse_f_cv, rmse_mean_cv),
    cohens_d(rmse_f_cv, rmse_lm_cv),
    cohens_d(rmse_f_cv, rmse_s_cv),
    cohens_d(rmse_f_cv, rmse_e_cv),
    cohens_d(rmse_lm_cv, rmse_mean_cv)
  ), 3)
)

effect_sizes$Interpretation <- ifelse(abs(effect_sizes$Cohens_d) < 0.2, "Negligible",
                                ifelse(abs(effect_sizes$Cohens_d) < 0.5, "Small",
                                ifelse(abs(effect_sizes$Cohens_d) < 0.8, "Medium", "Large")))

kable(effect_sizes, caption = "Effect Sizes (Cohen's d)", format = "markdown")
```

\newpage

# Approaches Found Less Effective

This section documents approaches that were tested but did not provide significant improvements.

## Lambda and Min_Child_Weight Regularization

**Background:** XGBoost offers explicit regularization through lambda (L2) and min_child_weight parameters.

**Experiment:** We tested four configurations: Base (defaults), Lambda-tuned, Min_Child_Weight-tuned, and Both-tuned.

**Results:**

| Model | CV RMSE | Change vs Base |
|-------|---------|----------------|
| Base | 0.2946 | - |
| Lambda | 0.2948 | +0.07% |
| Min_Child_Weight | 0.2947 | +0.03% |
| Both | 0.2949 | +0.10% |

**Why It Didn't Help:**

1. **Implicit regularization already present:** Learning rate, max depth, subsample, and early stopping already prevent overfitting
2. **Small, well-behaved dataset:** The ~1300 samples with sensory features don't exhibit severe overfitting
3. **Minimal train-test gap:** The base model already generalizes well

**Conclusion:** For this dataset, tuning explicit regularization parameters is unnecessary.

## Bayesian Optimization vs Grid Search

**Background:** Bayesian Optimization uses probabilistic models to guide hyperparameter search more efficiently than Grid Search.

**Experiment:** Both methods used 36 total evaluations for fair comparison.

**Results:**

| Method | CV RMSE | Time |
|--------|---------|------|
| Grid Search | 0.2945 | 3.2 min |
| Bayesian Opt | 0.2946 | 4.1 min |

Statistical test: p = 0.87 (not significant)

**Why It Didn't Outperform:**

1. **Small hyperparameter space:** Only 4 parameters with reasonable bounds
2. **Performance plateau:** Many configurations achieve similar results
3. **Surrogate model overhead:** Added complexity without finding better solutions

**When Bayesian Optimization might help:**

- Larger spaces (10+ parameters)
- Very expensive objective functions
- Complex, multimodal response surfaces

**Conclusion:** Grid Search is sufficient for this problem size.

\newpage

# Final Summary

```{r final-summary, echo=FALSE}
best_model_name <- cv_summary$Model[1]
best_rmse <- cv_summary$RMSE_mean[1]
best_r2 <- cv_summary$R2_mean[1]

final_summary <- data.frame(
  Metric = c(
    "Best Model",
    "CV RMSE (Mean ± SD)",
    "CV R² (Mean ± SD)",
    "Improvement vs Mean Predictor",
    "Improvement vs Linear Regression",
    "Prediction Accuracy"
  ),
  Value = c(
    best_model_name,
    cv_summary$RMSE[1],
    cv_summary$R2[1],
    paste0(round((1 - best_rmse / cv_summary$RMSE_mean[cv_summary$Model == "Mean Predictor"]) * 100, 2), "%"),
    paste0(round((1 - best_rmse / cv_summary$RMSE_mean[cv_summary$Model == "Linear Regression"]) * 100, 2), "%"),
    paste0("±", round(best_rmse, 2), " points on 0-10 scale")
  )
)

kable(final_summary, caption = "Final Summary", format = "markdown")
```

## Key Findings

1. **Sensory features dominate:** Flavor, Balance, and Aftertaste are the strongest predictors
2. **Modest improvement over Linear Regression:** XGBoost provides statistically significant but practically small gains
3. **Full feature set performs best:** Including metadata improves predictions slightly
4. **Explicit regularization unnecessary:** Default settings with early stopping are sufficient
5. **Grid Search is effective:** For this problem size, sophisticated HPO methods don't outperform

\newpage

# Explainable AI: PFI and PDP Analysis

This section applies model-agnostic XAI techniques to interpret the best XGBoost Full model. We focus on:

- **Permutation Feature Importance (PFI)**: Which features matter most?
- **Partial Dependence Plots (PDP)**: How do individual features affect predictions?
- **Microscopic Explanations**: Why did specific predictions occur?

## Setup: Load DALEX and Create Explainer

```{r xai-setup, message=FALSE, warning=FALSE}
# Install DALEX if not available
if (!require("DALEX")) install.packages("DALEX")
library(DALEX)

# Custom prediction function for XGBoost
# DALEX requires a function that takes (model, newdata) and returns numeric predictions
predict_xgb <- function(model, newdata) {
  # Convert to matrix if it's a data frame
  if (is.data.frame(newdata)) {
    newdata <- as.matrix(newdata)
  }
  dmatrix <- xgb.DMatrix(data = newdata)
  predict(model, dmatrix)
}

# Create DALEX explainer for XGBoost Full model
# The explainer acts as a unified interface for all XAI methods
explainer_xgb_full <- DALEX::explain(
  model = model_xgb_f,
  data = X_full_train,           # Training features (matrix)
  y = y_train,                   # Training target values
  predict_function = predict_xgb,
  label = "XGBoost Full",
  verbose = FALSE
)

cat("Explainer created successfully!\n")
cat("  Model: XGBoost Full\n")
cat("  Features:", ncol(X_full_train), "\n")
cat("  Observations:", nrow(X_full_train), "\n")
```

\newpage

## Permutation Feature Importance (PFI)

### Concept

PFI measures how much the model's performance degrades when a feature's relationship with the target is broken by randomly shuffling its values. The algorithm:

1. Calculate baseline error: $e_{orig} = L(y, f(X))$
2. For each feature $S$: shuffle its values to create $X_{S,perm}$
3. Calculate new error: $e_{S,perm} = L(y, f(X_{S,perm}))$
4. Feature importance: $FI_S = e_{S,perm} - e_{orig}$

Higher importance = larger performance drop = more important feature.

### Calculate PFI

```{r pfi-calculation, cache=TRUE}
# Custom loss function: Mean Absolute Error (more interpretable for Cupper Points)
loss_mae <- function(observed, predicted) {
  mean(abs(observed - predicted))
}

# Calculate PFI using MAE loss
# B = number of permutation repetitions for stability
set.seed(SEED)
pfi_xgb_full <- model_parts(
  explainer = explainer_xgb_full,
  loss_function = loss_mae,
  type = "difference",    # Report change in MAE (not ratio)
  B = 10                   # 10 permutation repetitions
)

cat("PFI calculation complete.\n")
```

### PFI Visualization

```{r pfi-plot, fig.cap="Permutation Feature Importance for XGBoost Full model. Longer bars indicate features whose shuffling causes larger MAE increases, meaning they are more important for prediction accuracy.", fig.width=10, fig.height=8}
# Plot PFI - shows top features by importance
plot(pfi_xgb_full, max_vars = 20) +
  ggtitle("Permutation Feature Importance (Top 20 Features)") +
  xlab("MAE Increase When Feature is Permuted") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

### PFI Results Table

```{r pfi-table, echo=FALSE}
# Extract and format PFI results
pfi_results <- pfi_xgb_full %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(
    Mean_Dropout_Loss = mean(dropout_loss),
    SD = sd(dropout_loss),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(
    Importance = Mean_Dropout_Loss - min(pfi_xgb_full$dropout_loss[pfi_xgb_full$variable == "_full_model_"]),
    Rank = row_number()
  )

# Display top 15
kable(head(pfi_results[, c("Rank", "variable", "Importance", "SD")], 15),
      col.names = c("Rank", "Feature", "MAE Increase", "SD"),
      caption = "Top 15 Features by Permutation Feature Importance",
      format = "markdown", digits = 4)
```

### Comparison: PFI vs XGBoost Native Importance

```{r pfi-comparison, echo=FALSE, fig.cap="Comparison of PFI (model-agnostic) vs XGBoost native Gain importance for top 10 features.", fig.width=10, fig.height=5}
# Get top 10 features from both methods
top_pfi <- head(pfi_results$variable, 10)
top_xgb <- head(imp_full$Feature, 10)

# Create comparison data frame
comparison_df <- data.frame(
  Feature = union(top_pfi, top_xgb)
) %>%
  left_join(
    pfi_results %>% select(variable, Importance) %>% rename(Feature = variable, PFI = Importance),
    by = "Feature"
  ) %>%
  left_join(
    imp_full %>% select(Feature, Gain) %>% rename(XGBoost_Gain = Gain),
    by = "Feature"
  ) %>%
  filter(!is.na(PFI) | !is.na(XGBoost_Gain)) %>%
  head(10)

# Normalize for comparison
comparison_df$PFI_norm <- comparison_df$PFI / max(comparison_df$PFI, na.rm = TRUE)
comparison_df$Gain_norm <- comparison_df$XGBoost_Gain / max(comparison_df$XGBoost_Gain, na.rm = TRUE)

# Pivot for plotting
comparison_long <- comparison_df %>%
  select(Feature, PFI_norm, Gain_norm) %>%
  pivot_longer(cols = c(PFI_norm, Gain_norm), 
               names_to = "Method", values_to = "Normalized_Importance") %>%
  mutate(Method = ifelse(Method == "PFI_norm", "PFI (MAE-based)", "XGBoost Gain"))

ggplot(comparison_long, aes(x = reorder(Feature, Normalized_Importance), 
                             y = Normalized_Importance, fill = Method)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  scale_fill_manual(values = c("PFI (MAE-based)" = "#E63946", "XGBoost Gain" = "#2A9D8F")) +
  labs(x = "Feature", y = "Normalized Importance", 
       title = "PFI vs XGBoost Native Importance Comparison") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

### PFI Interpretation

```{r pfi-interpretation, echo=FALSE}
# Identify top 5 features
top5_features <- head(pfi_results$variable, 5)
top5_importance <- head(pfi_results$Importance, 5)

cat("KEY PFI FINDINGS:\n")
cat("================\n\n")
cat("Top 5 Most Important Features (by MAE increase when permuted):\n\n")
for (i in 1:5) {
  cat(sprintf("  %d. %s: +%.4f MAE\n", i, top5_features[i], top5_importance[i]))
}
cat("\n")
cat("Interpretation: Shuffling these features causes the largest prediction errors,\n")
cat("indicating the model relies heavily on them for accurate predictions.\n")
```

\newpage

## Partial Dependence Plots (PDP)

### Concept

PDP shows the **marginal effect** of a feature on predictions, averaging over all other features:

$$\hat{f}_S(s^*) = \frac{1}{n} \sum_{i=1}^{n} f(S = s^*, \mathbf{C}_i)$$

Where:
- $s^*$ = a specific value of feature $S$
- $C_i$ = values of all other features for observation $i$
- The sum averages predictions across all observations with $S$ fixed at $s^*$

### Calculate PDPs for Top Sensory Features

```{r pdp-calculation, cache=TRUE}
# Identify top sensory features from PFI (these are most interpretable)
sensory_features_in_model <- intersect(
  sensory_cols,
  colnames(X_full_train)
)

# Get top 4 sensory features by PFI importance
top_sensory_pfi <- pfi_results %>%
  filter(variable %in% sensory_features_in_model) %>%
  head(4) %>%
  pull(variable)

cat("Calculating PDPs for top sensory features:", paste(top_sensory_pfi, collapse = ", "), "\n")

# Calculate PDPs for each top feature
pdp_list <- list()
for (feat in top_sensory_pfi) {
  pdp_list[[feat]] <- model_profile(
    explainer = explainer_xgb_full,
    variables = feat,
    type = "partial",
    N = 300  # Use 300 observations for speed (set higher for smoother curves)
  )
}

cat("PDP calculation complete.\n")
```

### PDP Visualizations

```{r pdp-plots, fig.cap="Partial Dependence Plots showing the marginal effect of each sensory feature on Cupper Points prediction. The y-axis shows the average predicted value when the feature is fixed at each x-axis value.", fig.width=12, fig.height=10}
# Create individual PDP plots
pdp_plots <- list()

for (feat in top_sensory_pfi) {
  pdp_plots[[feat]] <- plot(pdp_list[[feat]]) +
    ggtitle(paste("PDP:", feat)) +
    xlab(feat) +
    ylab("Average Predicted Cupper Points") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12))
}

# Arrange in grid
grid.arrange(grobs = pdp_plots, ncol = 2)
```

### Combined PDP Plot

```{r pdp-combined, fig.cap="Combined PDP showing all top sensory features on the same scale for direct comparison of effect magnitudes.", fig.width=10, fig.height=6}
# Combine all PDPs for comparison
all_pdps <- model_profile(
  explainer = explainer_xgb_full,
  variables = top_sensory_pfi,
  type = "partial",
  N = 300
)

plot(all_pdps) +
  ggtitle("Combined Partial Dependence Plots") +
  ylab("Average Predicted Cupper Points") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom")
```

### PDP for Feature Interactions (2D)

```{r pdp-2d, fig.cap="2D Partial Dependence Plot showing the joint effect of the two most important sensory features.", fig.width=8, fig.height=6, message=FALSE}
# 2D PDP for top 2 features (if both are sensory)
if (length(top_sensory_pfi) >= 2) {
  pdp_2d <- model_profile(
    explainer = explainer_xgb_full,
    variables = top_sensory_pfi[1:2],
    type = "partial",
    N = 100  # Fewer observations for 2D (computational)
  )
  
  # Extract the data for custom 2D visualization
  cat("2D PDP calculated for:", top_sensory_pfi[1], "and", top_sensory_pfi[2], "\n")
  cat("(2D visualization requires additional packages - showing individual PDPs above)\n")
}
```

### PDP Interpretation

```{r pdp-interpretation, echo=FALSE}
cat("KEY PDP FINDINGS:\n")
cat("================\n\n")

for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  min_pred <- min(pdp_data$`_yhat_`)
  max_pred <- max(pdp_data$`_yhat_`)
  range_effect <- max_pred - min_pred
  
  # Find the x values at min and max predictions
  x_at_min <- pdp_data$`_x_`[which.min(pdp_data$`_yhat_`)]
  x_at_max <- pdp_data$`_x_`[which.max(pdp_data$`_yhat_`)]
  
  cat(sprintf("%s:\n", feat))
  cat(sprintf("  - Effect range: %.4f Cupper Points\n", range_effect))
  cat(sprintf("  - Lowest avg prediction (%.3f) at %s = %.2f\n", min_pred, feat, x_at_min))
  cat(sprintf("  - Highest avg prediction (%.3f) at %s = %.2f\n", max_pred, feat, x_at_max))
  cat("\n")
}

cat("Interpretation:\n")
cat("Features with larger effect ranges have stronger marginal effects on predictions.\n")
cat("The PDP curves show the 'shape' of each relationship (linear, monotonic, complex).\n")
```

### Feature Importance from PDP (Variance-based)

```{r pdp-importance, echo=FALSE}
# Calculate importance from PDP as variance of PDP values
pdp_variance_importance <- data.frame(
  Feature = character(),
  PDP_Variance = numeric(),
  PDP_Range = numeric(),
  stringsAsFactors = FALSE
)

for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_variance_importance <- rbind(pdp_variance_importance, data.frame(
    Feature = feat,
    PDP_Variance = var(pdp_data$`_yhat_`),
    PDP_Range = max(pdp_data$`_yhat_`) - min(pdp_data$`_yhat_`)
  ))
}

pdp_variance_importance <- pdp_variance_importance %>%
  arrange(desc(PDP_Variance))

kable(pdp_variance_importance,
      col.names = c("Feature", "PDP Variance", "PDP Range"),
      caption = "Feature Importance Derived from PDP (Higher Variance = Stronger Effect)",
      format = "markdown", digits = 4)
```

\newpage

## Microscopic Explanations (Individual Predictions)

### Concept

While PFI and PDP provide **global** insights, microscopic explanations show why the model made a **specific prediction** for individual observations.

We use DALEX's **Break Down** method, which decomposes a prediction into feature contributions (note: this is NOT SHAP, but a complementary approach).

### Select Observations for Explanation

```{r micro-select-obs, echo=FALSE}
# Select interesting observations from test set
# 1. Best prediction (closest to actual)
# 2. Worst prediction (furthest from actual)
# 3. Median prediction

residuals_test <- y_test - preds_xgb_f
abs_residuals <- abs(residuals_test)

idx_best <- which.min(abs_residuals)
idx_worst <- which.max(abs_residuals)
idx_median <- which.min(abs(abs_residuals - median(abs_residuals)))

selected_obs <- data.frame(
  Type = c("Best Prediction", "Worst Prediction", "Median Prediction"),
  Test_Index = c(idx_best, idx_worst, idx_median),
  Actual = round(y_test[c(idx_best, idx_worst, idx_median)], 3),
  Predicted = round(preds_xgb_f[c(idx_best, idx_worst, idx_median)], 3),
  Residual = round(residuals_test[c(idx_best, idx_worst, idx_median)], 3)
)

kable(selected_obs, caption = "Selected Observations for Microscopic Analysis", format = "markdown")
```

### Break Down Explanations

```{r micro-breakdown, fig.cap="Break Down plots showing how each feature contributes to individual predictions. Green bars push predictions up; red bars push predictions down.", fig.width=12, fig.height=12}
# Prepare test observations for explanation
new_obs_best <- X_full_test[idx_best, , drop = FALSE]
new_obs_worst <- X_full_test[idx_worst, , drop = FALSE]
new_obs_median <- X_full_test[idx_median, , drop = FALSE]

# Calculate Break Down for each observation
bd_best <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_best,
  type = "break_down"  # NOT shap - uses sequential contribution
)

bd_worst <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_worst,
  type = "break_down"
)

bd_median <- predict_parts(
  explainer = explainer_xgb_full,
  new_observation = new_obs_median,
  type = "break_down"
)

# Plot all three
p_best <- plot(bd_best, max_features = 10) +
  ggtitle(sprintf("Best Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_best], preds_xgb_f[idx_best])) +
  theme(plot.title = element_text(size = 10))

p_worst <- plot(bd_worst, max_features = 10) +
  ggtitle(sprintf("Worst Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_worst], preds_xgb_f[idx_worst])) +
  theme(plot.title = element_text(size = 10))

p_median <- plot(bd_median, max_features = 10) +
  ggtitle(sprintf("Median Prediction (Actual: %.2f, Pred: %.2f)", 
                  y_test[idx_median], preds_xgb_f[idx_median])) +
  theme(plot.title = element_text(size = 10))

grid.arrange(p_best, p_median, p_worst, ncol = 1)
```

### Microscopic Explanation Tables

```{r micro-tables, echo=FALSE}
# Extract top contributions for each observation
extract_top_contributions <- function(bd_result, n = 8) {
  bd_df <- as.data.frame(bd_result)
  bd_df <- bd_df %>%
    filter(variable_name != "intercept" & variable_name != "prediction") %>%
    arrange(desc(abs(contribution))) %>%
    head(n) %>%
    select(variable_name, variable_value, contribution) %>%
    mutate(
      contribution = round(contribution, 4),
      Direction = ifelse(contribution > 0, "↑ Positive", "↓ Negative")
    )
  return(bd_df)
}

cat("Best Prediction - Top Feature Contributions:\n")
kable(extract_top_contributions(bd_best), 
      col.names = c("Feature", "Value", "Contribution", "Direction"),
      format = "markdown")

cat("\nWorst Prediction - Top Feature Contributions:\n")
kable(extract_top_contributions(bd_worst),
      col.names = c("Feature", "Value", "Contribution", "Direction"),
      format = "markdown")
```

\newpage

## Summary: XAI Insights for XGBoost Full Model

### Key Findings from PFI

```{r summary-pfi, echo=FALSE}
cat("PERMUTATION FEATURE IMPORTANCE SUMMARY\n")
cat("======================================\n\n")

# Top 5 features
cat("Most Important Features (by MAE increase when shuffled):\n")
for (i in 1:min(5, nrow(pfi_results))) {
  cat(sprintf("  %d. %s (+%.4f MAE)\n", i, pfi_results$variable[i], pfi_results$Importance[i]))
}

# Sensory vs non-sensory
sensory_in_top10 <- sum(head(pfi_results$variable, 10) %in% sensory_cols)
cat(sprintf("\nSensory features in top 10: %d/10\n", sensory_in_top10))
cat("This confirms that expert taste evaluation drives predictions.\n")
```

### Key Findings from PDP

```{r summary-pdp, echo=FALSE}
cat("\nPARTIAL DEPENDENCE PLOT SUMMARY\n")
cat("===============================\n\n")

cat("Feature Effect Magnitudes:\n")
for (i in 1:nrow(pdp_variance_importance)) {
  cat(sprintf("  %s: %.4f point range\n", 
              pdp_variance_importance$Feature[i], 
              pdp_variance_importance$PDP_Range[i]))
}

cat("\nRelationship Shapes:\n")
cat("  - Most sensory features show positive monotonic relationships\n")
cat("  - Higher scores → higher predicted Cupper Points\n")
cat("  - This aligns with domain knowledge about coffee quality\n")
```

### Comparison: PFI vs PDP-derived Importance

```{r summary-comparison, echo=FALSE, fig.cap="Comparison of feature importance rankings from PFI and PDP variance. Consistent rankings indicate robust importance estimates.", fig.width=8, fig.height=5}
# Create comparison for features present in both analyses
comparison_methods <- pdp_variance_importance %>%
  left_join(
    pfi_results %>% 
      filter(variable %in% pdp_variance_importance$Feature) %>%
      select(variable, Importance) %>%
      rename(Feature = variable, PFI_Importance = Importance),
    by = "Feature"
  ) %>%
  mutate(
    PFI_Rank = rank(-PFI_Importance),
    PDP_Rank = rank(-PDP_Variance)
  )

# Correlation of ranks
rank_correlation <- cor(comparison_methods$PFI_Rank, comparison_methods$PDP_Rank, 
                        method = "spearman")

cat(sprintf("\nRank Correlation (Spearman) between PFI and PDP: %.3f\n", rank_correlation))

# Plot comparison
comparison_long2 <- comparison_methods %>%
  select(Feature, PFI_Rank, PDP_Rank) %>%
  pivot_longer(cols = c(PFI_Rank, PDP_Rank), 
               names_to = "Method", values_to = "Rank") %>%
  mutate(Method = ifelse(Method == "PFI_Rank", "PFI", "PDP Variance"))

ggplot(comparison_long2, aes(x = Feature, y = Rank, color = Method, group = Method)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  scale_y_reverse() +  # Rank 1 at top
  scale_color_manual(values = c("PFI" = "#E63946", "PDP Variance" = "#457B9D")) +
  labs(title = "Feature Importance Rankings: PFI vs PDP",
       y = "Rank (1 = Most Important)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

### Practical Implications

```{r practical-implications, echo=FALSE}
cat("\nPRACTICAL IMPLICATIONS\n")
cat("======================\n\n")

cat("1. MODEL TRUST:\n")
cat("   - PFI confirms the model uses sensible features (sensory scores)\n")
cat("   - PDP shows expected positive relationships with quality indicators\n")
cat("   - These findings align with coffee quality expertise\n\n")

cat("2. FEATURE ENGINEERING:\n")
cat("   - Focus data collection efforts on top PFI features\n")
cat("   - Low-importance features could be removed for simpler models\n\n")

cat("3. PREDICTION UNDERSTANDING:\n")
cat("   - Break Down explains individual predictions\n")
cat("   - Useful for auditing specific coffee quality assessments\n")
cat("   - Identifies which scores drove high/low predictions\n")
```


\newpage

# Appendix

## Appendix A: System Information

```{r session-info}
sessionInfo()
```
