---
title: "Support Vector Regression for Coffee Quality Prediction"
author: "Luisa Kalkert"
date: "`r Sys.Date()`"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Support Vector Regression

### Data Preprocessing and Cleaning 

#### Feature selection:
In SVM Regression, categorical variables are One-Hot Encoded using dummy variables.
This means, features with many unique values will lead to a sparse feature matrix. This is disadvantageous as it increases computational cost and poses the risk of overfitting. 
Additionally, categories with only a few or even only one observation most likely will not add to model robustness, but are adding noise to the signal.
Therefore, we are removing categorical features with many unique values, only keeping columns with a low number of unique values.
E.g. we are not including the "Region" column with 344 unique values in the model training, because we'd on average have less than 4 observations per Region. 
For features with a high number of missing values, we either remove the column or impute the missing values.
For instance for the "Variety" column, we have about 15% missing values and 8% entries "other" at 30 different varieties. 
Keeping those, would mean adding 30 sparse columns, and at the same time the information gain is most likely not too high with this many missing values.
Therefore, we are not including this feature. 


#### Imputation:
For the altitude column, we have about 17% missing values. Here, we are imputing the missing values from the "Region" column.
For each observation with missing altitude, we are imputing the median altitude of all coffees from the same region.
This covers almost all missing values. For the remaining observations, we are imputing the mean altitude of all coffees from the same country of origin.
As the "Region" and "Country" columns are not used in the model training, we are not creating correlated features with this imputation.
We are imputing the missing values after the train/test split, using only values from the training set, to avoid data leakage from the test set into the training set.


#### Further preprocessing:
The bag weight column contains both pounds and kilograms, so we are converting all to kilograms. 
However, for some entries, it is unclear which unit is used. For these, we are averaging the weight in pounds and kilograms.
This means for those entries the weight is certainly incorrect, but should be within an order of magnitude of the correct weight. 
If we can assume that speciality coffee is sold in smaller quantities, we can still gain valuable information from the weight.


#### Scaling: 
The SVM method used for model training has a built in scaling functionality. 
Both predictor and target variables are scaled to have expected value 0 and variance 1. [https://rdrr.io/rforge/e1071/man/svm.html]
This transformation is based on the training data and applied to train and test data sets.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Read the data
coffee_data <- read.csv("data/arabica_data_cleaned.csv", na.strings = c("", "NA"))



# Avoid categorical variables with many unique values because they will create many dummy variables. 
columns_to_use <- c(
  "Number.of.Bags",
  "Bag.Weight",
  # "Country.of.Origin", Increased runtime too much
  "Processing.Method", 
  "Aroma",
  "Flavor",
  "Aftertaste",
  "Acidity",
  "Body",
  "Balance",
  "Uniformity",
  "Clean.Cup",
  "Sweetness",
  "Cupper.Points", # Target variable
  "Moisture",
  "Category.One.Defects",
  "Quakers",
  "Category.Two.Defects",
  "altitude_mean_meters" # Impute missing values from region and country origins
)




## Bag Weight

# Get bags with weight in pounds
bags_with_weight_in_pounds_indices <- grepl("lbs", coffee_data$Bag.Weight)

# Get bags with weight in kg
bags_with_weight_in_kg_indices <- grepl("kg", coffee_data$Bag.Weight)

# Get bags with neither pounds nor kg
bags_with_neither_pounds_nor_kg_indices <- !bags_with_weight_in_pounds_indices & !bags_with_weight_in_kg_indices

# Get bags with both pounds and kg (incorrectly formatted)
bags_with_both_pounds_and_kg_indices <- bags_with_weight_in_pounds_indices & bags_with_weight_in_kg_indices

# Remove bags with both pounds and kg from bags with weight indices and bags with weight in kg indices
bags_with_weight_in_pounds_indices <-
  bags_with_weight_in_pounds_indices & !bags_with_both_pounds_and_kg_indices

bags_with_weight_in_kg_indices <-
  bags_with_weight_in_kg_indices & !bags_with_both_pounds_and_kg_indices


# Sanity check
# print(paste("Number of bags with weight in pounds:", sum(bags_with_weight_in_pounds_indices)))
# print(paste("Number of bags with weight in kg:", sum(bags_with_weight_in_kg_indices)))
# print(paste("Number of bags with neither pounds nor kg:", sum(bags_with_neither_pounds_nor_kg_indices)))
# print(paste("Number of bags with both pounds and kg:", sum(bags_with_both_pounds_and_kg_indices)))
# print(paste("Number of bags with weight:", length(coffee_data$Bag.Weight)))

# Extract numeric value from Bag.Weight column
coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
# Multiply bags with weight in pounds by 0.453592 to convert to kg
coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] <- coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] * 0.453592

# For bags with neither pounds nor kg, average the weight in pounds and kg
coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] <- coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] * (1 + 0.453592)/2
coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] <- coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] * (1 + 0.453592)/2



# For the Processing Method fill NAs with "Unknown"
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"



# Split data into training and test set
# Separate features and target


# Simple random 80/20 split
SEED <- 123
set.seed(SEED)
n_total <- nrow(coffee_data)
n_train <- floor(0.8 * n_total)
train_index <- sample(1:n_total, n_train)

train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]



## Altitude

library(tidyverse)


# Create dict of regions and their median altitudes
region_altitudes <- train_data %>%
  group_by(Region) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

region_altitudes <- region_altitudes[complete.cases(region_altitudes),]

# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters)) 
train_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(train_data$Region[missing_altitude_indices], region_altitudes$Region)]


# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(test_data$Region[missing_altitude_indices], region_altitudes$Region)]




### Repeat imputation for country of origin

# Create dict of regions and their median altitudes
country_altitudes <- train_data %>%
  group_by(Country.of.Origin) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

country_altitudes <- country_altitudes[complete.cases(country_altitudes),]


# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters))
train_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(train_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]

# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(test_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]




# Remove columns we are not using
train_data <- train_data[,columns_to_use]
test_data <- test_data[,columns_to_use]


# Remove remaining NAs
train_data <- train_data[complete.cases(train_data),]
test_data <- test_data[complete.cases(test_data),]


# Add One Hot encoding
train_data <- data.frame(model.matrix(~ ., data = train_data))[,-1]
test_data  <- data.frame(model.matrix(~ ., data = test_data))[,-1]

# Check if categories align between train and test 
# print("All Categories match between train and Test data")
# print(all(colnames(train_data) == colnames(test_data)))
  

```

```{r, echo=FALSE}
# print(paste0(
#   "Number of training samples: ", nrow(train_data), "\n",
#   "Number of test samples: ", nrow(test_data), "\n",
#   "Number of features: ", ncol(train_data), "\n",
#   "Average cupper points in training set: ", mean(train_data$Cupper.Points), "\n",
#   "Average cupper points in test set: ", mean(test_data$Cupper.Points), "\n",
#   "Features used: ", paste(columns_to_use, collapse = ", ")

# ))

```

### Support Vector Regression - Hyperparameter Optimization

For the prediction of cupper points we train an SVM for regression on the data set, holding out the test data set for final evaluation. 
To choose appropriate hyperparameters, we conduct a hyperparameter optimization on the training data set using 10-fold cross-validation. 
We use a grid search for different values for cost and epsilon for a radial, linear, and 2-degree polynomial kernel. We compare the mean absolute error between runs to choose the best-performing hyperparameter combination. Finally, we train a model with these hyperparameters on the entire training set and evaluate on the test data set.




```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3}




####  Fit SVM model with polynomial kernel (degree 2) 
library(e1071)

# Gridsearch 
epsilon = seq(0,2,0.1)
cost = 2^(-10:3)
# epsilon = seq(0,1,0.5)
# cost = 2^(-2:-1)


# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "polynomial",
                   degree = 2,
                   ranges = list(epsilon = epsilon, cost = cost)
)


# Show Results and visualize tuning as a heatmap 
# print(tuneResult)
library(ggplot2)

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 
 library(dplyr)
 library(tidyr)
 
 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Set up layout for side-by-side plots
par(mfrow = c(1, 3))

# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using 2nd\nDegree Polynomial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


# Best parameters 
best_error <- max(data.frame(data_wide[,-1]))

# Add polynomial kernel results to best results
best_runs <- data.frame(
  kernel = "polynomial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
)




# Fit SVM model with linear kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "linear",
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)


 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Linear Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


best_error <- max(data.frame(data_wide[,-1]))

# add linear kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "linear",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))



# Fit SVM model with radial kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Radial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)

# Reset plotting parameters
par(mfrow = c(1, 1))

best_error <- max(data.frame(data_wide[,-1]))

# add radial kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "radial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))

```
These figures show the mean absolute error for the different hyperparameters by kernel. The best performing hyperparameter combination for each kernel are marked with orange points.

```{r}


# Compare best results for each kernel by getting the row with the lowest error
best_runs <- best_runs[order(best_runs$err),]

best_run <- best_runs[1,]

cat("Best SVM kernel and parameters selected:\n")
print(best_run[,c("kernel", "epsilon", "cost")])






# Train final model with best parameters
epsilon = best_run$epsilon
cost = best_run$cost

svm_model <- svm(Cupper.Points ~ ., data = train_data,
                 cost = cost,
                 epsilon = epsilon, 
                 kernel = best_run$kernel)

# Predict on test set
predictions <- predict(svm_model, test_data)

# Evaluate model performance
mae <- mean(abs(predictions - test_data[, "Cupper.Points"]), na.rm = TRUE)
mse <- mean((predictions - test_data[, "Cupper.Points"])^2, na.rm = TRUE)
rmse <- sqrt(mse)
# print(paste("RMSE:", rmse))
print(paste("MAE:", mae))


```
The best performing kernel is the linear kernel, which means the relationship between the features and the target variable is best approximated by a multidimensional hyperplane.

```{r, echo=FALSE, fig.height=3}
# Plot the predictions vs the true values
plot(predictions, test_data[, "Cupper.Points"],
     xlab = "Predictions",
     ylab = "True Values",
     main = "Predictions vs True Values")

# get predictions for a random data point from the test data set
random_index <- sample(1:nrow(test_data), 1)
random_data <- test_data[random_index,]
random_prediction <- predict(svm_model, random_data)
# print("Random Data Point")
# print(random_data)
# print(paste("Prediction:", random_prediction))
# print(paste("True value:", random_data[, "Cupper.Points"]))

```

Here, we can see the predictions vs. the true values for the test data set. We can see that aside from two outliers, the predictions are close to the true values.










