---
title: "Machine Learning II Project Report"
subtitle: "The Analysis of the Coffee Quality Dataset"
abstract:
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
  word_document:
    toc: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# CLEAR ENVIRONMENT
rm(list = ls(all.names = TRUE))

# LOAD PACKAGES
library(ggplot2)
library(corrplot)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(skimr)
library(e1071)
library(tidyverse)
library(patchwork)
library(DALEX)

# GLOBAL CHUNK OPTIONS
# 'dev' checks if it's PDF to use high-quality rendering, otherwise standard
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  fig.align = "center",
  fig.width = 6, 
  fig.height = 4
)

# GGPLOT THEME
theme_set(theme_minimal())  # check ?theme_bw for other themes available
```

\newpage

# Introduction

This report explores the Coffee Quality dataset, walking through everything from initial descriptive analysis to the math and application of machine learning. Our goal is to use these models to move beyond simple numbers and extract real, meaningful insights into what defines a high-quality cup of coffee.

## Descriptive Analysis

**Summary of the Descriptive Analysis:**
The dataset has 1,311 coffee evaluations (observations) with 44 different variables. It’s a mix of sensory scores (like how the coffee tastes) and environmental data (like where it was grown). Out of these, **24 variables are categorical and 20 are numerical**. For our project, we decided to use Cupper Points as our target variable. We avoided "Total Cup Points" because that number is just a simple sum of the other scores, which would make the prediction too easy and less meaningful.

**Dealing with Missing Data**
One of the first things we noticed was that a lot of the categorical (text) data is missing. Because there were so many gaps, it was hard to use many of these variables in our models. Instead, we focused our energy on the variables that really mattered and tried to fill in the blanks for critical info like Altitude.
```{r}
# Read the data
coffee_data <- read.csv("~/ML2 arabica-coffee-analayze/arabica_data_cleaned.csv", na.strings = c("", "NA", " ", "n/a"))
# 1. Select the features you want to focus on
selected_features <- coffee_data %>% 
  select(Cupper.Points, altitude_mean_meters, Aroma, Flavor, unit_of_measurement, Processing.Method, 
         Bag.Weight, Variety, Lot.Number, Quakers, Country.of.Origin)

# 2. Calculate missing values and prepare for plotting
missing_data <- selected_features %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Missing_Count") %>%
  mutate(Percentage = (Missing_Count / nrow(coffee_data)) * 100) %>%
  arrange(desc(Missing_Count)) # Sort so the biggest problem is first

# 3. Create the bar plot
ggplot(missing_data, aes(x = reorder(Feature, -Missing_Count), y = Missing_Count)) +
  geom_col(fill = "firebrick", color = "white") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            vjust = -0.5, size = 3.5) + # Add percentage labels above bars
  theme_minimal() +
  ylim(0, max(missing_data$Missing_Count) + 50) + # Add space for labels
  labs(title = "Missing Values Count in Selected Features",
       x = "Feature",
       y = "Number of Missing Observations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**The Distribution of the Variables**
When we looked at the histograms and boxplots for the sensory variables, we saw that they all look very similar. Most of the scores for things like Aroma or Flavor are clustered closely together, usually between 7 and 8. We also noticed several outliers*(coffee samples that scored much higher or lower than the rest)*.
```{r}
# Prepare the data in 'long' format
long_data <- coffee_data %>%
  select(Aroma, Flavor, Cupper.Points) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Score")

# Create the Horizontal Combined Boxplot (Left side)
p_box <- ggplot(long_data, aes(y = Feature, x = Score, fill = Feature)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.shape = 16) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Outlier & Range Comparison",
       y = "Feature", x = "Score") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

# Create Faceted Histograms (Right side)
p_hist <- ggplot(long_data, aes(x = Score, fill = Feature)) +
  geom_histogram(bins = 20, color = "white", alpha = 0.8) +
  facet_wrap(~Feature, ncol = 1) + 
  theme_light() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Distribution Shapes",
       x = "Score", y = "Count") +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold", color = "black"),
        plot.title = element_text(face = "bold"))

# Combine Side-by-Side
# We use '|' to put them together. 
(p_box | p_hist) + 
  plot_layout(widths = c(1.2, 1)) + 
  plot_annotation(title = "Sensory Feature Profile: Range and Distribution",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```
**Correlation Between Numerical Features**
Finally, our correlation matrix showed us how these variables lean on each other. Most sensory features are strongly linked. if a coffee smells great (Aroma), it almost always tastes great (Flavor). However, some things, like Moisture, didn't follow the crowd. Moisture has a very low correlation with the other scores and even shows a slight negative trend, meaning more moisture sometimes leads to a slightly lower quality score.
```{r , fig.width=7, fig.height=5, echo=FALSE, fig.align='center'}
# Select only the numerical sensory columns (adjust indices as needed)
# Usually columns like Aroma:Cupper.Points
sensory_cols <- coffee_data %>% select(Aroma, Flavor, Aftertaste, Acidity, Body, Balance, Sweetness, Uniformity, Clean.Cup, Moisture)

# Calculate correlation matrix
M <- cor(sensory_cols, use = "complete.obs")

# Visualize
corrplot(M, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", 
         mar = c(0, 0, 1.5, 0),
         title = "Correlation Between Sensory Attributes"
         )
```

# Data Preprocessing and Cleaning 

**Feature selection:**
In SVM Regression, categorical variables are One-Hot Encoded using dummy variables.
This means, features with many unique values will lead to a sparse feature matrix. This is disadvantageous as it increases computational cost and poses the risk of overfitting. 
Additionally, categories with only a few or even only one observation most likely will not add to model robustness, but are adding noise to the signal.
Therefore, we are removing categorical features with many unique values, only keeping columns with a low number of unique values.
E.g. we are not including the "Region" column with 344 unique values in the model training, because we'd on average have less than 4 observations per Region. 
For features with a high number of missing values, we either remove the column or impute the missing values.
For instance for the "Variety" column, we have about 15% missing values and 8% entries "other" at 30 different varieties. 
Keeping those, would mean adding 30 sparse columns, and at the same time the information gain is most likely not too high with this many missing values.
Therefore, we are not including this feature. 


**Imputation:**
For the altitude column, we have about 17% missing values. Here, we are imputing the missing values from the "Region" column.
For each observation with missing altitude, we are imputing the median altitude of all coffees from the same region.
This covers almost all missing values. For the remaining observations, we are imputing the mean altitude of all coffees from the same country of origin.
As the "Region" and "Country" columns are not used in the model training, we are not creating correlated features with this imputation.
We are imputing the missing values after the train/test split, using only values from the training set, to avoid data leakage from the test set into the training set.


**Further preprocessing:**
The bag weight column contains both pounds and kilograms, so we are converting all to kilograms. 
However, for some entries, it is unclear which unit is used. For these, we are averaging the weight in pounds and kilograms.
This means for those entries the weight is certainly incorrect, but should be within an order of magnitude of the correct weight. 
If we can assume that speciality coffee is sold in smaller quantities, we can still gain valuable information from the weight.


**Scaling:** 
The SVM method used for model training has a built in scaling functionality. 
Both predictor and target variables are scaled to have expected value 0 and variance 1. [https://rdrr.io/rforge/e1071/man/svm.html]
This transformation is based on the training data and applied to train and test data sets.


# A Brief Mathematical Overview of the ML Methods Used
## Support Vector Machine for Regression
```{r, fig.width=10, fig.height=6, out.width='100%', fig.align='center'}
set.seed(123)


x <- seq(-10, 10, 0.5) + rnorm(length(seq(-10, 10, 0.5)), mean = 0, sd = 0.1)
y <- 1/170 * (x^3 - 3* x^2 - 70*x + 5 + rnorm(length(x), mean = 0, sd = 170))

library(e1071)

epsilon <- 1
cost <- 1

model <- svm(y ~ x, data = data.frame(x = x, y = y), epsilon = epsilon, cost = cost)
predictedY <- predict(model, data.frame(x = x))

plot(x, y, pch = 16, col = "grey70")
lines(x, predictedY, col = "darkgreen")


# Add ε-tube
polygon(
 c(x, rev(x)),
 c(predictedY + epsilon, rev(predictedY - epsilon)),
 border = NA,
 col = rgb(1, 0, 0, 0.15)
)

# mark the 17 points that are the furthest from the predictedY 
idx_furthest <- order(abs(predictedY - y), decreasing = TRUE)[1:17]
points(x[idx_furthest], y[idx_furthest], 
       col = "#000000", 
       bg = "#ffffff00",
       pch = 21,
       cex = 1.6,
       lwd = 1)


# Add an arrow (dashed, double sided, between ypred at x = 10 and ypred at x = 10   + 1 )
arrows(10, predictedY[length(predictedY)], 10, predictedY[length(predictedY)] + epsilon, length = 0.1, lty = 2, col = "black", code = 3)
# add a text "ε" at the end of the arrow
text(10, predictedY[length(predictedY)] + 1/2, expression(epsilon), pos = 4)

# Add an arrow between 5,-5 and 5,-2.5 with text xsi_i star (flipped direction)
arrows(5.05, -4, 5.05, -2.28, length = 0.1, lty = 2, col = "black")
text(5.05, -3.2, expression(xi[j]^"*"), pos = 4)


# Add an arrow between the 13th last point and the 13th last point - 1 with text xsi_j
arrows(x[length(x) - 12], y[length(y) - 12], x[length(x) - 12], y[length(y) - 12]- 0.5, length = 0.1, lty = 2, col = "black")
text(x[length(x) - 12], y[length(y) - 12] - 0.3, expression(xi[k]), pos = 4)

```


Support Vector Regression expands support Vector Machines from discrete classification to a continuos regression model.

For the regression task, instead of fitting a hyperplane, we are trying to fit a (multidimensional) function $f(x)$ to the data. Similar as in the classification task, we draw a margin around the function. However, now the margin $\varepsilon$ is fixed at the start. (Note that $\varepsilon$ here, is not the slack variable as in Hasties and Tibshirani, but rather defines the width of the margin. Also, we don't aim to optimize the margin anymore, but instead fix it before the computation starts). We want to find a function f(x) that is as flat as possible, while fitting all points into the required margin. Again, we allow for some slack using slack variables. A wider margin decreases the risk of overfitting, while a smaller margin captures more intricacies of the data. 

Mathematically speaking, want to find a function $f(x)$ with $| f(x_i) - y_i |  \leq \varepsilon$ for all $i$. To avoid overfitting, we also introduce the constraint of making the function as flat as possible.
"Flatness" in this context is a measure of how sensitive the function is to change. For a linear function $f(x) = x^\top \beta + b$ increasing this flatness can be expressed through minimizing the norm of the function gradient: $\lVert \beta \rVert$. Minimizing the expression $\frac{1}{2} \lVert \beta \rVert^2$ leads to the same optimum, while allowing for more elegant mathematical solutions, and is therefore used for computation.


Again, we want to allow for some slack, so as in the SVM for classification, we are introducing slack variables $\xi_i$ and $\xi_i^\ast$. These slack variables allow for some room for margin violations. 
The optimization problem we want to solve is to minimize: 

$$
\frac{1}{2} \lVert \beta \rVert^2 + K \sum_{i=1}^n (\xi_i + \xi_i^\ast)
$$
subject to:
$$
y_i - f(x_i) \leq \varepsilon + \xi_i, 
$$
$$
f(x_i) - y_i \leq \varepsilon + \xi_i^\ast
$$
and 
$$
\xi_i > 0, \xi_i^\ast > 0 \quad \forall i
$$


where $K$ is fixed and $\beta$, $\xi$ and $\xi^\ast$ are variable.
The factor $K$ is introduced as cost for regularization purposes. It is choosen a priori to determine how strong the tradeoff between flatness and overfitting should be. A larger $K$ leads to a stronger impact of the term, therefore penalizing stronger margin violations more heavily and resulting in a less flat function curve (higher bias, lower variance). Conversly, a smaller $K$ leads to a flatter curve (lower bias, higher variance).


As in the SVM for classification, this can be generalized to non-linear functions by transforming the features using kernels.

## XGBoost

# Applications
## Support Vector Regression
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3}
# Avoid categorical variables with many unique values because they will create many dummy variables. 
columns_to_use <- c(
  "Number.of.Bags",
  "Bag.Weight",
  # "Country.of.Origin", Increased runtime too much
  "Processing.Method", 
  "Aroma",
  "Flavor",
  "Aftertaste",
  "Acidity",
  "Body",
  "Balance",
  "Uniformity",
  "Clean.Cup",
  "Sweetness",
  "Cupper.Points", # Target variable
  "Moisture",
  "Category.One.Defects",
  "Quakers",
  "Category.Two.Defects",
  "altitude_mean_meters" # Impute missing values from region and country origins
)




## Bag Weight

# Get bags with weight in pounds
bags_with_weight_in_pounds_indices <- grepl("lbs", coffee_data$Bag.Weight)

# Get bags with weight in kg
bags_with_weight_in_kg_indices <- grepl("kg", coffee_data$Bag.Weight)

# Get bags with neither pounds nor kg
bags_with_neither_pounds_nor_kg_indices <- !bags_with_weight_in_pounds_indices & !bags_with_weight_in_kg_indices

# Get bags with both pounds and kg (incorrectly formatted)
bags_with_both_pounds_and_kg_indices <- bags_with_weight_in_pounds_indices & bags_with_weight_in_kg_indices

# Remove bags with both pounds and kg from bags with weight indices and bags with weight in kg indices
bags_with_weight_in_pounds_indices <-
  bags_with_weight_in_pounds_indices & !bags_with_both_pounds_and_kg_indices

bags_with_weight_in_kg_indices <-
  bags_with_weight_in_kg_indices & !bags_with_both_pounds_and_kg_indices


# Sanity check
# print(paste("Number of bags with weight in pounds:", sum(bags_with_weight_in_pounds_indices)))
# print(paste("Number of bags with weight in kg:", sum(bags_with_weight_in_kg_indices)))
# print(paste("Number of bags with neither pounds nor kg:", sum(bags_with_neither_pounds_nor_kg_indices)))
# print(paste("Number of bags with both pounds and kg:", sum(bags_with_both_pounds_and_kg_indices)))
# print(paste("Number of bags with weight:", length(coffee_data$Bag.Weight)))

# Extract numeric value from Bag.Weight column
coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
# Multiply bags with weight in pounds by 0.453592 to convert to kg
coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] <- coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] * 0.453592

# For bags with neither pounds nor kg, average the weight in pounds and kg
coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] <- coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] * (1 + 0.453592)/2
coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] <- coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] * (1 + 0.453592)/2



# For the Processing Method fill NAs with "Unknown"
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"



# Split data into training and test set
# Separate features and target


# Simple random 80/20 split
SEED <- 123
set.seed(SEED)
n_total <- nrow(coffee_data)
n_train <- floor(0.8 * n_total)
train_index <- sample(1:n_total, n_train)

train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]



## Altitude



# Create dict of regions and their median altitudes
region_altitudes <- train_data %>%
  group_by(Region) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

region_altitudes <- region_altitudes[complete.cases(region_altitudes),]

# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters)) 
train_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(train_data$Region[missing_altitude_indices], region_altitudes$Region)]


# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(test_data$Region[missing_altitude_indices], region_altitudes$Region)]




### Repeat imputation for country of origin

# Create dict of regions and their median altitudes
country_altitudes <- train_data %>%
  group_by(Country.of.Origin) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

country_altitudes <- country_altitudes[complete.cases(country_altitudes),]


# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters))
train_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(train_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]

# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(test_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]




# Remove columns we are not using
train_data <- train_data[,columns_to_use]
test_data <- test_data[,columns_to_use]


# Remove remaining NAs
train_data <- train_data[complete.cases(train_data),]
test_data <- test_data[complete.cases(test_data),]


# Add One Hot encoding
train_data <- data.frame(model.matrix(~ ., data = train_data))[,-1]
test_data  <- data.frame(model.matrix(~ ., data = test_data))[,-1]

# Check if categories align between train and test 
# print("All Categories match between train and Test data")
# print(all(colnames(train_data) == colnames(test_data)))
  

```

```{r, echo=FALSE}
# print(paste0(
#   "Number of training samples: ", nrow(train_data), "\n",
#   "Number of test samples: ", nrow(test_data), "\n",
#   "Number of features: ", ncol(train_data), "\n",
#   "Average cupper points in training set: ", mean(train_data$Cupper.Points), "\n",
#   "Average cupper points in test set: ", mean(test_data$Cupper.Points), "\n",
#   "Features used: ", paste(columns_to_use, collapse = ", ")

# ))

```

### Support Vector Regression - Hyperparameter Optimization

For the prediction of cupper points we train an SVM for regression on the data set, holding out the test data set for final evaluation. 
To choose appropriate hyperparameters, we conduct a hyperparameter optimization on the training data set using 10-fold cross-validation. 
We use a grid search for different values for cost and epsilon for a radial, linear, and 2-degree polynomial kernel. We compare the mean absolute error between runs to choose the best-performing hyperparameter combination. Finally, we train a model with these hyperparameters on the entire training set and evaluate on the test data set.



```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3}




####  Fit SVM model with polynomial kernel (degree 2) 


# Gridsearch 
epsilon = seq(0,2,0.1)
cost = 2^(-10:3)
# epsilon = seq(0,1,0.5)
# cost = 2^(-2:-1)


# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "polynomial",
                   degree = 2,
                   ranges = list(epsilon = epsilon, cost = cost)
)


# Show Results and visualize tuning as a heatmap 
# print(tuneResult)

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 

 
 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Set up layout for side-by-side plots
par(mfrow = c(1, 3))

# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using 2nd\nDegree Polynomial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


# Best parameters 
best_error <- max(data.frame(data_wide[,-1]))

# Add polynomial kernel results to best results
best_runs <- data.frame(
  kernel = "polynomial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
)




# Fit SVM model with linear kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "linear",
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)


 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Linear Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


best_error <- max(data.frame(data_wide[,-1]))

# add linear kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "linear",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))



# Fit SVM model with radial kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Radial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)

# Reset plotting parameters
par(mfrow = c(1, 1))

best_error <- max(data.frame(data_wide[,-1]))

# add radial kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "radial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))

```
These figures show the mean absolute error for the different hyperparameters by kernel. The best performing hyperparameter combination for each kernel are marked with orange points.

```{r}


# Compare best results for each kernel by getting the row with the lowest error
best_runs <- best_runs[order(best_runs$err),]

best_run <- best_runs[1,]

cat("Best SVM kernel and parameters selected:\n")
print(best_run[,c("kernel", "epsilon", "cost")])






# Train final model with best parameters
epsilon = best_run$epsilon
cost = best_run$cost

svm_model <- svm(Cupper.Points ~ ., data = train_data,
                 cost = cost,
                 epsilon = epsilon, 
                 kernel = best_run$kernel)

# Predict on test set
predictions <- predict(svm_model, test_data)

# Evaluate model performance
mae <- mean(abs(predictions - test_data[, "Cupper.Points"]), na.rm = TRUE)
mse <- mean((predictions - test_data[, "Cupper.Points"])^2, na.rm = TRUE)
rmse <- sqrt(mse)
# print(paste("RMSE:", rmse))
print(paste("MAE:", mae))


```
The best performing kernel is the linear kernel, which means the relationship between the features and the target variable is best approximated by a multidimensional hyperplane.

```{r, echo=FALSE, fig.height=3}
# Plot the predictions vs the true values
plot(predictions, test_data[, "Cupper.Points"],
     xlab = "Predictions",
     ylab = "True Values",
     main = "Predictions vs True Values")

# get predictions for a random data point from the test data set
random_index <- sample(1:nrow(test_data), 1)
random_data <- test_data[random_index,]
random_prediction <- predict(svm_model, random_data)
# print("Random Data Point")
# print(random_data)
# print(paste("Prediction:", random_prediction))
# print(paste("True value:", random_data[, "Cupper.Points"]))

```

Here, we can see the predictions vs. the true values for the test data set. We can see that aside from two outliers, the predictions are close to the true values.



## XGBoost

# XAI and Comparison of the Models SVR and XGBoost
## Permutation Feature Importance (PFI)
**FOR SVM PFI;** The chart shows us exactly which "ingredients" the model thinks are most important for a high-quality cup of coffee. Flavor and Aftertaste are the clear winners here; when the model loses track of these, its accuracy drops the most. Balance and Acidity also help quite a bit, but they aren't as critical as the overall taste. Interestingly, all the "technical" details—like how much the Bag Weighs, the Altitude where it was grown, or the Moisture level don't effect significantly. Basically, the model has learned that if the coffee tastes great and has a clean finish, the technical background don't change the final score very much.
```{r}
# Create an "Explainer" object. Tell Dalex how to talk to the svm_model
explainer_svm <- DALEX::explain(
  model = svm_model, 
  data = test_data[,-which(names(test_data)=="Cupper.Points")], 
  y = test_data$Cupper.Points, 
  label = "SVM Coffee Model"
)


loss_mae <- function(observed, predicted){
  mean(abs(observed - predicted))
}
# Calculate Feature Importance PFI with loss_mae
pfi_results <- model_parts(
  explainer = explainer_svm,
  loss_function = loss_mae,
  type = "difference" # Shows how much the error INCREASES
)

# Visualize the results
plot(pfi_results) +
  ggtitle("Feature Importance (PFI) for Coffee Quality", 
          subtitle = "How much the error increases when a feature is shuffled") +
  theme_minimal()
```

## Partial Dependence Plot (PDP)
**FOR SVM PDP;** The plots show how the model’s quality predictions change as we look at different scores for *Acidity, Aroma, and Flavor*. The upward-sloping blue lines tell a very logical story: as these sensory scores go up, the model's predicted quality score also rises. By adding the red dots (our actual coffee samples), we can see that the model is doing a good job of finding the "middle ground" where most of the data lives. Also, we can see that the model doesn’t get distracted by the outliers, the model stays focused on the general trend, which shows it has built a very stable and reliable "logic" for what makes a good cup of coffee.
```{r}
# Calculate the Marginal Effects (PDP)
# We choose a few key features to look at (e.g., Flavor, Altitude, and Aroma)
pdp_results <- model_profile(
  explainer = explainer_svm,
  variables = c("Flavor", "Acidity", "Aroma"),
  type = "partial" # This specifies Partial Dependence
)

# 1. Prepare your raw data to match the facets
# We take the test data, select the variables in your PDP, and make them "long"
test_data_long <- test_data %>%
  select(Cupper.Points, Flavor, Acidity, Aroma) %>%
  pivot_longer(cols = c(Flavor, Acidity, Aroma), 
               names_to = "_vname_",   # This must match the internal DALEX name
               values_to = "_x_")      # This must match the internal DALEX name

# 2. Plot the PDP and overlay the points
plot(pdp_results) +
  geom_jitter(data = test_data_long, 
              aes(x = `_x_`, y = Cupper.Points), 
              alpha = 0.2, color = "darkred", width = 0.05, height = 0, inherit.aes = FALSE) +
  ggtitle("Partial Dependence Plots for Coffee Quality", 
          subtitle = "Red dots = Actual observations | Blue line = Model average") +
  theme_bw()
```

## Accumulated Local Effects (ALE)
ALE is a better option to see the feature importance in a model who has correlated features.Because The ALE plot is just a way to see how one specific feature (like Flavor) changes the final score, without getting confused by other features.
**FOR SVM ALE;**
The dashed line at zero represents an average cup of coffee in our data. When the blue line is above this zero mark, it means that specific feature is adding "bonus points" to the quality, while anything below zero means it is pulling the score down. Because the blue line slopes upward to the right, we can see the model correctly learned that higher sensory scores lead to higher quality predictions. The red dots represent the actual coffee samples, and since they cluster tightly around the blue line, it proves the model's internal logic is a very strong match for the real-world results as we have seen in PDP plots as well.
```{r}
# Calculate ALE instead of PDP
ale_results <- model_profile(
  explainer = explainer_svm,
  variables = c("Flavor", "Aroma", "Acidity"),
  type = "accumulated" # This triggers the ALE method
)

# We subtract the mean of the predictions (_yhat_) so the line centers at 0
ale_results$agr_profiles$`_yhat_` <- ale_results$agr_profiles$`_yhat_` - mean(ale_results$agr_profiles$`_yhat_`)

# 3. Use your centered dots from before
avg_cupper_points <- mean(test_data$Cupper.Points, na.rm = TRUE)

test_data_centered <- test_data %>%
  mutate(Centered_Cupper_Points = Cupper.Points - avg_cupper_points) %>%
  select(Centered_Cupper_Points, Flavor, Acidity, Aroma) %>%
  pivot_longer(cols = c(Flavor, Acidity, Aroma), 
               names_to = "_vname_",   
               values_to = "_x_")

# 4. Plot - Now they will align at zero
plot(ale_results) +
  geom_jitter(data = test_data_centered, 
              aes(x = `_x_`, y = Centered_Cupper_Points), 
              inherit.aes = FALSE, 
              alpha = 0.2, 
              color = "darkred", 
              width = 0.05, 
              height = 0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Aligned Centered ALE Plots",
       y = "Relative Effect on Cupper Points (Centered at 0)") +
  theme_bw()
```


## Shapley Additive Explanations (SHAP)
The SHAP is a Local Explanation, which acts like a "scorecard" for one specific cup of coffee to show why it received its particular prediction.

**FOR SVM SHAP;** 
In this specific example *(the number of observation is 100)*, the model's prediction was dragged down primarily because the Balance, Flavor, and Aftertaste scores were lower than average, as shown by the red bars extending to the left. On the other hand, the green bars show that this coffee’s Aroma and Clean Cup scores actually helped boost its final rating. Essentially, this plot shows exactly which characteristics made this specific cup of coffee better or worse than a typical sample.

```{r}
# 1. Select a specific coffee from your test set (e.g., the first one)
new_coffee <- test_data[100, ]

# 2. Calculate SHAP values for this specific coffee
shap_local <- predict_parts(explainer = explainer_svm, 
                             new_observation = new_coffee, 
                             type = "shap",
                             B = 25
                             ) # B is the number of re-orderings to stabilize the result

# 3. Plot the local explanation
plot(shap_local) + 
  theme_bw() +
  labs(title = "Local Explanation for a Single Coffee Sample",
       subtitle = "Contribution of each feature to the specific Cupper Points prediction")
```
