---
title: "XGBoost for Coffee Quality Prediction (Comparable to SVM)"
subtitle: "Using Identical Preprocessing and Data Split for Fair Comparison"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```


```{r config}
library(knitr)
config <- data.frame(
  Parameter = c("Random Seed", "Train/Test Split", "CV Folds (HPO)", "Early Stopping", "Max Rounds"),
  Value = c("123", "80/20", "10", "50", "2000")
)
kable(config, caption = "Configuration (Aligned with SVM)", format = "markdown")
```

# Data Preprocessing (Identical to SVM)

```{r load-libraries, message=FALSE, warning=FALSE}
library(xgboost)
library(tidyverse)
library(knitr)

SEED <- 123
MODEL_DIR <- "saved_models"
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)
```

```{r load-data}
# Read the data (same as SVM)
coffee_data <- read.csv("arabica_data_cleaned.csv", na.strings = c("", "NA"))
```

```{r define-features}
# Identical feature set from SVM - avoiding categorical variables with many unique values
columns_to_use <- c(
  "Number.of.Bags",
  "Bag.Weight",
  # "Country.of.Origin", # Excluded - too many categories
  "Processing.Method", 
  "Aroma",
  "Flavor",
  "Aftertaste",
  "Acidity",
  "Body",
  "Balance",
  "Uniformity",
  "Clean.Cup",
  "Sweetness",
  "Cupper.Points", # Target variable
  "Moisture",
  "Category.One.Defects",
  "Quakers",
  "Category.Two.Defects",
  "altitude_mean_meters" # Impute missing values from region and country origins
)

# Define sensory columns for later use in XAI
sensory_cols <- c("Aroma", "Flavor", "Aftertaste", "Acidity", "Body", 
                  "Balance", "Uniformity", "Clean.Cup", "Sweetness", 
                  "Moisture", "Category.One.Defects", "Quakers")
```



```{r bag-weight-processing}
# Get bags with weight in pounds
bags_with_weight_in_pounds_indices <- grepl("lbs", coffee_data$Bag.Weight)

# Get bags with weight in kg
bags_with_weight_in_kg_indices <- grepl("kg", coffee_data$Bag.Weight)

# Get bags with neither pounds nor kg
bags_with_neither_pounds_nor_kg_indices <- !bags_with_weight_in_pounds_indices & !bags_with_weight_in_kg_indices

# Get bags with both pounds and kg (incorrectly formatted)
bags_with_both_pounds_and_kg_indices <- bags_with_weight_in_pounds_indices & bags_with_weight_in_kg_indices

# Remove bags with both pounds and kg from individual indices
bags_with_weight_in_pounds_indices <- bags_with_weight_in_pounds_indices & !bags_with_both_pounds_and_kg_indices
bags_with_weight_in_kg_indices <- bags_with_weight_in_kg_indices & !bags_with_both_pounds_and_kg_indices

# Extract numeric value from Bag.Weight column
coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))

# Convert pounds to kg
coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] <- 
  coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] * 0.453592

# For bags with neither pounds nor kg, average the weight in pounds and kg
coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] <- 
  coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] * (1 + 0.453592) / 2
coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] <- 
  coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] * (1 + 0.453592) / 2
```



```{r processing-method}
# Fill NAs with "Unknown" (same as SVM)
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"
```



```{r train-test-split}
# Simple random 80/20 split with SAME SEED as SVM
set.seed(SEED)
n_total <- nrow(coffee_data)
n_train <- floor(0.8 * n_total)
train_index <- sample(1:n_total, n_train)

train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]
```


```{r altitude-imputation}
# Create lookup of regions and their median altitudes FROM TRAINING DATA ONLY
region_altitudes <- train_data %>%
  group_by(Region) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

region_altitudes <- region_altitudes[complete.cases(region_altitudes), ]

# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters)) 
train_data[missing_altitude_indices, "altitude_mean_meters"] <- 
  region_altitudes$altitude_median_mean_meters[match(
    train_data$Region[missing_altitude_indices], 
    region_altitudes$Region
  )]

# Impute for test data (using training set statistics)
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- 
  region_altitudes$altitude_median_mean_meters[match(
    test_data$Region[missing_altitude_indices], 
    region_altitudes$Region
  )]

# Fallback: Country of Origin median (for remaining NAs)
country_altitudes <- train_data %>%
  group_by(Country.of.Origin) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

country_altitudes <- country_altitudes[complete.cases(country_altitudes), ]

# Impute remaining NAs for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters))
train_data[missing_altitude_indices, "altitude_mean_meters"] <- 
  country_altitudes$altitude_median_mean_meters[match(
    train_data$Country.of.Origin[missing_altitude_indices], 
    country_altitudes$Country.of.Origin
  )]

# Impute remaining NAs for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- 
  country_altitudes$altitude_median_mean_meters[match(
    test_data$Country.of.Origin[missing_altitude_indices], 
    country_altitudes$Country.of.Origin
  )]
```



```{r final-prep}
# Keep only columns we're using
train_data <- train_data[, columns_to_use]
test_data <- test_data[, columns_to_use]

# Remove remaining NAs
train_data <- train_data[complete.cases(train_data), ]
test_data <- test_data[complete.cases(test_data), ]

# One-Hot Encoding (same approach as SVM)
train_data <- data.frame(model.matrix(~ ., data = train_data))[, -1]
test_data <- data.frame(model.matrix(~ ., data = test_data))[, -1]

# Verify categories align
categories_match <- all(colnames(train_data) == colnames(test_data))
```

```{r data-summary}
data_summary <- data.frame(
  Metric = c("Training samples", "Test samples", "Features (after encoding)", 
             "Categories aligned", "Mean Cupper Points (Train)", "Mean Cupper Points (Test)"),
  Value = c(nrow(train_data), nrow(test_data), ncol(train_data) - 1,
            ifelse(categories_match, "Yes", "No"),
            round(mean(train_data$Cupper.Points), 4),
            round(mean(test_data$Cupper.Points), 4))
)
kable(data_summary, caption = "Dataset Summary", format = "markdown")
```

# Baseline Models

```{r metrics-function}
get_metrics <- function(pred, obs) {
  residuals <- obs - pred
  mae <- mean(abs(residuals))
  rmse <- sqrt(mean(residuals^2))
  ss_res <- sum(residuals^2)
  ss_tot <- sum((obs - mean(obs))^2)
  r2 <- 1 - (ss_res / ss_tot)
  return(list(MAE = mae, RMSE = rmse, R2 = r2))
}
```

```{r baselines}
y_train <- train_data$Cupper.Points
y_test <- test_data$Cupper.Points
X_train <- as.matrix(train_data[, colnames(train_data) != "Cupper.Points"])
X_test <- as.matrix(test_data[, colnames(test_data) != "Cupper.Points"])

# Mean Predictor Baseline
preds_mean <- rep(mean(y_train), length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)

# Linear Regression Baseline
model_lm <- lm(Cupper.Points ~ ., data = train_data)
preds_lm <- predict(model_lm, newdata = test_data)
metrics_lm <- get_metrics(preds_lm, y_test)
```

```{r baseline-table}
baseline_results <- data.frame(
  Model = c("Mean Predictor", "Linear Regression"),
  MAE = c(round(metrics_mean$MAE, 4), round(metrics_lm$MAE, 4)),
  RMSE = c(round(metrics_mean$RMSE, 4), round(metrics_lm$RMSE, 4)),
  R2 = c(round(metrics_mean$R2, 4), round(metrics_lm$R2, 4))
)
kable(baseline_results, caption = "Baseline Model Performance", format = "markdown")
```

```{r lm-coefficients}
lm_coefs <- summary(model_lm)$coefficients
lm_coefs_df <- data.frame(
  Feature = rownames(lm_coefs),
  Estimate = round(lm_coefs[, 1], 4),
  Std_Error = round(lm_coefs[, 2], 4),
  P_Value = format(lm_coefs[, 4], digits = 3)
)
kable(head(lm_coefs_df[order(abs(lm_coefs_df$Estimate), decreasing = TRUE), ], 8), 
      caption = "Top Linear Regression Coefficients", format = "markdown", row.names = FALSE)
```

# XGBoost Hyperparameter Optimization

Using 10-fold cross-validation.

```{r xgb-hpo}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest <- xgb.DMatrix(data = X_test, label = y_test)

# Grid search (similar scope to SVM's epsilon/cost grid)
tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1),
  max_depth = c(3, 5, 7),
  subsample = c(0.7, 0.8, 0.9),
  colsample_bytree = c(0.7, 0.8, 0.9)
)

# Run HPO with 10-fold CV (matching SVM's tune() default)
hpo_results <- data.frame()

for (i in 1:nrow(tune_grid)) {
  params <- list(
    objective = "reg:squarederror",
    eval_metric = "mae",  # Using MAE to match SVM's primary metric
    eta = tune_grid$eta[i],
    max_depth = tune_grid$max_depth[i],
    subsample = tune_grid$subsample[i],
    colsample_bytree = tune_grid$colsample_bytree[i],
    nthread = 1
  )
  
  set.seed(SEED)
  cv_result <- xgb.cv(
    params = params,
    data = dtrain,
    nrounds = 2000,
    nfold = 10,  # Match SVM's default CV folds
    early_stopping_rounds = 50,
    verbose = 0
  )
  
  # Extract best iteration and MAE
  eval_log <- cv_result$evaluation_log
  mae_col <- grep("test.*mae.*mean", names(eval_log), value = TRUE)[1]
  best_iter <- cv_result$best_iteration
  if (is.null(best_iter) || length(best_iter) == 0) {
    best_iter <- which.min(eval_log[[mae_col]])
  }
  best_mae <- eval_log[[mae_col]][best_iter]
  
  hpo_results <- rbind(hpo_results, data.frame(
    eta = params$eta,
    max_depth = params$max_depth,
    subsample = params$subsample,
    colsample_bytree = params$colsample_bytree,
    best_nrounds = best_iter,
    cv_mae = best_mae
  ))
}

# Find best parameters
best_params <- hpo_results[which.min(hpo_results$cv_mae), ]
```

```{r hpo-results}
kable(best_params, caption = "Optimal XGBoost Hyperparameters", format = "markdown", digits = 4, row.names = FALSE)
```

# Train XGBoost Model

```{r train-final}
final_params <- list(
  objective = "reg:squarederror",
  eval_metric = "mae",
  eta = best_params$eta,
  max_depth = best_params$max_depth,
  subsample = best_params$subsample,
  colsample_bytree = best_params$colsample_bytree,
  nthread = 1
)

set.seed(SEED)
model_xgb <- xgb.train(
  params = final_params,
  data = dtrain,
  nrounds = best_params$best_nrounds,
  verbose = 0
)

# Predictions
preds_xgb <- predict(model_xgb, dtest)
metrics_xgb <- get_metrics(preds_xgb, y_test)
```

```{r save-model}
# Save the trained model for further analysis
model_file <- file.path(MODEL_DIR, "model_xgb_comparable.rds")
saveRDS(model_xgb, model_file)
```

Model saved to: `r model_file`

# Learning Curves

```{r learning-curves}
# Learning curves show how training and validation error evolve with more boosting rounds.
# A widening gap between curves indicates overfitting - the model is memorizing training data.
train_mae <- numeric()
test_mae <- numeric()
max_rounds <- min(500, best_params$best_nrounds * 2)

for (r in seq(10, max_rounds, by = 10)) {
  set.seed(SEED)
  temp_model <- xgb.train(params = final_params, data = dtrain, nrounds = r, verbose = 0)
  train_mae <- c(train_mae, mean(abs(predict(temp_model, dtrain) - y_train)))
  test_mae <- c(test_mae, mean(abs(predict(temp_model, dtest) - y_test)))
}

training_history <- data.frame(
  iter = seq(10, max_rounds, by = 10), 
  train_mae = train_mae, 
  test_mae = test_mae
)

learning_curve_df <- training_history %>%
  pivot_longer(cols = c(train_mae, test_mae), names_to = "Dataset", values_to = "MAE") %>%
  mutate(Dataset = ifelse(Dataset == "train_mae", "Training", "Validation"))
```

```{r learning-curves-plot, fig.cap="Learning curve for XGBoost model. The gap between training (blue) and validation (red) curves indicates overfitting. The vertical dashed line marks the optimal stopping point determined by early stopping.", fig.height=4}
ggplot(learning_curve_df, aes(x = iter, y = MAE, color = Dataset)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = best_params$best_nrounds, linetype = "dashed", color = "darkgreen", linewidth = 1) +
  annotate("text", x = best_params$best_nrounds + 10, y = max(learning_curve_df$MAE) * 0.9,
           label = paste("Optimal =", best_params$best_nrounds), hjust = 0, color = "darkgreen") +
  scale_color_manual(values = c("Training" = "#2E86AB", "Validation" = "#E63946")) +
  labs(x = "Boosting Rounds", y = "MAE", title = "Learning Curves") +
  theme_minimal()
```

```{r learning-curve-metrics}
best_row <- which.min(abs(training_history$iter - best_params$best_nrounds))
if (length(best_row) == 0) best_row <- nrow(training_history)
final_train_mae <- training_history$train_mae[best_row]
final_test_mae <- training_history$test_mae[best_row]
overfitting_gap <- final_test_mae - final_train_mae
gap_pct <- round(100 * overfitting_gap / final_test_mae, 2)
```

> **Overfitting Analysis:** The training-validation gap is `r round(overfitting_gap, 4)` MAE (`r gap_pct`% of validation MAE). This gap indicates whether the model is memorizing training data rather than learning generalizable patterns.

# Results

```{r final-results}
final_results <- data.frame(
  Model = c("Mean Predictor", "Linear Regression", "XGBoost"),
  MAE = c(metrics_mean$MAE, metrics_lm$MAE, metrics_xgb$MAE),
  RMSE = c(metrics_mean$RMSE, metrics_lm$RMSE, metrics_xgb$RMSE),
  R2 = c(metrics_mean$R2, metrics_lm$R2, metrics_xgb$R2)
)

final_results$MAE <- round(final_results$MAE, 4)
final_results$RMSE <- round(final_results$RMSE, 4)
final_results$R2 <- round(final_results$R2, 4)

kable(final_results, caption = "Final Model Comparison (Test Set)", format = "markdown")
```




# Predictions vs True Values

```{r pred-vs-true, fig.height=4, fig.cap="Predictions vs True Values for XGBoost model. Points close to the diagonal line indicate accurate predictions."}
plot(preds_xgb, y_test,
     xlab = "Predictions",
     ylab = "True Values",
     main = "XGBoost: Predictions vs True Values",
     pch = 19,
     col = rgb(0.2, 0.4, 0.6, 0.5))
abline(0, 1, col = "red", lwd = 2)
```

# Feature Importance

```{r feature-importance, fig.height=5, fig.cap="Top 15 features by importance (Gain). Higher gain indicates the feature contributes more to reducing prediction error."}
importance <- xgb.importance(feature_names = colnames(X_train), model = model_xgb)
xgb.plot.importance(importance[1:min(15, nrow(importance)), ], main = "XGBoost Feature Importance (Gain)")
```

```{r importance-table}
kable(head(importance, 10), caption = "Top 10 Features by Importance", format = "markdown", digits = 4)
```

# Approaches Found Less Effective

## Lambda and Min_Child_Weight Regularization

XGBoost offers explicit regularization through lambda (L2) and min_child_weight parameters. Testing showed no improvement because: (1) implicit regularization through learning rate, max depth, subsample, and early stopping already prevents overfitting; (2) the ~1300 samples with sensory features don't exhibit severe overfitting; (3) minimal train-test gap in base configuration.

## Bayesian Optimization vs Grid Search

Bayesian Optimization did not outperform Grid Search because: (1) only 4 parameters with reasonable bounds; (2) many configurations achieve similar results (performance plateau); (3) surrogate model overhead without finding better solutions. Grid Search is sufficient for this problem size.

# Explainable AI: PFI and PDP Analysis

This section applies model-agnostic XAI techniques to interpret the XGBoost model.

```{r xai-setup}
if (!require("DALEX")) install.packages("DALEX", quiet = TRUE)
library(DALEX)

predict_xgb <- function(model, newdata) {
  if (is.data.frame(newdata)) newdata <- as.matrix(newdata)
  dmatrix <- xgb.DMatrix(data = newdata)
  predict(model, dmatrix)
}

explainer_xgb <- DALEX::explain(
  model = model_xgb, 
  data = X_train, 
  y = y_train,
  predict_function = predict_xgb, 
  label = "XGBoost", 
  verbose = FALSE
)
```

## Permutation Feature Importance (PFI)

PFI measures how much model performance degrades when a feature's relationship with the target is broken by randomly shuffling its values. Higher importance = larger performance drop = more important feature.

```{r pfi-calculation}
# PFI is model-agnostic and measures actual predictive reliance.
# Unlike XGBoost's Gain metric, PFI accounts for feature interactions and correlations.
loss_mae <- function(observed, predicted) mean(abs(observed - predicted))
set.seed(SEED)
pfi_xgb <- model_parts(
  explainer = explainer_xgb, 
  loss_function = loss_mae, 
  type = "difference", 
  B = 10
)
```

```{r pfi-plot, fig.cap="Permutation Feature Importance for XGBoost model. Bars show average MAE increase when each feature is shuffled (with error bars for standard deviation across 10 permutations). Longer bars = more important features.", fig.width=10, fig.height=7}
plot(pfi_xgb, max_vars = 15)
```

```{r pfi-table}
pfi_results <- pfi_xgb %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(Mean_Dropout_Loss = mean(dropout_loss), SD = sd(dropout_loss), .groups = "drop") %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(Importance = Mean_Dropout_Loss - min(pfi_xgb$dropout_loss[pfi_xgb$variable == "_full_model_"]),
         Rank = row_number())
kable(head(pfi_results[, c("Rank", "variable", "Importance", "SD")], 10),
      col.names = c("Rank", "Feature", "MAE Increase", "SD"),
      caption = "Top 10 Features by Permutation Feature Importance", format = "markdown", digits = 4)
```

> **PFI Interpretation:** The top features by PFI are sensory attributes. This confirms that expert taste evaluation drives predictions and the model uses sensible features. Note that PFI measures *predictive reliance*, not causal importance.

## Partial Dependence Plots (PDP)

PDP shows the **marginal effect** of a feature on predictions, averaging over all other features.

```{r pdp-calculation}
# PDP reveals how the model responds to changes in each feature.
# Expected pattern: higher sensory scores → higher predicted Cupper Points
sensory_features_in_model <- intersect(sensory_cols, colnames(X_train))
top_sensory_pfi <- pfi_results %>%
  filter(variable %in% sensory_features_in_model) %>%
  head(4) %>%
  pull(variable)

pdp_list <- list()
for (feat in top_sensory_pfi) {
  pdp_list[[feat]] <- model_profile(
    explainer = explainer_xgb, 
    variables = feat, 
    type = "partial", 
    N = 300
  )
}
```

```{r pdp-combined, fig.cap="Partial Dependence Plots for top sensory features. Each curve shows how predicted Cupper Points change as that feature varies (holding others constant). All show positive relationships: higher sensory scores lead to higher quality predictions.", fig.width=10, fig.height=5}
pdp_combined_data <- data.frame()
for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_combined_data <- rbind(pdp_combined_data, data.frame(
    Feature = feat, 
    x = pdp_data$`_x_`, 
    yhat = pdp_data$`_yhat_`
  ))
}

ggplot(pdp_combined_data, aes(x = x, y = yhat, color = Feature)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Feature, scales = "free_x") +
  labs(title = "Partial Dependence Plots - Top Sensory Features", 
       x = "Feature Value", 
       y = "Avg Predicted Cupper Points") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "none")
```

```{r pdp-variance}
pdp_variance_importance <- data.frame(
  Feature = character(), 
  PDP_Variance = numeric(), 
  PDP_Range = numeric(), 
  stringsAsFactors = FALSE
)

for (feat in top_sensory_pfi) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  pdp_variance_importance <- rbind(pdp_variance_importance, data.frame(
    Feature = feat, 
    PDP_Variance = var(pdp_data$`_yhat_`),
    PDP_Range = max(pdp_data$`_yhat_`) - min(pdp_data$`_yhat_`)
  ))
}

pdp_variance_importance <- pdp_variance_importance %>% arrange(desc(PDP_Variance))
kable(pdp_variance_importance, 
      col.names = c("Feature", "PDP Variance", "PDP Range"),
      caption = "Feature Importance from PDP (Variance-based)", 
      format = "markdown", digits = 4)
```

## Why PFI and PDP Rankings Differ

```{r pfi-pdp-comparison}
rank_corr <- cor(1:4, match(pdp_variance_importance$Feature, head(pfi_results$variable, 4)), method = "spearman")
```

The Spearman correlation between PFI and PDP rankings is **`r round(rank_corr, 2)`**, indicating potential disagreement. This is expected because PFI and PDP measure fundamentally different things:

- **PFI** asks: "How much does model performance degrade when this feature's signal is destroyed?"
- **PDP variance** asks: "How much does the average prediction change across this feature's range?"

Features with large PDP range may have lower PFI if most observations cluster in a narrow range where the PDP is relatively flat. Conversely, features with smaller PDP range may have higher PFI if the model relies on them consistently across the data distribution.

# Summary

**Key Results:**

- XGBoost MAE: `r round(metrics_xgb$MAE, 4)`
- XGBoost RMSE: `r round(metrics_xgb$RMSE, 4)`
- XGBoost R²: `r round(metrics_xgb$R2, 4)`

**Model saved to:** `r model_file`

---

*Report generated: `r Sys.Date()`*
