---
title: "Coffee Quality Prediction: SVM & XGBoost Analysis"
date: "`r Sys.Date()`"
fontsize: 11pt

output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex

header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{multicol}
  - \usepackage{caption}
  - \captionsetup[table]{skip=3pt}
  - \captionsetup[figure]{skip=5pt}
  - \setlength{\tabcolsep}{4pt}
  - \renewcommand{\arraystretch}{0.9}
  - \setlength{\parskip}{3pt}
  - \setlength{\floatsep}{8pt}
  - \setlength{\textfloatsep}{8pt}

geometry: margin=0.85in
---

```{r setup, include=FALSE}
# Optional: Clear environment before running (uncomment if needed)
# rm(list = ls())

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 3.5, fig.height = 2.5, out.width = "50%",
  fig.pos = "H"
)

# Model directory for saving/loading trained models
MODEL_DIR <- "saved_models"
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)
```

# Support Vector Regression

## Data Preprocessing and Cleaning 

### Feature selection:
In SVM Regression, categorical variables are One-Hot Encoded using dummy variables.
This means, features with many unique values will lead to a sparse feature matrix. This is disadvantageous as it increases computational cost and poses the risk of overfitting. 
Additionally, categories with only a few or even only one observation most likely will not add to model robustness, but are adding noise to the signal.
Therefore, we are removing categorical features with many unique values, only keeping columns with a low number of unique values.
E.g. we are not including the "Region" column with 344 unique values in the model training, because we'd on average have less than 4 observations per Region. 
For features with a high number of missing values, we either remove the column or impute the missing values.
For instance for the "Variety" column, we have about 15% missing values and 8% entries "other" at 30 different varieties. 
Keeping those, would mean adding 30 sparse columns, and at the same time the information gain is most likely not too high with this many missing values.
Therefore, we are not including this feature. 


### Imputation:
For the altitude column, we have about 17% missing values. Here, we are imputing the missing values from the "Region" column.
For each observation with missing altitude, we are imputing the median altitude of all coffees from the same region.
This covers almost all missing values. For the remaining observations, we are imputing the mean altitude of all coffees from the same country of origin.
As the "Region" and "Country" columns are not used in the model training, we are not creating correlated features with this imputation.
We are imputing the missing values after the train/test split, using only values from the training set, to avoid data leakage from the test set into the training set.


### Further preprocessing:
The bag weight column contains both pounds and kilograms, so we are converting all to kilograms. 
However, for some entries, it is unclear which unit is used. For these, we are averaging the weight in pounds and kilograms.
This means for those entries the weight is certainly incorrect, but should be within an order of magnitude of the correct weight. 
If we can assume that speciality coffee is sold in smaller quantities, we can still gain valuable information from the weight.


### Scaling: 
The SVM method used for model training has a built in scaling functionality. 
Both predictor and target variables are scaled to have expected value 0 and variance 1. [https://rdrr.io/rforge/e1071/man/svm.html]
This transformation is based on the training data and applied to train and test data sets.


```{r svm-preprocessing, echo=FALSE, message=FALSE, warning=FALSE}
# Read the data
coffee_data <- read.csv("arabica_data_cleaned.csv", na.strings = c("", "NA"))



# Avoid categorical variables with many unique values because they will create many dummy variables. 
columns_to_use <- c(
  "Number.of.Bags",
  "Bag.Weight",
  # "Country.of.Origin", Increased runtime too much
  "Processing.Method", 
  "Aroma",
  "Flavor",
  "Aftertaste",
  "Acidity",
  "Body",
  "Balance",
  "Uniformity",
  "Clean.Cup",
  "Sweetness",
  "Cupper.Points", # Target variable
  "Moisture",
  "Category.One.Defects",
  "Quakers",
  "Category.Two.Defects",
  "altitude_mean_meters" # Impute missing values from region and country origins
)




## Bag Weight

# Get bags with weight in pounds
bags_with_weight_in_pounds_indices <- grepl("lbs", coffee_data$Bag.Weight)

# Get bags with weight in kg
bags_with_weight_in_kg_indices <- grepl("kg", coffee_data$Bag.Weight)

# Get bags with neither pounds nor kg
bags_with_neither_pounds_nor_kg_indices <- !bags_with_weight_in_pounds_indices & !bags_with_weight_in_kg_indices

# Get bags with both pounds and kg (incorrectly formatted)
bags_with_both_pounds_and_kg_indices <- bags_with_weight_in_pounds_indices & bags_with_weight_in_kg_indices

# Remove bags with both pounds and kg from bags with weight indices and bags with weight in kg indices
bags_with_weight_in_pounds_indices <-
  bags_with_weight_in_pounds_indices & !bags_with_both_pounds_and_kg_indices

bags_with_weight_in_kg_indices <-
  bags_with_weight_in_kg_indices & !bags_with_both_pounds_and_kg_indices


# Sanity check
# print(paste("Number of bags with weight in pounds:", sum(bags_with_weight_in_pounds_indices)))
# print(paste("Number of bags with weight in kg:", sum(bags_with_weight_in_kg_indices)))
# print(paste("Number of bags with neither pounds nor kg:", sum(bags_with_neither_pounds_nor_kg_indices)))
# print(paste("Number of bags with both pounds and kg:", sum(bags_with_both_pounds_and_kg_indices)))
# print(paste("Number of bags with weight:", length(coffee_data$Bag.Weight)))

# Extract numeric value from Bag.Weight column
coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
# Multiply bags with weight in pounds by 0.453592 to convert to kg
coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] <- coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] * 0.453592

# For bags with neither pounds nor kg, average the weight in pounds and kg
coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] <- coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] * (1 + 0.453592)/2
coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] <- coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] * (1 + 0.453592)/2



# For the Processing Method fill NAs with "Unknown"
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"



# Split data into training and test set
# Separate features and target


# Simple random 80/20 split
SEED <- 123
set.seed(SEED)
n_total <- nrow(coffee_data)
n_train <- floor(0.8 * n_total)
train_index <- sample(1:n_total, n_train)

train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]



## Altitude

library(tidyverse)


# Create dict of regions and their median altitudes
region_altitudes <- train_data %>%
  group_by(Region) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

region_altitudes <- region_altitudes[complete.cases(region_altitudes),]

# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters)) 
train_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(train_data$Region[missing_altitude_indices], region_altitudes$Region)]


# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(test_data$Region[missing_altitude_indices], region_altitudes$Region)]




### Repeat imputation for country of origin

# Create dict of regions and their median altitudes
country_altitudes <- train_data %>%
  group_by(Country.of.Origin) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

country_altitudes <- country_altitudes[complete.cases(country_altitudes),]


# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters))
train_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(train_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]

# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(test_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]




# Remove columns we are not using
train_data <- train_data[,columns_to_use]
test_data <- test_data[,columns_to_use]


# Remove remaining NAs
train_data <- train_data[complete.cases(train_data),]
test_data <- test_data[complete.cases(test_data),]


# Add One Hot encoding
train_data <- data.frame(model.matrix(~ ., data = train_data))[,-1]
test_data  <- data.frame(model.matrix(~ ., data = test_data))[,-1]

# Check if categories align between train and test 
# print("All Categories match between train and Test data")
# print(all(colnames(train_data) == colnames(test_data)))
  

```

```{r svm-data-info, echo=FALSE}
# print(paste0(
#   "Number of training samples: ", nrow(train_data), "\n",
#   "Number of test samples: ", nrow(test_data), "\n",
#   "Number of features: ", ncol(train_data), "\n",
#   "Average cupper points in training set: ", mean(train_data$Cupper.Points), "\n",
#   "Average cupper points in test set: ", mean(test_data$Cupper.Points), "\n",
#   "Features used: ", paste(columns_to_use, collapse = ", ")

# ))

```

## Support Vector Regression - Hyperparameter Optimization

For the prediction of cupper points we train an SVM for regression on the data set, holding out the test data set for final evaluation. 
To choose appropriate hyperparameters, we conduct a hyperparameter optimization on the training data set using 10-fold cross-validation. 
We use a grid search for different values for cost and epsilon for a radial, linear, and 2-degree polynomial kernel. We compare the mean absolute error between runs to choose the best-performing hyperparameter combination. Finally, we train a model with these hyperparameters on the entire training set and evaluate on the test data set.


```{r svm-hyperparameter-tuning, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.5, fig.width=8, out.width="95%"}




####  Fit SVM model with polynomial kernel (degree 2) 
library(e1071)

# Gridsearch 
epsilon = seq(0,2,0.1)
cost = 2^(-10:3)
# epsilon = seq(0,1,0.5)
# cost = 2^(-2:-1)


# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "polynomial",
                   degree = 2,
                   ranges = list(epsilon = epsilon, cost = cost)
)


# Show Results and visualize tuning as a heatmap 
# print(tuneResult)
library(ggplot2)

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 
 library(dplyr)
 library(tidyr)
 
 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Set up layout for side-by-side plots
par(mfrow = c(1, 3))

# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using 2nd\nDegree Polynomial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


# Best parameters 
best_error <- max(data.frame(data_wide[,-1]))

# Add polynomial kernel results to best results
best_runs <- data.frame(
  kernel = "polynomial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
)




# Fit SVM model with linear kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "linear",
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)


 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Linear Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


best_error <- max(data.frame(data_wide[,-1]))

# add linear kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "linear",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))



# Fit SVM model with radial kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Radial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)

# Reset plotting parameters
par(mfrow = c(1, 1))

best_error <- max(data.frame(data_wide[,-1]))

# add radial kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "radial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))

```
These figures show the mean absolute error for the different hyperparameters by kernel. The best performing hyperparameter combination for each kernel are marked with orange points.

```{r svm-final-model}
# Check if saved model exists
svm_model_file <- file.path(MODEL_DIR, "model_svm.rds")

if (file.exists(svm_model_file)) {
  # Load existing model
  svm_model <- readRDS(svm_model_file)
  # cat("SVM model loaded from:", svm_model_file, "\n")
  
  # Still need to determine best_run for display
  best_runs <- best_runs[order(best_runs$err),]
  best_run <- best_runs[1,]
} else {
  # Compare best results for each kernel by getting the row with the lowest error
  best_runs <- best_runs[order(best_runs$err),]
  best_run <- best_runs[1,]
  
  # Train final model with best parameters
  svm_model <- svm(Cupper.Points ~ ., data = train_data,
                   cost = best_run$cost,
                   epsilon = best_run$epsilon, 
                   kernel = as.character(best_run$kernel))
  
  # Save model for future runs
  saveRDS(svm_model, svm_model_file)
  cat("SVM model trained and saved to:", svm_model_file, "\n")
}

cat("Best SVM kernel and parameters selected:\n")
print(best_run[,c("kernel", "epsilon", "cost")])

# Predict on test set
svm_predictions <- predict(svm_model, test_data)

# Evaluate model performance
svm_mae <- mean(abs(svm_predictions - test_data[, "Cupper.Points"]), na.rm = TRUE)
svm_mse <- mean((svm_predictions - test_data[, "Cupper.Points"])^2, na.rm = TRUE)
svm_rmse <- sqrt(svm_mse)
svm_ss_res <- sum((test_data[, "Cupper.Points"] - svm_predictions)^2)
svm_ss_tot <- sum((test_data[, "Cupper.Points"] - mean(test_data[, "Cupper.Points"]))^2)
svm_r2 <- 1 - (svm_ss_res / svm_ss_tot)

print(paste("MAE:", svm_mae))
```
The best performing kernel is the linear kernel, which means the relationship between the features and the target variable is best approximated by a multidimensional hyperplane.

```{r svm-predictions-plot, echo=FALSE, fig.height=2.8, fig.width=4, out.width="50%"}
# Plot the predictions vs the true values
plot(svm_predictions, test_data[, "Cupper.Points"],
     xlab = "Predictions",
     ylab = "True Values",
     main = "SVM: Predictions vs True Values")

# get predictions for a random data point from the test data set
random_index <- sample(1:nrow(test_data), 1)
random_data <- test_data[random_index,]
random_prediction <- predict(svm_model, random_data)
# print("Random Data Point")
# print(random_data)
# print(paste("Prediction:", random_prediction))
# print(paste("True value:", random_data[, "Cupper.Points"]))

```

Here, we can see the predictions vs. the true values for the test data set. We can see that aside from two outliers, the predictions are close to the true values.

```{r save-svm-results-for-comparison}
# Save SVM results for use in Part 3 comparison
# These variables will be used after XGBoost training
svm_results <- list(
  mae = svm_mae,
  rmse = svm_rmse,
  r2 = svm_r2,
  predictions = svm_predictions,
  test_cupper_points = test_data[, "Cupper.Points"],
  best_run = best_run
)
```

# XGBoost Model

```{r xgb-load-libraries}
library(xgboost)   # Core gradient boosting algorithm
library(dplyr)     # Data manipulation (group_by, summarise, mutate, %>%)
library(ggplot2)   # Visualization
library(tidyr)     # Data reshaping (pivot_longer)
library(knitr)     # Table formatting
library(gridExtra) # Multi-panel plot layouts

# Global Constants
SEED       <- 123
CV_FOLDS   <- 10
EARLY_STOP <- 50
MAX_ROUNDS <- 2000
MODEL_DIR  <- "saved_models"

if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)
```

## Data Preprocessing

To ensure that the SVM and XGBoost models are comparable, we apply the same data preprocessing described above to XGBoost, except we skip feature scaling. 

SVM requires scaling because it calculates distances between data points to find the optimal separating hyperplane. Features with larger ranges would dominate the distance calculations, which would make the model incorrect. XGBoost is a tree-based method, it is only important if values are higher or lower (e.g., “is Flavor > 7.5?”), so scaling is unnecessary.

```{r xgb-load-data}
coffee_data <- read.csv("arabica_data_cleaned.csv", na.strings = c("", "NA"))
```

```{r xgb-define-features}
columns_to_use <- c(
  "Number.of.Bags", "Bag.Weight", "Processing.Method", 
  "Aroma", "Flavor", "Aftertaste", "Acidity", "Body", "Balance",
  "Uniformity", "Clean.Cup", "Sweetness", "Cupper.Points",
  "Moisture", "Category.One.Defects", "Quakers", "Category.Two.Defects",
  "altitude_mean_meters"
)
```

```{r xgb-bag-weight-processing}
bags_lbs <- grepl("lbs", coffee_data$Bag.Weight)
bags_kg <- grepl("kg", coffee_data$Bag.Weight)
bags_neither <- !bags_lbs & !bags_kg
bags_both <- bags_lbs & bags_kg
bags_lbs <- bags_lbs & !bags_both
bags_kg <- bags_kg & !bags_both

coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
coffee_data$Bag.Weight[bags_lbs] <- coffee_data$Bag.Weight[bags_lbs] * 0.453592
coffee_data$Bag.Weight[bags_neither] <- coffee_data$Bag.Weight[bags_neither] * (1 + 0.453592) / 2
coffee_data$Bag.Weight[bags_both] <- coffee_data$Bag.Weight[bags_both] * (1 + 0.453592) / 2
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"
```

```{r xgb-train-test-split}
set.seed(SEED)
n_total <- nrow(coffee_data)
train_index <- sample(1:n_total, floor(0.8 * n_total))
train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]
```

```{r xgb-altitude-imputation}
# Hierarchical imputation: Region median → Country median (training data only)
region_alt <- train_data %>% group_by(Region) %>%
  summarise(alt_med = median(altitude_mean_meters, na.rm = TRUE)) %>% filter(!is.na(alt_med))

for (df_name in c("train_data", "test_data")) {
  df <- get(df_name)
  missing <- which(is.na(df$altitude_mean_meters))
  df$altitude_mean_meters[missing] <- region_alt$alt_med[match(df$Region[missing], region_alt$Region)]
  assign(df_name, df)
}

country_alt <- train_data %>% group_by(Country.of.Origin) %>%
  summarise(alt_med = median(altitude_mean_meters, na.rm = TRUE)) %>% filter(!is.na(alt_med))

for (df_name in c("train_data", "test_data")) {
  df <- get(df_name)
  missing <- which(is.na(df$altitude_mean_meters))
  df$altitude_mean_meters[missing] <- country_alt$alt_med[match(df$Country.of.Origin[missing], country_alt$Country.of.Origin)]
  assign(df_name, df)
}
```

```{r xgb-final-prep}
train_data <- train_data[, columns_to_use]
test_data <- test_data[, columns_to_use]
train_data <- train_data[complete.cases(train_data), ]
test_data <- test_data[complete.cases(test_data), ]
train_data <- data.frame(model.matrix(~ ., data = train_data))[, -1]
test_data <- data.frame(model.matrix(~ ., data = test_data))[, -1]
```

## Configuration & Data

```{r xgb-config-data-table, results='asis'}
cat("\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{ll}
\\toprule
\\textbf{Config} & \\textbf{Value} \\\\
\\midrule
Seed & ", SEED, " \\\\
Split & 80/20 \\\\
CV Folds & ", CV_FOLDS, " \\\\
Early Stop & ", EARLY_STOP, " \\\\
Max Rounds & ", MAX_ROUNDS, " \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}
\\hfill
\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{lr}
\\toprule
\\textbf{Data} & \\textbf{Value} \\\\
\\midrule
Train n & ", nrow(train_data), " \\\\
Test n & ", nrow(test_data), " \\\\
Features & ", ncol(train_data) - 1, " \\\\
Target $\\bar{x}$ & ", round(mean(train_data$Cupper.Points), 2), " \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}
\\hfill
\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{lrrr}
\\toprule
\\textbf{Model} & \\textbf{MAE} & \\textbf{R²} \\\\
\\midrule
Mean & 0.309 & 0.00 \\\\
Linear & 0.146 & 0.59 \\\\
XGBoost & 0.141 & 0.68 \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}

\\vspace{6pt}")
```

```{r xgb-metrics-function}
get_metrics <- function(pred, obs) {
  res <- obs - pred
  list(MAE = mean(abs(res)), RMSE = sqrt(mean(res^2)), R2 = 1 - sum(res^2)/sum((obs - mean(obs))^2))
}
```

```{r xgb-baselines}
y_train <- train_data$Cupper.Points
y_test <- test_data$Cupper.Points
X_train <- as.matrix(train_data[, colnames(train_data) != "Cupper.Points"])
X_test <- as.matrix(test_data[, colnames(test_data) != "Cupper.Points"])

preds_mean <- rep(mean(y_train), length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)

model_lm <- lm(Cupper.Points ~ ., data = train_data)
preds_lm <- predict(model_lm, newdata = test_data)
metrics_lm <- get_metrics(preds_lm, y_test)
```

**Baseline models**

We use two baseline models: the Mean Predictor and Linear Regression.

The **Mean Predictor** has the $R^2$ = 0, which confirms that our target has meaningful variance to explain. The MAE of 0.309 means that without any features, we'd be off by about 0.3 points on average.

**Linear Regression** provides an interpretable baseline with $R^2$ of 0.59, showing that 59% of the variance in cupping scores can be explained by linear combinations of features.
Any improvement from the XGBoost models means capturing non-linear relationships that linear regression misses.

## Hyperparameter Optimization

```{r xgb-hpo}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest <- xgb.DMatrix(data = X_test, label = y_test)

tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1), max_depth = c(3, 5, 7),
  subsample = c(0.7, 0.8, 0.9), colsample_bytree = c(0.7, 0.8, 0.9)
)

hpo_results <- data.frame()
for (i in 1:nrow(tune_grid)) {
  params <- list(objective = "reg:squarederror", eval_metric = "mae",
    eta = tune_grid$eta[i], max_depth = tune_grid$max_depth[i],
    subsample = tune_grid$subsample[i], colsample_bytree = tune_grid$colsample_bytree[i], nthread = 1)
  
  set.seed(SEED)
  cv_result <- xgb.cv(params = params, data = dtrain, nrounds = MAX_ROUNDS,
    nfold = CV_FOLDS, early_stopping_rounds = EARLY_STOP, verbose = 0)
  
  eval_log <- cv_result$evaluation_log
  mae_col <- grep("test.*mae.*mean", names(eval_log), value = TRUE)[1]
  best_iter <- cv_result$best_iteration
  if (is.null(best_iter) || length(best_iter) == 0) best_iter <- which.min(eval_log[[mae_col]])
  
  hpo_results <- rbind(hpo_results, data.frame(
    eta = params$eta, max_depth = params$max_depth, subsample = params$subsample,
    colsample_bytree = params$colsample_bytree, nrounds = best_iter, cv_mae = eval_log[[mae_col]][best_iter]))
}
best_params <- hpo_results[which.min(hpo_results$cv_mae), ]
```

```{r xgb-hpo-table, results='asis'}
cat("\\begin{center}
\\scriptsize
\\begin{tabular}{cccccc}
\\toprule
eta & max\\_depth & subsample & colsample & rounds & CV MAE \\\\
\\midrule
", best_params$eta, " & ", best_params$max_depth, " & ", best_params$subsample, " & ", 
best_params$colsample_bytree, " & ", best_params$nrounds, " & ", round(best_params$cv_mae, 3), " \\\\
\\bottomrule
\\end{tabular}
\\end{center}")
```

**Hyperparameter selection**

We tune four main parameters to balance bias and variance:

**eta (learning rate):** 0.01–0.1. Smaller values make each tree contribute less, reducing overfitting but needing more rounds. 
    
**max_depth:** 3–7. Controls tree complexity. Shallow trees capture simple patterns; deeper trees can model complex patterns but may overfit, especially on our dataset with about 1000 samples.
    
**subsample:** 0.7–0.9. Fraction of rows used per tree. Reduces overfitting by adding randomness. We avoid below 0.7, as it would leave too few samples per tree.
    
**colsample_bytree:** 0.7–0.9. Fraction of features per tree. Prevents any one feature from dominating.
    
These ranges are conservative. With about 1000 samples, extreme values (e.g., eta=0.3, max_depth=10) would overfit, but mild regularization is enough. 

**Lambda and Min_Child_Weight Regularization**  
We tested additional hyperparameters, but they were ineffective. XGBoost’s lambda (L2) and min_child_weight provided no benefit because implicit regularization from learning rate, max_depth, subsample, and early stopping already prevents overfitting. Extra regularization slightly worsened the metrics, so it was omitted.

**Bayesian Optimization vs Grid Search**  
We also tested Bayesian Optimization, but it was ineffective. With only 4 parameters and a small search space, it added overhead, slowed hyperparameter tuning, and offered no meaningful improvement. Grid search is sufficient for this dataset and model.

**10-fold CV:** Matches SVM for fair comparison. Each fold has about 100 samples, giving stable MAE estimates. Fewer folds increase variance; leave-one-out is too slow.

**Early stopping (50 rounds):** Stops training if validation MAE doesn’t improve for 50 rounds, automatically choosing the optimal number of trees and avoiding overfitting.


```{r xgb-train-final}
xgb_model_file <- file.path(MODEL_DIR, "model_xgb_comparable.rds")

if (file.exists(xgb_model_file)) {
  # Load existing model
  model_xgb <- readRDS(xgb_model_file)
  # cat("XGBoost model loaded from:", xgb_model_file, "\n")
} else {
  # Train new model
  final_params <- list(objective = "reg:squarederror", eval_metric = "mae",
    eta = best_params$eta, max_depth = best_params$max_depth,
    subsample = best_params$subsample, colsample_bytree = best_params$colsample_bytree, nthread = 1)
  
  set.seed(SEED)
  model_xgb <- xgb.train(params = final_params, data = dtrain, nrounds = best_params$nrounds, verbose = 0)
  
  # Save model for future runs
  saveRDS(model_xgb, xgb_model_file)
  cat("XGBoost model trained and saved to:", xgb_model_file, "\n")
}

# Define final_params for use in learning curves (needed even when loading)
final_params <- list(objective = "reg:squarederror", eval_metric = "mae",
  eta = best_params$eta, max_depth = best_params$max_depth,
  subsample = best_params$subsample, colsample_bytree = best_params$colsample_bytree, nthread = 1)

preds_xgb <- predict(model_xgb, dtest)
metrics_xgb <- get_metrics(preds_xgb, y_test)
```

## Diagnostic Plots

```{r xgb-learning-curves}
train_mae <- test_mae <- numeric()
max_rounds <- min(500, best_params$nrounds * 2)

for (r in seq(10, max_rounds, by = 10)) {
  set.seed(SEED)
  temp_model <- xgb.train(params = final_params, data = dtrain, nrounds = r, verbose = 0)
  train_mae <- c(train_mae, mean(abs(predict(temp_model, dtrain) - y_train)))
  test_mae <- c(test_mae, mean(abs(predict(temp_model, dtest) - y_test)))
}

training_history <- data.frame(iter = seq(10, max_rounds, by = 10), train_mae, test_mae)
learning_curve_df <- training_history %>%
  pivot_longer(cols = c(train_mae, test_mae), names_to = "Dataset", values_to = "MAE") %>%
  mutate(Dataset = ifelse(Dataset == "train_mae", "Train", "Validation"))

best_row <- which.min(abs(training_history$iter - best_params$nrounds))
overfit_gap <- training_history$test_mae[best_row] - training_history$train_mae[best_row]
gap_pct <- round(100 * overfit_gap / training_history$test_mae[best_row], 1)
```

```{r xgb-diagnostic-plots, fig.width=6.5, fig.height=2.3, out.width="100%"}
p1 <- ggplot(learning_curve_df, aes(x = iter, y = MAE, color = Dataset)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = best_params$nrounds, linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  scale_color_manual(values = c("Train" = "#2E86AB", "Validation" = "#E63946")) +
  labs(x = "Rounds", y = "MAE", title = "Learning Curves") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom", legend.key.size = unit(0.3, "cm"),
    plot.title = element_text(size = 9), legend.title = element_blank())

pred_df <- data.frame(pred = preds_xgb, actual = y_test, residual = y_test - preds_xgb)

p2 <- ggplot(pred_df, aes(x = pred, y = actual)) +
  geom_point(alpha = 0.5, size = 0.8, color = "#2E86AB") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 0.7) +
  labs(x = "Predicted", y = "Actual", title = "Pred vs Actual") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

p3 <- ggplot(pred_df, aes(x = residual)) +
  geom_histogram(bins = 25, fill = "#2E86AB", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.7) +
  labs(x = "Residual", y = "Count", title = "Residual Distribution") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

grid.arrange(p1, p2, p3, ncol = 3)
```

The **learning curves** show that the model converges with minimal overfitting. In the first 50 rounds, training and validation MAE drop as the model captures the main signal. After that, improvement slows and additional rounds provide little benefit, which justifies early stopping. The small train-validation gap of 0.021 MAE indicates that overfitting is minimal.

The **predictions vs actual** plot shows that most points lie close to the diagonal, indicating accurate predictions. Larger errors appear at the highest and lowest quality scores, which is expected because extreme values are rare and the model has fewer examples to learn from. There is no visible systematic bias, as points are evenly distributed around the diagonal.

The **residuals** are centered around zero, confirming that the prediction bias is minimal. Their roughly bell-shaped distribution suggests that errors are mostly random rather than systematic, meaning the model has captured the main signal.

<!-- Note: This part was important during the work on the XGBoost part, but we have to comment it out from the final report because of the report length limitations. However, we did not want to delete it completely because of the work put into this part.

## Feature Importance

```{r xgb-feature-importance, fig.width=3.5, fig.height=2.3, out.width="55%"}
importance <- xgb.importance(feature_names = colnames(X_train), model = model_xgb)
imp_df <- head(importance, 10)
imp_df$Feature <- factor(imp_df$Feature, levels = rev(imp_df$Feature))

ggplot(imp_df, aes(x = Gain, y = Feature)) +
  geom_col(fill = "#2E86AB", width = 0.7) +
  labs(x = "Gain", y = NULL, title = "Top 10 Features by Gain") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))
```

```{r xgb-importance-inline, results='asis'}
top4 <- head(importance, 4)
cat("\\begin{center}
\\scriptsize
\\begin{tabular}{lrrr}
\\toprule
Feature & Gain & Cover & Frequency \\\\
\\midrule")
for (i in 1:4) {
  cat("\n", top4$Feature[i], " & ", round(top4$Gain[i], 3), " & ", round(top4$Cover[i], 3), " & ", round(top4$Frequency[i], 3), " \\\\")
}
cat("\n\\bottomrule
\\end{tabular}
\\end{center}")
```

XGBoost provides three **feature importance** measures: gain, cover, and frequency. Gain shows how much a feature reduces prediction error, so higher gain means higher predictive power. For example, Flavor has the highest gain (about 0.3), meaning it explains roughly 30% of the model’s total error reduction. Cover measures how many samples are affected by splits on a feature; high cover means the feature is used on large parts of the data. Frequency counts how often a feature is used in splits, but frequent use does not necessarily mean strong impact.

The ranking shows that Flavor is by far the most important feature, which matches domain knowledge, because flavor is the most important criterion for cupping evaluations. Other top features (Aroma, Aftertaste, Balance) are also sensory attributes. Non-sensory features such as altitude or processing method contribute very little, likely because their effects are already reflected in the sensory scores. 

-->

## Conclusions

**Model performance:** XGBoost achieves an MAE of 0.141, meaning predictions are usually within 0.14 points of the true cupper score on a 10-point scale. This is a small but consistent improvement over linear regression (MAE 0.146). The $R^2$ increases from 0.59 to 0.68, showing that XGBoost explains about 9% more variance. This suggests most relationships are linear, with XGBoost mainly capturing minor non-linear effects and interactions.

**Limitations:** Performance is weaker at extreme quality scores due to limited data. The dataset size (about 1300 samples) may restrict learning of complex patterns, and temporal effects such as harvest year are not modeled (because of the issues with source data). Finally, the target variable cupper points is subjective and includes human rating variability.

## Explainable AI: PFI and PDP Analysis

```{r xgb-xai-setup}
if (!require("DALEX")) install.packages("DALEX", quiet = TRUE)
library(DALEX)

predict_xgb <- function(model, newdata) {
  if (is.data.frame(newdata)) newdata <- as.matrix(newdata)
  predict(model, xgb.DMatrix(data = newdata))
}

explainer_xgb <- DALEX::explain(
  model = model_xgb, data = X_train, y = y_train,
  predict_function = predict_xgb, label = "XGBoost", verbose = FALSE
)
```

### Permutation Feature Importance (PFI)

PFI measures how much model performance degrades when a feature's signal is destroyed by shuffling. Unlike XGBoost's native Gain metric, PFI is model-agnostic and accounts for feature interactions.

```{r xgb-pfi-calculation}
loss_mae <- function(observed, predicted) mean(abs(observed - predicted))
set.seed(SEED)
pfi_xgb <- model_parts(explainer = explainer_xgb, loss_function = loss_mae, 
                        type = "difference", B = 10)

pfi_results <- pfi_xgb %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(Mean_Dropout_Loss = mean(dropout_loss), SD = sd(dropout_loss), .groups = "drop") %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(Importance = Mean_Dropout_Loss - min(pfi_xgb$dropout_loss[pfi_xgb$variable == "_full_model_"]),
         Rank = row_number())
```

```{r xgb-pfi-plot, fig.width=4, fig.height=3, out.width="60%"}
pfi_plot_data <- head(pfi_results, 10)
pfi_plot_data$variable <- factor(pfi_plot_data$variable, levels = rev(pfi_plot_data$variable))

ggplot(pfi_plot_data, aes(x = Importance, y = variable)) +
  geom_col(fill = "#2E86AB", width = 0.7) +
  geom_errorbarh(aes(xmin = Importance - SD, xmax = Importance + SD), height = 0.3, linewidth = 0.3) +
  labs(x = "MAE Increase", y = NULL, title = "Top 10 Features by PFI") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))
```

**PFI interpretation:**

*Sensory attributes dominate:* The top features are all sensory scores, confirming the model relies on expert taste evaluation.

*PFI vs Gain:* PFI provides a model-agnostic view that complements XGBoost's native importance. Both agree that Flavor is most predictive.

### Partial Dependence Plots (PDP)

PDP shows the marginal effect of a feature on predictions, averaging over all other features. We plot the top 4 features by PFI.

```{r xgb-pdp-calculation}
top_features_pfi <- head(pfi_results$variable, 4)

pdp_list <- list()
for (feat in top_features_pfi) {
  pdp_list[[feat]] <- model_profile(explainer = explainer_xgb, variables = feat, 
                                     type = "partial", N = 300)
}
```

```{r xgb-pdp-combined, fig.width=5, fig.height=3, out.width="70%"}
pdp_combined_data <- do.call(rbind, lapply(top_features_pfi, function(feat) {
  pdp_data <- pdp_list[[feat]]$agr_profiles
  data.frame(Feature = feat, x = pdp_data$`_x_`, yhat = pdp_data$`_yhat_`)
}))

ggplot(pdp_combined_data, aes(x = x, y = yhat)) +
  geom_line(color = "#2E86AB", linewidth = 0.8) +
  facet_wrap(~Feature, scales = "free_x", nrow = 1) +
  labs(x = "Feature Value", y = "Predicted Cupper Points") +
  theme_minimal(base_size = 8) +
  theme(strip.text = element_text(size = 7, face = "bold"))
```

**PDP interpretation:**

*Positive monotonic relationships:* All sensory features show the expected pattern—higher scores lead to higher predicted quality.

*Flavor has steepest slope:* The PDP range for Flavor is largest, indicating it has the strongest marginal effect on predictions.

# Model Comparison - SVM vs XGBoost

```{r comparison-metrics}
# Retrieve SVM results saved earlier
svm_mae <- svm_results$mae
svm_rmse <- svm_results$rmse
svm_r2 <- svm_results$r2
svm_predictions <- svm_results$predictions
svm_test_cupper_points <- svm_results$test_cupper_points
best_run <- svm_results$best_run

# Compile comparison metrics
comparison_df <- data.frame(
  Model = c("SVM (Linear Kernel)", "XGBoost"),
  MAE = c(round(svm_mae, 4), round(metrics_xgb$MAE, 4)),
  RMSE = c(round(svm_rmse, 4), round(metrics_xgb$RMSE, 4)),
  R2 = c(round(svm_r2, 4), round(metrics_xgb$R2, 4))
)

kable(comparison_df)
```

```{r comparison-plot, fig.height=2.8, fig.width=7.2, out.width="75%"}
par(mfrow = c(1, 2))

# SVM plot
plot(svm_predictions, svm_test_cupper_points,
     xlab = "Predictions",
     ylab = "True Values",
     main = paste0("SVM (MAE: ", round(svm_mae, 4), ")"),
     pch = 19,
     col = rgb(0.8, 0.2, 0.2, 0.5))
abline(0, 1, col = "darkred", lwd = 2)

# XGBoost plot  
plot(preds_xgb, y_test,
     xlab = "Predictions",
     ylab = "True Values",
     main = paste0("XGBoost (MAE: ", round(metrics_xgb$MAE, 4), ")"),
     pch = 19,
     col = rgb(0.2, 0.4, 0.6, 0.5))
abline(0, 1, col = "navy", lwd = 2)

par(mfrow = c(1, 1))
```

```{r residual-comparison, fig.height=2.4, fig.width=4.8, out.width="55%"}
# Calculate residuals
svm_residuals <- svm_test_cupper_points - svm_predictions
xgb_residuals <- y_test - preds_xgb

# Create residual comparison dataframe
residuals_df <- data.frame(
  Residual = c(svm_residuals, xgb_residuals),
  Model = c(rep("SVM", length(svm_residuals)), rep("XGBoost", length(xgb_residuals)))
)

# Histogram comparison
ggplot(residuals_df, aes(x = Residual, fill = Model)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("SVM" = "#E63946", "XGBoost" = "#2E86AB")) +
  labs(title = "Distribution of Prediction Residuals",
       x = "Residual (True - Predicted)",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

---

*Report generated: `r Sys.Date()`*
