---
title: "Coffee Quality Prediction: SVM & XGBoost Analysis"
author: "Nataliia Remezova, Luisa Kalkert, Nina Immenroth, Yüsra Senem"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    latex_engine: xelatex
  word_document:
  html_document:
    number_sections: true
    df_print: paged
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{multicol}
  - \usepackage{caption}
  - \captionsetup[table]{skip=3pt}
  - \captionsetup[figure]{skip=5pt}
  - \setlength{\tabcolsep}{4pt}
  - \renewcommand{\arraystretch}{0.9}
  - \setlength{\parskip}{3pt}
  - \setlength{\floatsep}{8pt}
  - \setlength{\textfloatsep}{8pt}
geometry: margin=0.85in, top=0.5cm
bibliography: references.bib
---

```{r setup, include=FALSE}
# Optional: Clear environment before running (uncomment if needed)
# rm(list = ls())

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 3.5, fig.height = 2.5, out.width = "50%",
  fig.pos = "H"
)

# Set seed for reproducibility
SEED <- 123 # Global constant
set.seed(SEED)
knitr::opts_chunk$set(cache = TRUE)

# Model directory for saving/loading trained models
MODEL_DIR <- "saved_models"
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

library(ggplot2)
library(corrplot)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(skimr)
library(e1071)
library(tidyverse)
library(gridExtra)
library(stringr)
library(here)
library(shapviz)
library(patchwork)
library(DALEX)
```

# Introduction

In this report we set out to explore a data set about coffee quality. It covers everything from initial descriptive analysis to the mathematical foundations and application of two machine learning methods (SVM for Regression and XGBoost), as well as explainable AI analyses. We showcase how we use these techniques to extract meaningful insights into what makes a coffee expert choose their favorite coffees.

## Descriptive Analysis

**Summary of the Descriptive Analysis:**
The data used in this report originates from the website of the Coffee Quality Institiute and was scraped in 2018 by user jldbc. It can be found on GitHub ^[https://github.com/jldbc/coffee-quality-database].

The dataset has 1,311 coffee evaluations (observations) with 44 different variables. They features comprise a mix of sensory scores (e.g. Aroma, Acidity, Moisture) and agricultural cultivation data (e.g. Region, Altitude, Processing Method). Out of these, 24 variables are categorical and 20 are numerical. For our project, we decided to use **Cupper Points** as our target variable. “Cupper Points” is a score reflecting how much a professional coffee taster enjoyed an individual coffee. We avoided "Total Cup Points" as target variable as it is a linear combination of other sensory scores and hence not suitable for analysis.
\begin{minipage}[c]{0.48\textwidth}
```{r, fig.width=6, fig.height=4, out.width="100%"}
# Read the data
coffee_data <- read.csv(here("arabica_data_cleaned.csv"), na.strings = c("", "NA", " ", "n/a"))
# 1. Select the features you want to focus on
selected_features <- coffee_data %>% 
  select(Cupper.Points, altitude_mean_meters, Aroma, Flavor, unit_of_measurement, Processing.Method, 
         Bag.Weight, Variety, Lot.Number, Quakers, Country.of.Origin)

# 2. Calculate missing values and prepare for plotting
missing_data <- selected_features %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Missing_Count") %>%
  mutate(Percentage = (Missing_Count / nrow(coffee_data)) * 100) %>%
  arrange(desc(Missing_Count)) # Sort so the biggest problem is first

# 3. Create the bar plot
ggplot(missing_data, aes(x = reorder(Feature, -Missing_Count), y = Missing_Count)) +
  geom_col(fill = "firebrick", color = "white") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            vjust = -0.5, size = 3.5) + # Add percentage labels above bars
  theme_minimal() +
  ylim(0, max(missing_data$Missing_Count) + 50) + # Add space for labels
  labs(title = "Missing Values Count in Selected Features",
       x = "Feature",
       y = "Number of Missing Observations") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
\end{minipage} \begin{minipage}[c]{0.45\textwidth} \textbf{Dealing with Missing Data} The amount of data missing from the data set varies greatly by feature. Especially for categorical data, we can see high ratios of up to almost 80\% gaps. Variables with this high rates of missing entries will most likely not add useful signal to the model training. Instead we focus on features where either data is abundant or where only a medium amount ($<$20\%) of data is missing and there are sensible ways to impute missing values.\end{minipage} \begin{minipage}[c]{0.48\textwidth} \hspace{0.05\textwidth}
```{r, fig.width=6, fig.height=4, out.width="100%"}
# Prepare the data in 'long' format
long_data <- coffee_data %>%
  select(Aroma, Flavor, Cupper.Points) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Score")

# Create the Horizontal Combined Boxplot (Left side)
p_box <- ggplot(long_data, aes(y = Feature, x = Score, fill = Feature)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.shape = 16) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Outlier & Range Comparison",
       y = "Feature", x = "Score") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))

# Create Faceted Histograms (Right side)
p_hist <- ggplot(long_data, aes(x = Score, fill = Feature)) +
  geom_histogram(bins = 20, color = "white", alpha = 0.8) +
  facet_wrap(~Feature, ncol = 1) + 
  theme_light() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Distribution Shapes",
       x = "Score", y = "Count") +
  theme(legend.position = "none",
        strip.text = element_text(size = 8, face = "bold", color = "black"),
        plot.title = element_text(face = "bold"))

# Combine Side-by-Side
# We use '|' to put them together. 
(p_box | p_hist) + 
  plot_layout(widths = c(1.2, 1)) + 
  plot_annotation(title = "Sensory Feature Profile: Range and Distribution",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
```
\end{minipage} \begin{minipage}[c]{0.45\textwidth} \textbf{The Distribution of the Variables} Looking at the histograms for the sensory variables, we can see them following mutually similar distributions. A majority of the scores fall closely together in the interval between 7 and 8. A few outliers are noticeable in the extremes of the possible values. These could indicate a coffee taster having a strong opinion about the coffee, a special variety of coffee, or might be an error in the data.\end{minipage}
\begin{minipage}[c]{0.48\textwidth}
```{r , fig.width=7, fig.height=5, echo=FALSE, out.width="100%"}
# Select only the numerical sensory columns (adjust indices as needed)
# Usually columns like Aroma:Cupper.Points
sensory_cols <- coffee_data %>% select(Aroma, Flavor, Aftertaste, Acidity, Body, Balance, Sweetness, Uniformity, Clean.Cup, Moisture)

# Calculate correlation matrix
M <- cor(sensory_cols, use = "complete.obs")

# Visualize
corrplot(M, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", 
         mar = c(0, 0, 1.5, 0),
         title = "Correlation Between Sensory Attributes"
         )
```
\end{minipage} \begin{minipage}[c]{0.45\textwidth} \textbf{Correlation Between Numerical Features} Most of the sensory variables are highly correlated. For instance a pleasant smell, which translates to a high value for the Aroma score, is almost always accompanied by great taste as indicated by a high flavor score. Moisture stands out as the only sensory attribute that shows a slight negative or no correlation with all other variables in this group.\end{minipage}

# A Brief Mathematical Overview of SVM and XGBoost
## Support Vector Machine for Regression
```{r, fig.width=4.5, fig.height=2.5, out.width='65%', fig.align='center'}
par(mar = c(3, 3, 0.3, 1.5), mgp = c(1.8, 0.6, 0), cex = 0.8)

set.seed(SEED)
x <- seq(-10, 10, 0.5) + rnorm(length(seq(-10, 10, 0.5)), mean = 0, sd = 0.1)
y <- 1/170 * (x^3 - 3* x^2 - 70*x + 5 + rnorm(length(x), mean = 0, sd = 170))

library(e1071)
epsilon <- 1
cost <- 1

model <- svm(y ~ x, data = data.frame(x = x, y = y), epsilon = epsilon, cost = cost)
predictedY <- predict(model, data.frame(x = x))

plot(x, y, pch = 16, col = "grey70")
lines(x, predictedY, col = "darkgreen")

polygon(
 c(x, rev(x)),
 c(predictedY + epsilon, rev(predictedY - epsilon)),
 border = NA,
 col = rgb(1, 0, 0, 0.15)
)

idx_furthest <- order(abs(predictedY - y), decreasing = TRUE)[1:17]
points(x[idx_furthest], y[idx_furthest], 
       col = "#000000", bg = "#ffffff00", pch = 21, cex = 1.3, lwd = 1)

# ε arrow - original logic
arrows(10, predictedY[length(predictedY)], 10, predictedY[length(predictedY)] + epsilon, 
       length = 0.06, lty = 2, col = "black", code = 3)
text(10, predictedY[length(predictedY)] + 0.5, expression(epsilon), pos = 4, cex = 0.85)

# ξ_j* arrow - original logic  
arrows(5.05, -4, 5.05, -2.28, length = 0.06, lty = 2, col = "black")
text(5.05, -3.2, expression(xi[j]^"*"), pos = 4, cex = 0.85)

# ξ_k arrow - original logic
arrows(x[length(x) - 12], y[length(y) - 12], x[length(x) - 12], y[length(y) - 12] - 0.5, 
       length = 0.06, lty = 2, col = "black")
text(x[length(x) - 12], y[length(y) - 12] - 0.3, expression(xi[k]), pos = 4, cex = 0.85)
```
Support Vector Regression expands Support Vector Machines from discrete classification to a continuos regression model [@SVMregression].

For the regression task, instead of fitting a hyperplane, we are trying to fit a (multidimensional) function $f(x)$ to the data. Similar as in the classification task, we draw a margin around the function. However, now the margin $\varepsilon$ is fixed at the start. (Note that $\varepsilon$ here, is not the slack variable as the notation is commonly used for SVM in classification, e.g. in "An Introduction to Statistical Learning" [@HastiesSVM], but rather defines the width of the margin. Also, we do not aim to optimize the margin boundaries here, but instead fix them before computation starts). We want to find a function f(x) that is as flat as possible, while fitting all points into the required margin. Again, we allow for some slack using slack variables. A wider margin decreases the risk of overfitting, while a smaller margin captures more intricacies of the data. 

Mathematically speaking, want to find a function $f(x)$ with $| f(x_i) - y_i |  \leq \varepsilon$ for all $i$. To avoid overfitting, we also introduce the constraint of making the function as flat as possible.
"Flatness" in this context is a measure of how sensitive the function is to change. For a linear function $f(x) = x^\top \beta + b$ increasing this flatness can be expressed through minimizing the norm of the function gradient: $\lVert \beta \rVert$. Minimizing the expression $\frac{1}{2} \lVert \beta \rVert^2$ leads to the same optimum, while allowing for more elegant mathematical solutions, and is therefore used for computation.


Again, we want to allow for some slack, so as in the SVM for classification, we are introducing slack variables $\xi_i$ and $\xi_i^\ast$. These slack variables allow for some room for margin violations. 
The optimization problem we want to solve is to minimize: 

$$
\frac{1}{2} \lVert \beta \rVert^2 + K \sum_{i=1}^n (\xi_i + \xi_i^\ast)
$$
subject to:
$$
y_i - f(x_i) \leq \varepsilon + \xi_i, 
$$
$$
f(x_i) - y_i \leq \varepsilon + \xi_i^\ast
$$
and 
$$
\xi_i > 0, \xi_i^\ast > 0 \quad \forall i
$$


where $K$ is fixed and $\beta$, $\xi$ and $\xi^\ast$ are variable.
The factor $K$ is introduced as cost for regularization purposes. It is choosen a priori to determine how strong the tradeoff between flatness and overfitting should be. A larger $K$ leads to a stronger impact of the term, therefore penalizing stronger margin violations more heavily and resulting in a less flat function curve (higher bias, lower variance). Conversly, a smaller $K$ leads to a flatter curve (lower bias, higher variance).


As in the SVM for classification, this can be generalized to non-linear functions by transforming the features using kernels.

## XGBoost
XGBoost is an algorithm based on gradient boosted trees and second-order approximation [@chen2016xgboost]. It is widely used in machine learning due to its flexibility and strong empirical performance. These properties arise from a combination of algorithmic and systems-level optimizations, including efficient handling of sparse input data, approximate tree construction via weighted quantile sketches, and parallelized tree learning. In addition, careful optimization of memory access, data compression, and out-of-core computation enables the method to scale to very large datasets while maintaining computational efficiency.

This section describes the mathematical foundations of the algorithm, namely ensemble methods, regularization, and gradient boosting machines combined with second order approximation. 

**Tree Ensembles and Regularization**

XGBoost models the prediction as an additive ensemble of $K$ functions:
$$
\hat{y}_i = \sum_{k=1}^{K} f_k(x_i), \quad f_k \in \mathcal{F},
$$
where each $f_k$ is a regression tree belonging to the function space $\mathcal{F}$ [@chen2016xgboost; @james2021introduction]. Each tree acts as a weak learner that partitions the feature space into disjoint regions and assigns a constant prediction to each leaf [@james2021introduction].

The model is trained by minimizing a regularized objective function of the form
$$
\mathcal{L} = \sum_{i} \ell(y_i, \hat{y}_i) + \sum_{k=1}^{K} \Omega(f_k),
$$
where $\ell(\cdot)$ is a differentiable loss function and $\Omega(f)$ is a regularization term that penalizes model complexity [@chen2016xgboost]. This term is defined as
$$
\Omega(f) = \gamma T + \frac{1}{2} \lambda \sum_{j=1}^{T} w_j^2,
$$
with $T$ denoting the number of leaves in the tree and $w_j$ the weight of leaf $j$. The regularization term encourages simpler trees and smooth leaf weights, helping to prevent overfitting.

The ensemble structure allows the model to reduce bias by combining multiple weak learners, while regularization controls variance by limiting tree complexity and the magnitude of leaf weights [@james2021introduction]. This balance between bias and variance enables XGBoost to achieve strong predictive performance while maintaining good generalization.

**Gradient Boosting Machines**

The tree ensemble is constructed in a stage-wise manner, where weak learners are added iteratively to improve the model [@friedman2001greedy]. At iteration $t$, the prediction is given by
$$
\hat{y}_i^{(t)} = \hat{y}_i^{(t-1)} + f_t(x_i),
$$
with $f_t \in \mathcal{F}$ denoting the newly added regression tree.

Gradient tree boosting can be interpreted as gradient descent in function space, where each new learner is chosen to minimize the loss function in the direction of the negative gradient [@friedman2001greedy]. Specifically, the new tree at iteration $t$ is trained to fit the negative first order gradient
$$
-g_i^{(t)} = -\left. \frac{\partial \ell(y_i, \hat{y})}{\partial \hat{y}} \right|_{\hat{y} = \hat{y}_i^{(t-1)}}.
$$

XGBoost extends this framework by explicitly approximating the objective function using a second-order Taylor expansion of the loss around the current predictions:
$$
\ell(y_i, \hat{y}_i^{(t-1)} + f_t(x_i)) \approx \ell(y_i, \hat{y}_i^{(t-1)}) + g_i f_t(x_i) + \frac{1}{2} h_i f_t(x_i)^2,
$$
where
$$
h_i = \left. \frac{\partial^2 \ell(y_i, \hat{y})}{\partial \hat{y}^2} \right|_{\hat{y} = \hat{y}_i^{(t-1)}}
$$
is the second order derivative of the loss [@chen2016xgboost]. By incorporating both first- and second-order information, XGBoost enables efficient optimization of tree structures and leaf weights within each boosting iteration.

# Support Vector Regression

## Data Preprocessing and Cleaning 

### Feature selection
In SVM Regression, categorical variables are One-Hot Encoded using dummy variables.
This means, features with many unique values will lead to a sparse feature matrix. This is disadvantageous as it increases computational cost and poses the risk of overfitting. 
Additionally, categories with only a few or even only one observation most likely will not add to model robustness, but are adding noise to the signal.
Therefore, we are removing categorical features with many unique values, only keeping columns with a low number of unique values.
E.g. we are not including the "Region" column with 344 unique values in the model training, because we'd on average have less than 4 observations per Region. 
For features with a high number of missing values, we either remove the column or impute the missing values.
For instance for the "Variety" column, we have about 15% missing values and 8% entries "other" at 30 different varieties. 
Keeping those, would mean adding 30 sparse columns, and at the same time the information gain is most likely not too high with this many missing values.
Therefore, we are not including this feature. 


### Imputation
For the altitude column, we have about 17% missing values. Here, we are imputing the missing values from the "Region" column.
For each observation with missing altitude, we are imputing the median altitude of all coffees from the same region.
This covers almost all missing values. For the remaining observations, we are imputing the mean altitude of all coffees from the same country of origin.
As the "Region" and "Country" columns are not used in the model training, we are not creating correlated features with this imputation.
We are imputing the missing values after the train/test split, using only values from the training set, to avoid data leakage from the test set into the training set.


### Further preprocessing
The bag weight column contains both pounds and kilograms, so we are converting all to kilograms. 
However, for some entries, it is unclear which unit is used. For these, we are averaging the weight in pounds and kilograms.
This means for those entries the weight is certainly incorrect, but should be within an order of magnitude of the correct weight. 
If we can assume that speciality coffee is sold in smaller quantities, we can still gain valuable information from the weight.


### Scaling
The SVM method from the e1071 packaged used for model training has a built in scaling functionality. 
Both predictor and target variables are scaled to have expected value 0 and variance 1. [@SVMpackage]
This transformation is based on the training data and applied to train and test data sets.


```{r svm-preprocessing, echo=FALSE, message=FALSE, warning=FALSE}

# Avoid categorical variables with many unique values because they will create many dummy variables. 
columns_to_use <- c(
  "Number.of.Bags",
  "Bag.Weight",
  # "Country.of.Origin", Increased runtime too much
  "Processing.Method", 
  "Aroma",
  "Flavor",
  "Aftertaste",
  "Acidity",
  "Body",
  "Balance",
  "Uniformity",
  "Clean.Cup",
  "Sweetness",
  "Cupper.Points", # Target variable
  "Moisture",
  "Category.One.Defects",
  "Quakers",
  "Category.Two.Defects",
  "altitude_mean_meters" # Impute missing values from region and country origins
)




## Bag Weight

# Get bags with weight in pounds
bags_with_weight_in_pounds_indices <- grepl("lbs", coffee_data$Bag.Weight)

# Get bags with weight in kg
bags_with_weight_in_kg_indices <- grepl("kg", coffee_data$Bag.Weight)

# Get bags with neither pounds nor kg
bags_with_neither_pounds_nor_kg_indices <- !bags_with_weight_in_pounds_indices & !bags_with_weight_in_kg_indices

# Get bags with both pounds and kg (incorrectly formatted)
bags_with_both_pounds_and_kg_indices <- bags_with_weight_in_pounds_indices & bags_with_weight_in_kg_indices

# Remove bags with both pounds and kg from bags with weight indices and bags with weight in kg indices
bags_with_weight_in_pounds_indices <-
  bags_with_weight_in_pounds_indices & !bags_with_both_pounds_and_kg_indices

bags_with_weight_in_kg_indices <-
  bags_with_weight_in_kg_indices & !bags_with_both_pounds_and_kg_indices


# Sanity check
# print(paste("Number of bags with weight in pounds:", sum(bags_with_weight_in_pounds_indices)))
# print(paste("Number of bags with weight in kg:", sum(bags_with_weight_in_kg_indices)))
# print(paste("Number of bags with neither pounds nor kg:", sum(bags_with_neither_pounds_nor_kg_indices)))
# print(paste("Number of bags with both pounds and kg:", sum(bags_with_both_pounds_and_kg_indices)))
# print(paste("Number of bags with weight:", length(coffee_data$Bag.Weight)))

# Extract numeric value from Bag.Weight column
coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
# Multiply bags with weight in pounds by 0.453592 to convert to kg
coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] <- coffee_data$Bag.Weight[bags_with_weight_in_pounds_indices] * 0.453592

# For bags with neither pounds nor kg, average the weight in pounds and kg
coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] <- coffee_data$Bag.Weight[bags_with_neither_pounds_nor_kg_indices] * (1 + 0.453592)/2
coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] <- coffee_data$Bag.Weight[bags_with_both_pounds_and_kg_indices] * (1 + 0.453592)/2



# For the Processing Method fill NAs with "Unknown"
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"



# Split data into training and test set
# Separate features and target


# Simple random 80/20 split
set.seed(SEED)
n_total <- nrow(coffee_data)
n_train <- floor(0.8 * n_total)
train_index <- sample(1:n_total, n_train)

train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]



## Altitude

library(tidyverse)


# Create dict of regions and their median altitudes
region_altitudes <- train_data %>%
  group_by(Region) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

region_altitudes <- region_altitudes[complete.cases(region_altitudes),]

# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters)) 
train_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(train_data$Region[missing_altitude_indices], region_altitudes$Region)]


# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- region_altitudes$altitude_median_mean_meters[match(test_data$Region[missing_altitude_indices], region_altitudes$Region)]




### Repeat imputation for country of origin

# Create dict of regions and their median altitudes
country_altitudes <- train_data %>%
  group_by(Country.of.Origin) %>%
  summarise(altitude_median_mean_meters = median(altitude_mean_meters, na.rm = TRUE)) 

country_altitudes <- country_altitudes[complete.cases(country_altitudes),]


# Impute for train data
missing_altitude_indices <- which(is.na(train_data$altitude_mean_meters))
train_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(train_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]

# Impute for test data
missing_altitude_indices <- which(is.na(test_data$altitude_mean_meters))
test_data[missing_altitude_indices, "altitude_mean_meters"] <- country_altitudes$altitude_median_mean_meters[match(test_data$Country.of.Origin[missing_altitude_indices], country_altitudes$Country.of.Origin)]




# Remove columns we are not using
train_data <- train_data[,columns_to_use]
test_data <- test_data[,columns_to_use]


# Remove remaining NAs
train_data <- train_data[complete.cases(train_data),]
test_data <- test_data[complete.cases(test_data),]


# Add One Hot encoding
train_data <- data.frame(model.matrix(~ ., data = train_data))[,-1]
test_data  <- data.frame(model.matrix(~ ., data = test_data))[,-1]

# Check if categories align between train and test 
# print("All Categories match between train and Test data")
# print(all(colnames(train_data) == colnames(test_data)))
  

```

```{r svm-data-info, echo=FALSE}
# print(paste0(
#   "Number of training samples: ", nrow(train_data), "\n",
#   "Number of test samples: ", nrow(test_data), "\n",
#   "Number of features: ", ncol(train_data), "\n",
#   "Average cupper points in training set: ", mean(train_data$Cupper.Points), "\n",
#   "Average cupper points in test set: ", mean(test_data$Cupper.Points), "\n",
#   "Features used: ", paste(columns_to_use, collapse = ", ")

# ))

```

## Hyperparameter Optimization

For the prediction of cupper points we train an SVM for regression on the data set, holding out the test data set for final evaluation. 
To choose appropriate hyperparameters, we conduct a hyperparameter optimization on the training data set using 10-fold cross-validation. 
We use a grid search for different values for cost and epsilon for a radial, linear, and 2-degree polynomial kernel. We compare the mean absolute error between runs to choose the best-performing hyperparameter combination. Finally, we train a model with these hyperparameters on the entire training set and evaluate on the test data set.


```{r svm-hyperparameter-tuning, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2.5, fig.width=8, out.width="95%"}




####  Fit SVM model with polynomial kernel (degree 2) 
library(e1071)

# Gridsearch 
epsilon = seq(0,2,0.1)
cost = 2^(-10:3)
# epsilon = seq(0,1,0.5)
# cost = 2^(-2:-1)


# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "polynomial",
                   degree = 2,
                   ranges = list(epsilon = epsilon, cost = cost)
)


# Show Results and visualize tuning as a heatmap 
# print(tuneResult)
library(ggplot2)

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 
 library(dplyr)
 library(tidyr)
 
 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Set up layout for side-by-side plots
par(mfrow = c(1, 3))

# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using 2nd\nDegree Polynomial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


# Best parameters 
best_error <- max(data.frame(data_wide[,-1]))

# Add polynomial kernel results to best results
best_runs <- data.frame(
  kernel = "polynomial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
)




# Fit SVM model with linear kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   kernel = "linear",
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)


 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Linear Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)


best_error <- max(data.frame(data_wide[,-1]))

# add linear kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "linear",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))



# Fit SVM model with radial kernel 

# Tune automatically with cross-validation
tuneResult <- tune(svm, Cupper.Points ~ .,  data = data.frame(train_data),
                   ranges = list(epsilon = epsilon, cost = cost)
)

# Show Results and visualize tuning as a heatmap 

perf <- tuneResult$performances
best <- data.frame(
  epsilon = tuneResult$best.parameters$epsilon,
  cost    = tuneResult$best.parameters$cost
)

 # Reshape performance data so that each row is an epsilon value and
 # each column (after the first) is a cost value, holding the error.
 data_wide <- perf %>%
   select(epsilon, cost, error) %>%
   arrange(epsilon, cost) %>%
   pivot_wider(
     names_from = cost,
     values_from = error
   )
 
 # Extract x (epsilon), y (cost) and z (error matrix) for image()
 x_vals <- data_wide$epsilon
 y_vals <- as.numeric(names(data_wide)[-1])
 z_mat  <- as.matrix(data_wide[, -1])
 
# Use image() with a transformed y axis: log2(cost)
image(x = x_vals, y = log2(y_vals), z = log2(z_mat),
      xlab = "epsilon",
      ylab = "log2(cost)",
      main = "Errors using Radial Kernel",
      col = colorRampPalette(c("white", "navy"))(50))

points(best$epsilon, log2(best$cost), col = "orange", pch = 18, cex = 1.5)

# Reset plotting parameters
par(mfrow = c(1, 1))

best_error <- max(data.frame(data_wide[,-1]))

# add radial kernel to best runs
best_runs <- rbind(best_runs, data.frame(
  kernel = "radial",
  epsilon = best$epsilon,
  cost = best$cost,
  error = best_error
))

```
These figures show the mean absolute error for the different hyperparameters by kernel. The best performing hyperparameter combination for each kernel are marked with orange points.

```{r svm-final-model}
# Check if saved model exists
svm_model_file <- file.path(MODEL_DIR, "model_svm.rds")

if (file.exists(svm_model_file)) {
  # Load existing model
  svm_model <- readRDS(svm_model_file)
  # cat("SVM model loaded from:", svm_model_file, "\n")
  
  # Still need to determine best_run for display
  best_runs <- best_runs[order(best_runs$err),]
  best_run <- best_runs[1,]
} else {
  # Compare best results for each kernel by getting the row with the lowest error
  best_runs <- best_runs[order(best_runs$err),]
  best_run <- best_runs[1,]
  
  # Train final model with best parameters
  svm_model <- svm(Cupper.Points ~ ., data = train_data,
                   cost = best_run$cost,
                   epsilon = best_run$epsilon, 
                   kernel = as.character(best_run$kernel))
  
  # Save model for future runs
  saveRDS(svm_model, svm_model_file)
  # cat("SVM model trained and saved to:", svm_model_file, "\n")
}

# cat("Best SVM kernel and parameters selected:\n")
# print(best_run[,c("kernel", "epsilon", "cost")])

# Predict on test set
svm_predictions <- predict(svm_model, test_data)

# Evaluate model performance
svm_mae <- mean(abs(svm_predictions - test_data[, "Cupper.Points"]), na.rm = TRUE)
svm_mse <- mean((svm_predictions - test_data[, "Cupper.Points"])^2, na.rm = TRUE)
svm_rmse <- sqrt(svm_mse)
svm_ss_res <- sum((test_data[, "Cupper.Points"] - svm_predictions)^2)
svm_ss_tot <- sum((test_data[, "Cupper.Points"] - mean(test_data[, "Cupper.Points"]))^2)
svm_r2 <- 1 - (svm_ss_res / svm_ss_tot)

# print(paste("MAE:", svm_mae))
```

```{r svm-hpo-table, results='asis'}
cat("\\begin{center}
\\scriptsize
\\begin{tabular}{cccccc}
\\toprule
epsilon & cost & kernel & CV MAE \\\\
\\midrule
", best_run$epsilon, " & ", best_run$cost, " & ", best_run$kernel, " & ", round(best_run$error, 3), " \\\\
\\bottomrule
\\end{tabular}
\\end{center}")
```


The best performing kernel is the linear kernel, which means the relationship between the features and the target variable is best approximated by a multidimensional hyperplane.

```{r svm-predictions-plot, echo=FALSE, fig.height=2.8, fig.width=4, out.width="50%"}
# Plot the predictions vs the true values
# plot(svm_predictions, test_data[, "Cupper.Points"],
#      xlab = "Predictions",
#      ylab = "True Values",
#      main = "SVM: Predictions vs True Values")

# get predictions for a random data point from the test data set
random_index <- sample(1:nrow(test_data), 1)
random_data <- test_data[random_index,]
random_prediction <- predict(svm_model, random_data)
# print("Random Data Point")
# print(random_data)
# print(paste("Prediction:", random_prediction))
# print(paste("True value:", random_data[, "Cupper.Points"]))


# Here, we can see the predictions vs. the true values for the test data set. We can see that aside from two outliers, the predictions are close to the true values.

```



```{r save-svm-results-for-comparison}
# Save SVM results for use in Part 3 comparison
# These variables will be used after XGBoost training
svm_results <- list(
  mae = svm_mae,
  rmse = svm_rmse,
  r2 = svm_r2,
  predictions = svm_predictions,
  test_cupper_points = test_data[, "Cupper.Points"],
  best_run = best_run
)
```

# XGBoost Model

```{r xgb-load-libraries}
library(xgboost)   # Core gradient boosting algorithm
library(dplyr)     # Data manipulation (group_by, summarise, mutate, %>%)
library(ggplot2)   # Visualization
library(tidyr)     # Data reshaping (pivot_longer)
library(knitr)     # Table formatting
library(gridExtra) # Multi-panel plot layouts
library(shapviz)

# Global Constants
CV_FOLDS   <- 10
EARLY_STOP <- 50
MAX_ROUNDS <- 2000
MODEL_DIR  <- "saved_models"

if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)
```

## Data Preprocessing

To ensure that the SVM and XGBoost models are comparable, we apply the same data preprocessing described above to XGBoost, except we skip feature scaling. 

SVM requires scaling because it calculates distances between data points to find the optimal separating hyperplane. Features with larger ranges would dominate the distance calculations, which would make the model incorrect. XGBoost uses decision trees that select split points by maximizing information gain. Because these splits depend only on whether values are higher or lower than a threshold (e.g., “is Flavor > 7.5?”), feature scaling is unnecessary.

```{r xgb-load-data}
coffee_data <- read.csv(here("arabica_data_cleaned.csv"), na.strings = c("", "NA", " ", "n/a"))
```

```{r xgb-define-features}
columns_to_use <- c(
  "Number.of.Bags", "Bag.Weight", "Processing.Method", 
  "Aroma", "Flavor", "Aftertaste", "Acidity", "Body", "Balance",
  "Uniformity", "Clean.Cup", "Sweetness", "Cupper.Points",
  "Moisture", "Category.One.Defects", "Quakers", "Category.Two.Defects",
  "altitude_mean_meters"
)
```

```{r xgb-bag-weight-processing}
bags_lbs <- grepl("lbs", coffee_data$Bag.Weight)
bags_kg <- grepl("kg", coffee_data$Bag.Weight)
bags_neither <- !bags_lbs & !bags_kg
bags_both <- bags_lbs & bags_kg
bags_lbs <- bags_lbs & !bags_both
bags_kg <- bags_kg & !bags_both

coffee_data$Bag.Weight <- as.numeric(sub("^([^ ]+) .*", "\\1", coffee_data$Bag.Weight))
coffee_data$Bag.Weight[bags_lbs] <- coffee_data$Bag.Weight[bags_lbs] * 0.453592
coffee_data$Bag.Weight[bags_neither] <- coffee_data$Bag.Weight[bags_neither] * (1 + 0.453592) / 2
coffee_data$Bag.Weight[bags_both] <- coffee_data$Bag.Weight[bags_both] * (1 + 0.453592) / 2
coffee_data$Processing.Method[is.na(coffee_data$Processing.Method)] <- "Unknown"
```

```{r xgb-train-test-split}
set.seed(SEED)
n_total <- nrow(coffee_data)
train_index <- sample(1:n_total, floor(0.8 * n_total))
train_data <- coffee_data[train_index, ]
test_data <- coffee_data[-train_index, ]
```

```{r xgb-altitude-imputation}
# Hierarchical imputation: Region median → Country median (training data only)
region_alt <- train_data %>% group_by(Region) %>%
  summarise(alt_med = median(altitude_mean_meters, na.rm = TRUE)) %>% filter(!is.na(alt_med))

for (df_name in c("train_data", "test_data")) {
  df <- get(df_name)
  missing <- which(is.na(df$altitude_mean_meters))
  df$altitude_mean_meters[missing] <- region_alt$alt_med[match(df$Region[missing], region_alt$Region)]
  assign(df_name, df)
}

country_alt <- train_data %>% group_by(Country.of.Origin) %>%
  summarise(alt_med = median(altitude_mean_meters, na.rm = TRUE)) %>% filter(!is.na(alt_med))

for (df_name in c("train_data", "test_data")) {
  df <- get(df_name)
  missing <- which(is.na(df$altitude_mean_meters))
  df$altitude_mean_meters[missing] <- country_alt$alt_med[match(df$Country.of.Origin[missing], country_alt$Country.of.Origin)]
  assign(df_name, df)
}
```

```{r xgb-final-prep}
train_data <- train_data[, columns_to_use]
test_data <- test_data[, columns_to_use]
train_data <- train_data[complete.cases(train_data), ]
test_data <- test_data[complete.cases(test_data), ]
train_data <- data.frame(model.matrix(~ ., data = train_data))[, -1]
test_data <- data.frame(model.matrix(~ ., data = test_data))[, -1]
```

## Configuration & Data

```{r xgb-metrics-function}
get_metrics <- function(pred, obs) {
  res <- obs - pred
  list(MAE = mean(abs(res)), RMSE = sqrt(mean(res^2)), R2 = 1 - sum(res^2)/sum((obs - mean(obs))^2))
}
```

```{r xgb-baselines}
y_train <- train_data$Cupper.Points
y_test <- test_data$Cupper.Points
X_train <- as.matrix(train_data[, colnames(train_data) != "Cupper.Points"])
X_test <- as.matrix(test_data[, colnames(test_data) != "Cupper.Points"])

preds_mean <- rep(mean(y_train), length(y_test))
metrics_mean <- get_metrics(preds_mean, y_test)

model_lm <- lm(Cupper.Points ~ ., data = train_data)
preds_lm <- predict(model_lm, newdata = test_data)
metrics_lm <- get_metrics(preds_lm, y_test)
```

```{r xgb-config-data-table, results='asis'}
cat("\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{ll}
\\toprule
\\textbf{Config} & \\textbf{Value} \\\\
\\midrule
Seed & ", SEED, " \\\\
Split & 80/20 \\\\
CV Folds & ", CV_FOLDS, " \\\\
Early Stop & ", EARLY_STOP, " \\\\
Max Rounds & ", MAX_ROUNDS, " \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}
\\hfill
\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{lr}
\\toprule
\\textbf{Data} & \\textbf{Value} \\\\
\\midrule
Train n & ", nrow(train_data), " \\\\
Test n & ", nrow(test_data), " \\\\
Features & ", ncol(train_data) - 1, " \\\\
Target $\\bar{x}$ & ", round(mean(train_data$Cupper.Points), 2), " \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}
\\hfill
\\begin{minipage}[t]{0.32\\textwidth}
\\centering
\\scriptsize
\\begin{tabular}{lrrr}
\\toprule
\\textbf{Model} & \\textbf{MAE} & \\textbf{R²} \\\\
\\midrule
Mean & ", round(metrics_mean$MAE, 3), " & ", round(metrics_mean$R2, 2), " \\\\
Linear & ", round(metrics_lm$MAE, 3), " & ", round(metrics_lm$R2, 2), " \\\\
\\bottomrule
\\end{tabular}
\\end{minipage}

\\vspace{6pt}")
```


**Baseline models**

We use two baseline models: the Mean Predictor and Linear Regression.

The **Mean Predictor** has the $R^2$ = 0, which confirms that our target has meaningful variance to explain. The MAE of 0.309 means that without any features, we'd be off by about 0.3 points on average.

**Linear Regression** provides an interpretable baseline with $R^2$ of 0.59, showing that 59% of the variance in cupper points can be explained by linear combinations of features.
Any improvement from the XGBoost models means capturing non-linear relationships that linear regression misses.

## Hyperparameter Optimization

```{r xgb-hpo}
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest <- xgb.DMatrix(data = X_test, label = y_test)

tune_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1), max_depth = c(3, 5, 7),
  subsample = c(0.7, 0.8, 0.9), colsample_bytree = c(0.7, 0.8, 0.9)
)

hpo_results <- data.frame()
for (i in 1:nrow(tune_grid)) {
  params <- list(objective = "reg:squarederror", eval_metric = "mae",
    eta = tune_grid$eta[i], max_depth = tune_grid$max_depth[i],
    subsample = tune_grid$subsample[i], colsample_bytree = tune_grid$colsample_bytree[i], nthread = 1)
  
  set.seed(SEED)
  cv_result <- xgb.cv(params = params, data = dtrain, nrounds = MAX_ROUNDS,
    nfold = CV_FOLDS, early_stopping_rounds = EARLY_STOP, verbose = 0)
  
  eval_log <- cv_result$evaluation_log
  mae_col <- grep("test.*mae.*mean", names(eval_log), value = TRUE)[1]
  best_iter <- cv_result$best_iteration
  if (is.null(best_iter) || length(best_iter) == 0) best_iter <- which.min(eval_log[[mae_col]])
  
  hpo_results <- rbind(hpo_results, data.frame(
    eta = params$eta, max_depth = params$max_depth, subsample = params$subsample,
    colsample_bytree = params$colsample_bytree, nrounds = best_iter, cv_mae = eval_log[[mae_col]][best_iter]))
}
best_params <- hpo_results[which.min(hpo_results$cv_mae), ]
```

```{r xgb-hpo-table, results='asis'}
cat("\\begin{center}
\\scriptsize
\\begin{tabular}{cccccc}
\\toprule
eta & max\\_depth & subsample & colsample & rounds & CV MAE \\\\
\\midrule
", best_params$eta, " & ", best_params$max_depth, " & ", best_params$subsample, " & ", 
best_params$colsample_bytree, " & ", best_params$nrounds, " & ", round(best_params$cv_mae, 3), " \\\\
\\bottomrule
\\end{tabular}
\\end{center}")
```

**Hyperparameter selection**

We tune four main parameters to balance bias and variance:

**eta (learning rate):** 0.01–0.1. Smaller values make each tree contribute less, reducing overfitting but needing more rounds. 
    
**max_depth:** 3–7. Controls tree complexity. Shallow trees capture simple patterns; deeper trees can model complex patterns but may overfit, especially on our dataset with about 1000 samples.
    
**subsample:** 0.7–0.9. Fraction of rows used per tree. Reduces overfitting by adding randomness. We avoid below 0.7, as it would leave too few samples per tree.
    
**colsample_bytree:** 0.7–0.9. Fraction of features per tree. Prevents any one feature from dominating.
    
These ranges are conservative. With about 1000 samples, extreme values (e.g., eta=0.3, max_depth=10) would overfit, but mild regularization is enough. 

**Lambda and Min_Child_Weight Regularization**  
We tested additional hyperparameters, but they were ineffective. XGBoost's lambda (L2) and min_child_weight provided no benefit because implicit regularization from learning rate, max_depth, subsample, and early stopping already prevents overfitting. Extra regularization slightly worsened the metrics, so it was omitted.

**Bayesian Optimization vs Grid Search**  
We also tested Bayesian Optimization, but it was ineffective. With only 4 parameters and a small search space, it added overhead, slowed hyperparameter tuning, and offered no meaningful improvement. Grid search is sufficient for this dataset and model.

**10-fold CV:** Matches SVM for fair comparison. Each fold has about 100 samples, giving stable MAE estimates.

**Early stopping (50 rounds):** Stops training if validation MAE doesn't improve for 50 rounds, automatically choosing the optimal number of trees and avoiding overfitting.


```{r xgb-train-final}
xgb_model_file <- file.path(MODEL_DIR, "model_xgb_comparable.rds")

if (file.exists(xgb_model_file)) {
  # Load existing model
  model_xgb <- readRDS(xgb_model_file)
  # cat("XGBoost model loaded from:", xgb_model_file, "\n")
} else {
  # Train new model
  final_params <- list(objective = "reg:squarederror", eval_metric = "mae",
    eta = best_params$eta, max_depth = best_params$max_depth,
    subsample = best_params$subsample, colsample_bytree = best_params$colsample_bytree, nthread = 1)
  
  set.seed(SEED)
  model_xgb <- xgb.train(params = final_params, data = dtrain, nrounds = best_params$nrounds, verbose = 0)
  
  # Save model for future runs
  saveRDS(model_xgb, xgb_model_file)
  cat("XGBoost model trained and saved to:", xgb_model_file, "\n")
}

# Define final_params for use in learning curves (needed even when loading)
final_params <- list(objective = "reg:squarederror", eval_metric = "mae",
  eta = best_params$eta, max_depth = best_params$max_depth,
  subsample = best_params$subsample, colsample_bytree = best_params$colsample_bytree, nthread = 1)

preds_xgb <- predict(model_xgb, dtest)
metrics_xgb <- get_metrics(preds_xgb, y_test)
```

## Diagnostic Plots

```{r xgb-learning-curves}
train_mae <- test_mae <- numeric()
max_rounds <- min(500, best_params$nrounds * 2)

for (r in seq(10, max_rounds, by = 10)) {
  set.seed(SEED)
  temp_model <- xgb.train(params = final_params, data = dtrain, nrounds = r, verbose = 0)
  train_mae <- c(train_mae, mean(abs(predict(temp_model, dtrain) - y_train)))
  test_mae <- c(test_mae, mean(abs(predict(temp_model, dtest) - y_test)))
}

training_history <- data.frame(iter = seq(10, max_rounds, by = 10), train_mae, test_mae)
learning_curve_df <- training_history %>%
  pivot_longer(cols = c(train_mae, test_mae), names_to = "Dataset", values_to = "MAE") %>%
  mutate(Dataset = ifelse(Dataset == "train_mae", "Train", "Validation"))

best_row <- which.min(abs(training_history$iter - best_params$nrounds))
overfit_gap <- training_history$test_mae[best_row] - training_history$train_mae[best_row]
gap_pct <- round(100 * overfit_gap / training_history$test_mae[best_row], 1)
```

```{r xgb-diagnostic-plots, fig.width=6.5, fig.height=2.3, out.width="100%"}
p1 <- ggplot(learning_curve_df, aes(x = iter, y = MAE, color = Dataset)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = best_params$nrounds, linetype = "dashed", color = "darkgreen", linewidth = 0.5) +
  scale_color_manual(values = c("Train" = "#2E86AB", "Validation" = "#E63946")) +
  labs(x = "Rounds", y = "MAE", title = "Learning Curves") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom", legend.key.size = unit(0.3, "cm"),
    plot.title = element_text(size = 9), legend.title = element_blank())

pred_df <- data.frame(pred = preds_xgb, actual = y_test, residual = y_test - preds_xgb)

p2 <- ggplot(pred_df, aes(x = pred, y = actual)) +
  geom_point(alpha = 0.5, size = 0.8, color = "#2E86AB") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 0.7) +
  labs(x = "Predicted", y = "Actual", title = "Pred vs Actual") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

p3 <- ggplot(pred_df, aes(x = residual)) +
  geom_histogram(bins = 25, fill = "#2E86AB", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 0.7) +
  labs(x = "Residual", y = "Count", title = "Residual Distribution") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

grid.arrange(p1, p2, p3, ncol = 3)
```

The **learning curves** show that the model converges with minimal overfitting. In the first 50 rounds, training and validation MAE drop as the model captures the main signal. After that, improvement slows and additional rounds provide little benefit, which justifies early stopping. The small train-validation gap of 0.021 MAE indicates that overfitting is minimal.

The **predictions vs actual** plot shows that most points lie close to the diagonal, indicating accurate predictions. Larger errors appear at the highest and lowest quality scores, which is expected because extreme values are rare and the model has fewer examples to learn from. There is no visible systematic bias, as points are evenly distributed around the diagonal.

The **residuals** are centered around zero, confirming that the prediction bias is minimal. Their roughly bell-shaped distribution suggests that errors are mostly random rather than systematic, meaning the model has captured the main signal.

<!-- Note: This part was important during the work on the XGBoost part, but we have to comment it out from the final report because of the report length limitations. However, we did not want to delete it completely because of the work put into this part.

## Feature Importance

```{r xgb-feature-importance, fig.width=3.5, fig.height=2.3, out.width="55%"}
importance <- xgb.importance(feature_names = colnames(X_train), model = model_xgb)
imp_df <- head(importance, 10)
imp_df$Feature <- factor(imp_df$Feature, levels = rev(imp_df$Feature))

ggplot(imp_df, aes(x = Gain, y = Feature)) +
  geom_col(fill = "#2E86AB", width = 0.7) +
  labs(x = "Gain", y = NULL, title = "Top 10 Features by Gain") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))
```

```{r xgb-importance-inline, results='asis'}
top4 <- head(importance, 4)
cat("\\begin{center}
\\scriptsize
\\begin{tabular}{lrrr}
\\toprule
Feature & Gain & Cover & Frequency \\\\
\\midrule")
for (i in 1:4) {
  cat("\n", top4$Feature[i], " & ", round(top4$Gain[i], 3), " & ", round(top4$Cover[i], 3), " & ", round(top4$Frequency[i], 3), " \\\\")
}
cat("\n\\bottomrule
\\end{tabular}
\\end{center}")
```

XGBoost provides three **feature importance** measures: gain, cover, and frequency. Gain shows how much a feature reduces prediction error, so higher gain means higher predictive power. For example, Flavor has the highest gain (about 0.3), meaning it explains roughly 30% of the model's total error reduction. Cover measures how many samples are affected by splits on a feature; high cover means the feature is used on large parts of the data. Frequency counts how often a feature is used in splits, but frequent use does not necessarily mean strong impact.

The ranking shows that Flavor is by far the most important feature, which matches domain knowledge, because flavor is the most important criterion for cupping evaluations. Other top features (Aroma, Aftertaste, Balance) are also sensory attributes. Non-sensory features such as altitude or processing method contribute very little, likely because their effects are already reflected in the sensory scores. 

-->

## Conclusions

**Model performance:** XGBoost achieves an MAE of `r round(metrics_xgb$MAE, 3)`, meaning predictions are usually within `r round(metrics_xgb$MAE, 2)` points of the true cupper score on a 10-point scale. This is a small but consistent improvement over linear regression (MAE `r round(metrics_lm$MAE, 3)`). The $R^2$ increases from `r round(metrics_lm$R2, 2)` to `r round(metrics_xgb$R2, 2)`, showing that XGBoost explains about `r round((metrics_xgb$R2 - metrics_lm$R2) * 100)`% more variance.

**Limitations:** Performance is weaker at extreme quality scores due to limited data. The dataset size (about 1300 samples) may restrict learning of complex patterns, and temporal effects such as harvest year are not modeled (because of the issues with source data). Finally, the target variable cupper points is subjective and includes human rating variability.

# Explainable AI
```{r xgb-xai-setup}
if (!require("DALEX")) install.packages("DALEX", quiet = TRUE)
library(DALEX)

predict_xgb <- function(model, newdata) {
  if (is.data.frame(newdata)) newdata <- as.matrix(newdata)
  predict(model, xgb.DMatrix(data = newdata))
}

explainer_xgb <- DALEX::explain(
  model = model_xgb, data = X_train, y = y_train,
  predict_function = predict_xgb, label = "XGBoost", verbose = FALSE
)
```

### Permutation Feature Importance (PFI)

PFI measures how much model performance degrades when a feature's signal is destroyed by shuffling [@friedman2001greedy]. Unlike XGBoost's native Gain metric, PFI is model-agnostic and accounts for feature interactions.
```{r svm-explainer-setup}
# ---- SVM Explainer ----
explainer_svm <- DALEX::explain(
  model = svm_model, 
  data = test_data[, -which(names(test_data) == "Cupper.Points")], 
  y = test_data$Cupper.Points, 
  label = "SVM Coffee Model",
  verbose = FALSE
)

# ---- Loss function ----
loss_mae <- function(observed, predicted) {
  mean(abs(observed - predicted))
}
```
```{r svm-pfi-calc}
# ---- SVM PFI ----
set.seed(SEED)
pfi_svm <- model_parts(
  explainer = explainer_svm,
  loss_function = loss_mae,
  type = "difference"
)

# ---- Tidy SVM PFI ----
pfi_results_svm <- pfi_svm %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(
    Mean_Dropout_Loss = mean(dropout_loss),
    SD = sd(dropout_loss),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(
    Importance = Mean_Dropout_Loss -
      min(pfi_svm$dropout_loss[pfi_svm$variable == "_full_model_"]),
    Rank = row_number()
  )
```
```{r svm-ale-calc}
# ---- Select top 4 SVM features ----
top_features_svm <- head(pfi_results_svm$variable, 4)

# ---- SVM ALE ----
set.seed(SEED)
ale_plots_svm <- lapply(top_features_svm, function(feature) {
  model_profile(
    explainer = explainer_svm,
    variables = feature,
    type = "accumulated",
    N = 1000
  )
})

names(ale_plots_svm) <- top_features_svm

# ---- Center ALE at 0 ----
ale_plots_svm <- lapply(ale_plots_svm, function(mp) {
  mp$agr_profiles$`_yhat_` <-
    mp$agr_profiles$`_yhat_` - mean(mp$agr_profiles$`_yhat_`)
  mp
})
```
```{r plot-svm-pfi-ale, fig.width=10, fig.height=2.8, out.width="100%"}
pfi_plot_data_svm <- head(pfi_results_svm, 8)
pfi_plot_data_svm$variable <- factor(
  pfi_plot_data_svm$variable,
  levels = rev(pfi_plot_data_svm$variable)
)

p_pfi_svm <- ggplot(pfi_plot_data_svm, aes(x = Importance, y = variable)) +
  geom_col(fill = "#2E86AB", width = 0.8) +
  geom_errorbarh(
    aes(xmin = Importance - SD, xmax = Importance + SD),
    height = 0.35,
    linewidth = 0.4
  ) +
  labs(
    x = "MAE Increase",
    y = NULL,
    title = "PFI (SVM)"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(size = 10),
    plot.margin = margin(5, 10, 5, 5, "pt")
  )

ale_combined_data_svm <- do.call(
  rbind,
  lapply(names(ale_plots_svm), function(feat) {
    d <- ale_plots_svm[[feat]]$agr_profiles
    data.frame(
      Feature = feat,
      x = d$`_x_`,
      ale = d$`_yhat_`
    )
  })
)

p_ale_svm <- ggplot(ale_combined_data_svm, aes(x = x, y = ale)) +
  geom_line(color = "#2E86AB", linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
  facet_wrap(~Feature, scales = "free", nrow = 1) +
  labs(
    x = "Feature Value",
    y = "ALE",
    title = "ALE (SVM)"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    strip.text = element_text(size = 7, face = "bold"),
    plot.title = element_text(size = 10),
    plot.margin = margin(5, 5, 5, 10, "pt"),
    panel.spacing = unit(0.8, "lines")
  )

gridExtra::grid.arrange(
  p_pfi_svm,
  p_ale_svm,
  ncol = 2,
  widths = c(1.2, 2.8)   
)
```
```{r xgb-pfi-calc}
# ---- XGBoost PFI ----
set.seed(SEED)
pfi_xgb <- model_parts(
  explainer = explainer_xgb, 
  loss_function = loss_mae, 
  type = "difference", 
  B = 10
)

# ---- Tidy XGBoost PFI ----
pfi_results_xgb <- pfi_xgb %>%
  filter(variable != "_baseline_" & variable != "_full_model_") %>%
  group_by(variable) %>%
  summarise(
    Mean_Dropout_Loss = mean(dropout_loss), 
    SD = sd(dropout_loss), 
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Dropout_Loss)) %>%
  mutate(
    Importance = Mean_Dropout_Loss - min(pfi_xgb$dropout_loss[pfi_xgb$variable == "_full_model_"]),
    Rank = row_number()
  )
```
```{r xgb-ale-calc}
# ---- Select top 4 XGBoost features ----
top_features_xgb <- head(pfi_results_xgb$variable, 4)

# ---- XGBoost ALE ----
set.seed(SEED)
ale_plots_xgb <- lapply(top_features_xgb, function(feature) {
  model_profile(
    explainer = explainer_xgb,
    variables = feature,
    type = "accumulated",
    N = 1000
  )
})

names(ale_plots_xgb) <- top_features_xgb

# ---- Center ALE at 0 ----
ale_plots_xgb <- lapply(ale_plots_xgb, function(mp) {
  mp$agr_profiles$`_yhat_` <-
    mp$agr_profiles$`_yhat_` - mean(mp$agr_profiles$`_yhat_`)
  mp
})
```
```{r plot-xgb-pfi-ale, fig.width=10, fig.height=2.8, out.width="100%"}
pfi_plot_data_xgb <- head(pfi_results_xgb, 10)
pfi_plot_data_xgb$variable <- factor(
  pfi_plot_data_xgb$variable, 
  levels = rev(pfi_plot_data_xgb$variable)
)

p_pfi_xgb <- ggplot(pfi_plot_data_xgb, aes(x = Importance, y = variable)) +
  geom_col(fill = "#2E86AB", width = 0.7) +
  geom_errorbarh(
    aes(xmin = Importance - SD, xmax = Importance + SD), 
    height = 0.3, 
    linewidth = 0.4
  ) +
  labs(x = "MAE Increase", y = NULL, title = "PFI (XGBoost)") +
  theme_minimal(base_size = 9) +
  theme(
    plot.title = element_text(size = 10),
    plot.margin = margin(5, 10, 5, 5, "pt")
  )

ale_combined_data_xgb <- do.call(
  rbind,
  lapply(names(ale_plots_xgb), function(feat) {
    ale_data <- ale_plots_xgb[[feat]]$agr_profiles
    data.frame(
      Feature = feat,
      x = ale_data$`_x_`,
      ale = ale_data$`_yhat_`
    )
  })
)

p_ale_xgb <- ggplot(ale_combined_data_xgb, aes(x = x, y = ale)) +
  geom_line(color = "#2E86AB", linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~Feature, scales = "free", nrow = 1) +
  labs(
    x = "Feature Value",
    y = "ALE",
    title = "ALE (XGBoost)"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    strip.text = element_text(size = 8, face = "bold"),
    plot.title = element_text(size = 10),
    plot.margin = margin(5, 5, 5, 10, "pt"),
    panel.spacing = unit(0.8, "lines")
  )

grid.arrange(
  p_pfi_xgb,
  p_ale_xgb,
  ncol = 2,
  widths = c(1.2, 2.8)
)
```

The top features are all sensory scores, confirming the model relies on expert taste evaluation.


### Accumulated Local Effects (ALE)

ALE plots show the marginal effect of a feature on the model's predictions by averaging local changes in the prediction over the distribution of the other features [@apley2020visualizing]. It is chosen here instead of PDP (Partial Dependence plots), because it is more robust to multicollinearities and centered around 0. We plot the top four features ranked by permutation feature importance (PFI).

All sensory features show the expected pattern that higher scores lead to higher predicted quality, only in Balance there is a sudden dip to a negative impact on the cupper score around values of 7, after which the effect goes back to being positive. The ALE range for Flavor is largest, indicating it has the strongest marginal effect on predictions.“


### Shapley Additive Explanations (SHAP)

SHAP values provide a unified measure of feature importance based on cooperative game theory [@lundberg2017unified]. For each prediction, SHAP assigns each feature a value that represents its contribution to moving the prediction away from the baseline (average prediction). We utilize the SHAP values to inspect the microscopic effects of the feature values for two specific observations. For this we chose one observation with a high value for Flavor and one with a low value, since this is the most important feature according to PFI.

```{r plot-xgb-shap-waterfall, fig.width=9, fig.height=2.8, out.width="100%"}
set.seed(SEED)

flavor_col <- which(colnames(X_train) == "Flavor")
wf_plots_xgb <- vector("list", 2)

# Find min and max Flavor observations
obs_id_low <- which.min(X_train[, flavor_col])
obs_id_high <- which.max(X_train[, flavor_col])
obs_ids <- c(obs_id_low, obs_id_high)

for (i in 1:2) {
  obs_id <- obs_ids[i]
  new_obs_matrix <- X_train[obs_id, , drop = FALSE]
  actual_flavor <- round(X_train[obs_id, flavor_col], 2)
  storage.mode(new_obs_matrix) <- "numeric"
  
  sv <- shapviz(
    explainer_xgb$model,
    X_pred = new_obs_matrix
  )
  
  wf_plots_xgb[[i]] <- sv_waterfall(sv, max_display = 8) +
    ggtitle(paste("XGBoost: Flavor =", actual_flavor)) +
    theme_minimal(base_size = 8) +
    theme(
      plot.title = element_text(size = 9),
      axis.title = element_blank(),
      axis.text  = element_text(size = 7),
      plot.margin = margin(2, 5, 2, 5, "pt")
    )
}

gridExtra::grid.arrange(wf_plots_xgb[[1]], wf_plots_xgb[[2]], ncol = 2)
```

```{r plot-svm-shap-waterfall, fig.width=11, fig.height=3.2, out.width="100%", echo=FALSE}
library(patchwork)
svm_plots <- list()

flavor_col_train <- which(names(train_data) == "Flavor")
obs_id_low <- which.min(train_data$Flavor)
obs_id_high <- which.max(train_data$Flavor)
obs_ids <- c(obs_id_low, obs_id_high)

for (i in 1:2) {
  obs_id <- obs_ids[i]
  new_obs <- train_data[obs_id, , drop = FALSE]
  actual_flavor <- round(train_data$Flavor[obs_id], 2)
  
  bd_svm <- predict_parts(
    explainer = explainer_svm, 
    new_observation = new_obs, 
    type = "break_down"
  )
  
  svm_plots[[i]] <- plot(bd_svm, 
                         max_features = 8, 
                         digits = 2, 
                         shift_values = TRUE) + 
    ggtitle(paste("SVM: Flavor =", actual_flavor)) +
    theme_minimal(base_size = 8) + 
    theme(
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
      axis.title.x = element_blank(),
      axis.text.y = element_text(size = 7),
      plot.margin = margin(5, 8, 5, 8)
    )
}

svm_plots[[1]] + svm_plots[[2]]
```
The microscopic plots for both models show the expected effects of the Flavor feature on the overall prediction. For a Flavor of 0, both models strongly decrease their prediction, XGBoost by 1.74 and SVM by 2 cupper points. Interestingly, the XGBoost model derives an even greater effect (-2.82) from Aroma of 0 in this particular observation, even though Aroma is only the 5th most important feature according to PFI. 
The highest value of Flavor (8.83) leads to an increase of 0.425 cupper points in the XGBoost model and 0.39 in SVM. Since the sensory features are also highly correlated, their values are all relatively low or high in accordance to the Flavor feature. This leads to the effects being spread across the different features and could explain why in the first observation the aroma has an even stronger effect than the Flavor in the XGBoost model.

# Model Comparison - SVM vs XGBoost

```{r comparison-metrics}
# Retrieve SVM results saved earlier
svm_mae <- svm_results$mae
svm_rmse <- svm_results$rmse
svm_r2 <- svm_results$r2
svm_predictions <- svm_results$predictions
svm_test_cupper_points <- svm_results$test_cupper_points
best_run <- svm_results$best_run

# Compile comparison metrics
comparison_df <- data.frame(
  Model = c("SVM (Linear Kernel)", "XGBoost"),
  MAE = c(round(svm_mae, 4), round(metrics_xgb$MAE, 4)),
  RMSE = c(round(svm_rmse, 4), round(metrics_xgb$RMSE, 4)),
  R2 = c(round(svm_r2, 4), round(metrics_xgb$R2, 4))
)

kable(comparison_df)
```

```{r comparison-plots-combined, fig.height=2.4, fig.width=6.5, out.width="85%"}
# Create side-by-side scatter plots using ggplot2 for better layout control
library(gridExtra)

# SVM scatter plot
svm_df <- data.frame(pred = svm_predictions, actual = svm_test_cupper_points)
p_svm <- ggplot(svm_df, aes(x = pred, y = actual)) +
  geom_point(alpha = 0.5, size = 1.2, color = "#E63946") +
  geom_abline(slope = 1, intercept = 0, color = "darkred", linewidth = 0.8) +
  labs(x = "Predictions", y = "True Values", 
       title = paste0("SVM (MAE: ", round(svm_mae, 4), ")")) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9, face = "bold"))

# XGBoost scatter plot
xgb_df <- data.frame(pred = preds_xgb, actual = y_test)
p_xgb <- ggplot(xgb_df, aes(x = pred, y = actual)) +
  geom_point(alpha = 0.5, size = 1.2, color = "#2E86AB") +
  geom_abline(slope = 1, intercept = 0, color = "navy", linewidth = 0.8) +
  labs(x = "Predictions", y = "True Values", 
       title = paste0("XGBoost (MAE: ", round(metrics_xgb$MAE, 4), ")")) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9, face = "bold"))

# Residual histogram
svm_residuals <- svm_test_cupper_points - svm_predictions
xgb_residuals <- y_test - preds_xgb
residuals_df <- data.frame(
  Residual = c(svm_residuals, xgb_residuals),
  Model = c(rep("SVM", length(svm_residuals)), rep("XGBoost", length(xgb_residuals)))
)

p_resid <- ggplot(residuals_df, aes(x = Residual, fill = Model)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("SVM" = "#E63946", "XGBoost" = "#2E86AB")) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 0.5) +
  labs(x = "Residual", y = "Count", title = "Residual Distribution") +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9, face = "bold"),
        legend.position = "bottom", legend.key.size = unit(0.3, "cm"))

# Arrange all three plots side by side
grid.arrange(p_svm, p_xgb, p_resid, ncol = 3)
```

Both models achieve similar predictive accuracy, with XGBoost having slightly smaller MAE (`r round(metrics_xgb$MAE, 3)` vs `r round(svm_mae, 3)`) and larger $R^2$ (`r round(metrics_xgb$R2, 2)` vs `r round(svm_r2, 2)`). The scatter plots confirm that both models produce predictions closely aligned with true values, with no systematic bias evident in either case. The residual distributions are nearly identical and centered around zero, indicating that both approaches effectively capture the underlying signal in the data.


## Literature

<div id="refs"></div>

## Statement of AI Use 

Generative AI was used in this report to generate the R code for the analyses. Across all programming sections LLMs have been used to aid in code generation. The authors state that the generated code has been manualy adjusted, verified and tested to ensure high quality. AI was also used to help formatting the report: writing LaTeX mathematical formulas and adjusting the tables size and formatting. 

However, the text part of the report, the analysis of the plots, the research, the development the workflow etc. was done by hand using the stated sources as well as lecture material and examples from the exercises.

